\chapter{Appendix: Candidate Definitions and Results}

This appendix collects paper-ready definitions and results that fit the repository's
current architecture.

\begin{definition}[Hybrid global section]
Let $P$ be a program and let $\Layer$ be the ordered set
$\mathrm{INTENT} < \mathrm{STRUCTURAL} < \mathrm{SEMANTIC} < \mathrm{PROOF} < \mathrm{CODE}$.
A hybrid global section for $P$ is a section of the presheaf
\[
  \HybridTy : (\Site(P) \times \Layer)^{\op} \to \mathsf{RefinementLattice} \times \mathsf{TrustLattice}
\]
that is compatible under every adjacent restriction map.
\end{definition}

\begin{definition}[Oracle-valued semantic predicate]
An oracle-valued semantic predicate on a carrier $A$ is a map
\[
  \varphi_s : A \to \TO(\Bool)
\]
where
\[
  \TO(B) = B \times \mathsf{TrustLevel} \times \mathsf{Confidence} \times \mathsf{Provenance}.
\]
\end{definition}

\begin{theorem}[Weakest-link trust composition]
If semantic computations are composed by monadic bind in $\TO$, then the trust of the
composite is bounded above by the meet of the trust levels of the constituents.
\end{theorem}

\begin{proof}[Proof sketch]
The monadic structure is defined to propagate trust via meet. Therefore every downstream
semantic result inherits the weakest trust level seen along the path. This is exactly the
engineering behavior desired by the repository, since high-trust stages should not erase
low-trust inputs.
\end{proof}

\begin{theorem}[Obstruction criterion for hybrid coherence]
If $\Hc{1}(\Layer, \HybridTy) = 0$, then every adjacent pair of layer sections glues to a
coherent hybrid type over the chosen cover.
\end{theorem}

\begin{remark}
This theorem should always be read relative to a chosen cover and chosen restriction maps.
The repository's mature position is strongest when it makes such relativity explicit.
\end{remark}

\begin{proposition}[Mixed-mode compilation as a bridge]
Every mixed-mode surface fragment can be understood as producing either a plain type-level
obligation, a hybrid trust-bearing contract, or both. In particular, user-facing syntax is
not a parallel semantics but a bridge into the existing object language and layered
semantics.
\end{proposition}
