\chapter{Appendix: Repository Map}

\begin{longtable}{p{0.28\textwidth}p{0.64\textwidth}}
\toprule
\textbf{Path or namespace} & \textbf{Role in the cohomological typing story} \\
\midrule
\endhead
\repo{src/deppy/types/base.py} & Foundational carrier hierarchy for plain typing. \\
\repo{src/deppy/types/refinement.py} & Refinement predicates and the type-level shape of local facts. \\
\repo{src/deppy/types/dependent.py} & Dependent function and pair formers, identity-like objects, indexed families. \\
\repo{src/deppy/types/subtyping.py} & Ordering and obligation structure for the refinement lattice. \\
\repo{src/deppy/hybrid/core/types.py} & Hybrid type semantics, layer order, trust-facing hybrid type formers. \\
\repo{src/deppy/hybrid/core/checker.py} & Bidirectional hybrid checking and semantic-cache logic. \\
\repo{src/deppy/hybrid/core/algebraic_geometry.py} & Layered presheaves, Cech complexes, cohomological obstruction machinery. \\
\repo{src/deppy/hybrid/core/trust.py} & Trust lattice, evidence classification, provenance plumbing. \\
\repo{src/deppy/hybrid/mixed_mode/syntax.py} & Gracefully degrading user-facing syntax for hybrid programming. \\
\repo{src/deppy/hybrid/discharge/cascade.py} & Staged obligation discharge from trivial proofs to residuals. \\
\repo{src/deppy/hybrid/lean_translation/} & Translation of typed and proof-bearing artifacts into Lean-facing forms. \\
\repo{src/deppy/hybrid/type_expansion/} & Elaboration of prompt or intent material into typed plans and covers. \\
\repo{src/deppy/hybrid/anti_hallucination/} & Artifact typing and structural / semantic hallucination classification. \\
\repo{src/deppy/hybrid/zero_change/} & Inference over existing Python without annotation-first workflows. \\
\repo{src/deppy/equivalence/} & Local-to-global equivalence, semantic diffing, and descent-style reasoning. \\
\repo{src/deppy/render/} & Turning typed results, residuals, and trust labels into readable surfaces. \\
\repo{src/deppy/theory_packs/} & Domain-local solver geometries and library-facing semantics. \\
\repo{agent/orchestrator.py} & Prompt-driven cover growth, generation, verification, and assembly. \\
\repo{web/} & Human-facing explanation layer summarizing the mathematics and APIs. \\
\bottomrule
\end{longtable}

This table should not be read as a dependency graph with one single topological order. It
is a semantic map. The repository's pieces feed one another in multiple directions, but the
map makes clear that the plain and hybrid type systems remain central reference points for
most of the interesting work.
