% Chapter: Cubical Structure, the Two Dimension Systems, and Descent
% Part of the CCTT Monograph — Proof Calculus, Chapter II
%
% REWRITTEN to reflect the Synergy Atlas (synergies.tex).
% Central theme: the cubical and cohomological structures are unified,
% not independent.  HComp IS descent (Synergy 0).  The interval
% connections ARE the cover algebra (Synergies 10, 13).  GluePath IS
% parameterized descent (Synergy 9).

\chapter{Cubical Structure, the Two Dimension Systems, and Descent}
\label{ch:c4-cubical}

In Chapter~\ref{ch:c4-calculus} we introduced the syntax and typing rules of
$\mathrm{C}^4$.  In Chapter~\ref{ch:c4-descent} we introduced the cohomological
descent machinery: refinement covers, $\Cech$ data, and the sheaf gluing condition.
The present chapter reveals that these are not two independent theories sitting
side by side.  They are \emph{the same universal construction} seen from two
different angles.

The central thesis is this: $\mathrm{C}^4$ has \textbf{two independent
dimension systems}.  The \emph{cubical} dimensions are parameterized by the
interval $\interval$; they capture proof steps, equality witnesses, and
homotopies.  The \emph{cohomological} dimensions are parameterized by
refinement predicates $\varphi(x)$; they capture which inputs a proof applies
to, and how local proofs are assembled into global ones.  A proof lives in the
product space $\interval^m \times \site^n$, where $m$ is the number of proof
steps and $n$ is the number of input cases.

The identification of these two structures is not a mere analogy.  It is
a precise theorem: $\hfilling$ (homogeneous composition) and
$\mathrm{desc}$ ($\Cech$ descent) have the same universal property.  We prove
this as Theorem~\ref{thm:hcomp-is-descent} in
Section~\ref{sec:hcomp-is-descent}, which is the first major theorem of this
chapter.  Everything else follows from this identification.

After establishing the foundational identification, we show:
\begin{itemize}
\item The connection algebra $(\wedge, \vee, \sim)$ on $\interval$ is
  isomorphic (as a De Morgan algebra) to the cover algebra
  $(\wedge, \vee, \neg)$ on refinement predicates.  This is why cover
  simplification (Synergy~13) works: it is the same algebra.

\item \textsf{GluePath} (Synergy~9) is a \emph{parameterized} descent,
  in which the $\Cech$ data varies continuously over the interval.  This is
  strictly more expressive than either component alone.

\item Fiber-preserving homotopies (Synergy~20) are paths in proof space that
  respect the fibration over the refinement site.  They are the
  $E_1$-page data of the spectral sequence.

\item Overlap compatibility (Synergy~7) and cubical edge agreement are the
  same condition, stated in the two languages.

\item Kan operations ($\hfilling$, $\transp$) are derived as special cases of
  the descent rule applied in the cubical direction.

\item Univalence and function extensionality retain their standard proofs but
  are re-read through the lens of descent and fibered universes.
\end{itemize}


% ─────────────────────────────────────────────────────────
\section{The Two Dimension Systems}
\label{sec:two-dimensions}

\begin{definition}[The Two Dimension Systems of $\mathrm{C}^4$]
\label{def:two-dims}
$\mathrm{C}^4$ has two families of \textbf{dimensions}:

\medskip
\noindent\textbf{Cubical dimensions.}  These are \emph{interval variables}
$i, j, k : \interval$.  The interval $\interval$ is a type with two
distinguished endpoints $\mathsf{0}, \mathsf{1} : \interval$ and the
connection algebra $(\wedge, \vee, \sim)$.  An $m$-dimensional cubical
context is $i_1 : \interval, \ldots, i_m : \interval$; a term in such a
context is a \emph{cube} $A(i_1, \ldots, i_m)$.

\medskip
\noindent\textbf{Cohomological dimensions.}  These are \emph{refinement
predicates} $\varphi_1(x), \ldots, \varphi_n(x)$, each a Z3-decidable
formula in the input variable $x$.  They form a cover
$\covers = \{\varphi_1, \ldots, \varphi_n\}$ of the input domain
$X \defeq \PyObj$ whenever $\bigvee_k \varphi_k(x)$ is a Z3 tautology.
An $n$-dimensional cohomological context is given by such a cover;
a term in such a context is a tuple of \emph{fiber values}
$(t_1, \ldots, t_n)$ with $t_k$ defined under $\varphi_k$.
\end{definition}

\begin{definition}[The Product Proof Space]
\label{def:product-space}
The \textbf{product proof space} for a function with $n$ input cases and
a proof of $m$ steps is:
\[
  \interval^m \times \site^n
\]
where $\site^n \defeq \prod_{k=1}^n \{x \mid \varphi_k(x)\}$ is the
product of the $n$ fibers.  A point $(i_1, \ldots, i_m, k)$ in this space
represents: step $i_1, \ldots, i_m$ of the proof on input case $k$.

\medskip
\noindent The \textbf{cubical dimensions} $\interval^m$ parameterize
\emph{how} we prove: what equalities we traverse, what homotopies we
construct.  The \textbf{cohomological dimensions} $\site^n$ parameterize
\emph{what} we prove for: which slice of the input domain is under
consideration.  Full verification of a function requires filling the
entire product space.
\end{definition}

\begin{center}
\begin{tabular}{lll}
\toprule
& \textbf{Cubical dimensions} & \textbf{Cohomological dimensions} \\
\midrule
Parameterized by & Interval variables $i, j, k : \interval$ &
  Refinement predicates $\varphi(x)$ (Z3-decidable) \\
$0$-cells & Programs / proof terms & Input values \\
$1$-cells & Equality proofs (paths) & Fiber restrictions \\
$2$-cells & Proof coherence (squares) & Cover overlaps \\
Composition & $\hfilling$ (Kan filling) & Descent (gluing) \\
Transport & Along type paths & Along predicate implications \\
Obstruction & Kan condition & $\Hone(\covers, \presheaf) = 0$ \\
Decidability & Structural normalization & Z3 \\
Algebra & De Morgan on $\interval$ & De Morgan on $\mathbb{F}$ \\
\bottomrule
\end{tabular}
\end{center}

\begin{remark}[Why two systems?]
Cubical type theory alone can fill cubes but has no notion of ``which inputs
the proof applies to.''  Sheaf cohomology alone can decompose by input cases
but has no notion of proof paths connecting them.  The product structure
$\interval^m \times \site^n$ is what allows $\mathrm{C}^4$ to do both:
decompose a proof by input cases \emph{and} connect those decomposed proofs
by paths, all in a single formal framework.
\end{remark}


% ─────────────────────────────────────────────────────────
\section{Theorem 1: HComp Is Descent}
\label{sec:hcomp-is-descent}

This section contains the foundational identification of $\mathrm{C}^4$.
We prove that homogeneous composition ($\hfilling$) and $\Cech$ descent
($\mathrm{desc}$) are the same universal construction --- not merely analogous,
but formally identified in a single verification-condition generator.

\begin{definition}[$\Cech$ Descent Rule]
\label{def:descent-rule}
Let $\covers = \{\varphi_1, \ldots, \varphi_n\}$ be a cover of $X$
(i.e., $\bigvee_k \varphi_k$ is a Z3 tautology).  Let $\presheaf$ be a
sheaf of types on $\site$.  Given:
\begin{itemize}
\item \textbf{Local sections}: for each $k$, a term
  $s_k : \presheaf(\varphi_k)$, i.e., $\Gamma, \varphi_k(x) \vdash s_k : A$;
\item \textbf{Overlap compatibility}: for each pair $k, \ell$ with
  $\varphi_k \wedge \varphi_\ell$ Z3-satisfiable,
  \[
    \restr{s_k}{\varphi_k \wedge \varphi_\ell}
    \;=\;
    \restr{s_\ell}{\varphi_k \wedge \varphi_\ell}
    \quad\text{(as terms in $\Gamma, \varphi_k(x), \varphi_\ell(x) \vdash A$)}
  \]
\end{itemize}
the \textbf{descent rule} produces a global section:
\[
\frac{
  \forall k.\;\Gamma, \varphi_k(x) \vdash s_k : A \qquad
  \forall k, \ell.\;\text{(overlap compat.)}
}{
  \Gamma \vdash \mathrm{desc}(\covers, \{s_k\}) : A
}
\tag{desc}
\]
with computation rule: $\restr{\mathrm{desc}(\covers, \{s_k\})}{\varphi_k}
\equiv s_k$ for each $k$.
\end{definition}

\begin{definition}[Homogeneous Composition --- $\hfilling$]
\label{def:hcomp}
Given $A : \mathcal{U}_i$, a face formula $\varphi$ (a disjunction of
atomic constraints $r = \mathsf{0}$ or $r = \mathsf{1}$), a partial path
$\Gamma, \varphi, i : \interval \vdash u : A$, and a base point
$\Gamma \vdash u_0 : A\,[\varphi \mapsto u[\mathsf{0}/i]]$:
\[
\frac{\Gamma \vdash A : \mathcal{U}_i \qquad
      \Gamma, \varphi, i : \interval \vdash u : A \qquad
      \Gamma \vdash u_0 : A\,[\varphi \mapsto u[\mathsf{0}/i]]}
     {\Gamma \vdash \hfilling^i\, A\, [\varphi \mapsto u]\, u_0 : A}
\tag{hcomp}
\]
with the boundary law: when $\varphi$ holds,
$\hfilling^i\, A\, [\varphi \mapsto u]\, u_0 \equiv u[\mathsf{1}/i]$.
\end{definition}

\begin{theorem}[HComp is Descent --- Foundational Identification]
\label{thm:hcomp-is-descent}
Homogeneous composition $\hfilling$ and the $\Cech$ descent rule
$\mathrm{desc}$ are the same universal construction, under the following
identification:
\begin{enumerate}
\item Each \textbf{refinement fiber} $\varphi_k$ corresponds to a
  \textbf{cubical face} $[i_k = \varepsilon_k]$ of $\interval^n$.

\item Each \textbf{local proof} $s_k : A\,[\varphi_k]$ corresponds to a
  \textbf{partial boundary datum} $u_k : A$ defined on the face
  $\{i_k = \varepsilon_k\}$.

\item The \textbf{overlap compatibility condition}
  $\restr{s_k}{\varphi_k \wedge \varphi_\ell}
  = \restr{s_\ell}{\varphi_k \wedge \varphi_\ell}$
  corresponds to the \textbf{cubical edge agreement condition}:
  the partial data on face $[i_k = \varepsilon_k]$ and face
  $[i_\ell = \varepsilon_\ell]$ agree on their shared edge
  $[i_k = \varepsilon_k, i_\ell = \varepsilon_\ell]$.

\item The \textbf{Kan filling condition} (``$\hfilling$ exists'')
  corresponds to $\Hone(\covers, \presheaf) = 0$ for the $\Cech$ nerve
  of $\covers$.

\item For propositional coefficients (path types over types with
  decidable equality --- our setting, by Hedberg's theorem),
  $\Hone = 0$ automatically.  Hence $\hfilling$ always exists
  when the face data is consistent.
\end{enumerate}
In consequence, a single verification-condition generator
$\mathtt{compile\_filling}$ handles both $\mathsf{RefinementDescent}$
and $\mathsf{HComp}$ proof terms, differing only in whether the
face/fiber system is presented as Z3-decidable refinement predicates
(cohomological direction) or as cubical dimension constraints
(cubical direction).
\end{theorem}

\begin{proof}
We construct a formal translation in both directions.

\medskip
\noindent\textbf{Descent to HComp.}
Given descent data $(\covers, \{s_k\}, \{\text{compat}_{k\ell}\})$,
encode each fiber $\varphi_k$ as an atomic constraint $i_k = \mathsf{0}$
on a fresh interval variable $i_k$.  The partial path
$u : A$ is defined on the face $\bigvee_k (i_k = \mathsf{0})$ by
$u[\mathsf{0}/i_k] \defeq s_k$.  The overlap compatibility translates
directly to edge agreement: on the shared edge
$(i_k = \mathsf{0}) \wedge (i_\ell = \mathsf{0})$, the data
$s_k$ and $s_\ell$ agree by the compatibility hypothesis.
Apply $\hfilling^i\, A\, [\bigvee_k (i_k = \mathsf{0}) \mapsto u]\,
u_{\mathrm{base}}$ to obtain the global section.  By the boundary law,
its restriction to each face $[i_k = \mathsf{0}]$ recovers $s_k$.

\medskip
\noindent\textbf{HComp to Descent.}
Conversely, given the input to $\hfilling$ --- type $A$, face formula
$\varphi = \bigvee_k (r_k = \varepsilon_k)$, partial path $u : A\,[\varphi]$,
and base $u_0$ --- define a cover $\covers \defeq \{\varphi_k\}$ where
$\varphi_k \defeq (r_k = \varepsilon_k)$ viewed as a refinement predicate on
the interval variable $r_k$.  Each $s_k \defeq u[\varepsilon_k / r_k]$
is a local section.  Edge agreement gives overlap compatibility.
The output of $\hfilling$ is the global section produced by descent.

\medskip
\noindent\textbf{Why $\Hone = 0$.}
The coefficient sheaf $\presheaf$ assigns, to each refinement predicate
$\varphi$, the type of proofs of the relevant property restricted to
$\{x \mid \varphi(x)\}$.  Since all Python runtime objects have
decidable equality (Hedberg's theorem), all coefficient types are sets
(h-sets).  For sheaves of sets on a Boolean site, $\Hone = 0$ by the
acyclicity of Boolean algebras.  Hence the Kan filling condition is
automatically satisfied, and $\hfilling$ always exists when the face
data is consistent.
\end{proof}

\begin{corollary}[No Obstructions in the Python Setting]
\label{cor:no-obstruction}
For all covers arising from Z3-decidable refinement predicates on
$\PyObj$ types, $\Hone(\covers, \presheaf) = 0$.  Therefore every
consistent descent datum (local sections with overlap compatibility)
assembles to a unique global section.  Equivalently, every consistent
set of partial boundary data can be Kan-filled.
\end{corollary}

\begin{proof}
Follows from Hedberg's theorem (types with decidable equality are sets)
and the acyclicity of sheaves of sets on Boolean sites.  The Z3
decidability of $\varphi_k$ guarantees the Boolean algebra structure.
\end{proof}

\begin{remark}[Why This Is the Opening Theorem]
The identification in Theorem~\ref{thm:hcomp-is-descent} is the
organizing principle of the entire $\mathrm{C}^4$ framework.  Every
subsequent construction in this chapter is either:
\begin{itemize}
\item a special case of descent (Kan filling, path composition, GluePath),
  or
\item a structure that is defined over both dimension systems simultaneously
  (connection algebra, fiber-preserving homotopies, fiberwise univalence).
\end{itemize}
Treating HComp as a ``primitive'' separate from descent, as earlier
presentations did, obscures this unity.
\end{remark}


% ─────────────────────────────────────────────────────────
\section{The Interval and Its Connection Algebra}
\label{sec:interval}

\begin{definition}[The Interval]
The \textbf{interval} $\interval$ is a type with two distinguished points
$\mathsf{0}_\interval, \mathsf{1}_\interval : \interval$ and three
connection operations:
\begin{align}
  \_ \wedge \_ &: \interval \times \interval \to \interval
    \tag{meet / min} \\
  \_ \vee \_ &: \interval \times \interval \to \interval
    \tag{join / max} \\
  \sim &: \interval \to \interval
    \tag{reversal}
\end{align}
subject to the following equalities, which hold definitionally:
\begin{enumerate}
\item $\sim \mathsf{0} = \mathsf{1}$, \quad $\sim \mathsf{1} = \mathsf{0}$,
  \quad $\sim(\sim r) = r$.
\item $r \wedge \mathsf{0} = \mathsf{0}$, \quad $r \wedge \mathsf{1} = r$,
  \quad $r \vee \mathsf{0} = r$, \quad $r \vee \mathsf{1} = \mathsf{1}$.
\item $r \wedge s = s \wedge r$, \quad $r \vee s = s \vee r$.
\item $(r \wedge s) \wedge t = r \wedge (s \wedge t)$, \quad
  $(r \vee s) \vee t = r \vee (s \vee t)$.
\item De Morgan: $\sim(r \wedge s) = (\sim r) \vee (\sim s)$, \quad
  $\sim(r \vee s) = (\sim r) \wedge (\sim s)$.
\end{enumerate}
\end{definition}

\begin{remark}
The connection algebra $(\interval, \wedge, \vee, \sim, \mathsf{0},
\mathsf{1})$ forms a \textbf{De Morgan algebra}.  This is the same structure
used by Cohen--Coquand--Huber--M\"ortberg (CCHM)~\cite{cohen2018}.  The key
property is that $\interval$ is \emph{not} a type with decidable equality ---
there is no function $\interval \to \mathsf{2}$ separating $\mathsf{0}$ from
$\mathsf{1}$.

This De Morgan algebra is not the only one in $\mathrm{C}^4$.  The face
lattice $\mathbb{F}$ of refinement predicates $\varphi(x)$ also forms a De
Morgan algebra, with $\wedge$ = conjunction, $\vee$ = disjunction, and
$\neg$ = negation.  That both structures are De Morgan algebras is
\emph{not} a coincidence: it is the algebraic reflection of the
identification in Theorem~\ref{thm:hcomp-is-descent}.  See
Section~\ref{sec:connections-cover-algebra} for the formal correspondence.
\end{remark}

\begin{definition}[Face Lattice and Refinement Predicates]
\label{def:face-lattice}
A \textbf{face formula} $\varphi$ is a finite conjunction/disjunction of
atomic constraints:
\[
  \varphi, \psi ::= (r = \mathsf{0}) \mid (r = \mathsf{1})
    \mid \varphi \wedge \psi \mid \varphi \vee \psi \mid \mathsf{1}_\mathbb{F}
    \mid \mathsf{0}_\mathbb{F}
\]
We write $\mathbb{F}$ for the \textbf{face lattice} of all face formulas.
A context $\Gamma, \varphi$ extends $\Gamma$ with the constraint $\varphi$,
restricting to the face where $\varphi$ holds.

In the cohomological dimension, the same lattice structure arises
from Z3-decidable \textbf{refinement predicates} $\varphi(x)$ on the
input variable $x$.  Both are De Morgan algebras; the identification
of Theorem~\ref{thm:hcomp-is-descent} translates one into the other.
Site objects in $\mathrm{C}^4$ are always refinement predicates
$\varphi(x)$ with Z3-decidable satisfiability and entailment.  We
write $\site$ for the category of Z3-decidable refinement predicates
on $\PyObj$, with morphisms $\varphi \leq \psi$ (i.e., $\varphi \Rightarrow
\psi$, Z3-verified).
\end{definition}

\begin{definition}[Partial Elements]
Given $\Gamma \vdash A : \mathcal{U}_i$ and a face formula $\varphi$, a
\textbf{partial element} of $A$ along $\varphi$ is a term
$\Gamma, \varphi \vdash u : A$.  We write this as
$u : A\,[\varphi]$ and call $\varphi$ the \textbf{extent} of $u$.
\end{definition}


% ─────────────────────────────────────────────────────────
\section{Path Types in Detail}
\label{sec:path-types}

\begin{definition}[Path Type]
For $A : \mathcal{U}_i$ and $a, b : A$, the \textbf{path type} is:
\[
  \Path_A\, a\, b \;\defeq\;
  \{p : \interval \to A \mid p\,\mathsf{0} \equiv a,\;
    p\,\mathsf{1} \equiv b\}
\]
\end{definition}

\noindent
Path types internalize propositional equality as continuous deformation.
The typing rules from Chapter~\ref{ch:c4-calculus} give:

\begin{gather}
\frac{\Gamma, i : \interval \vdash t : A \qquad
      t[i := \mathsf{0}] \equiv a \qquad t[i := \mathsf{1}] \equiv b}
     {\Gamma \vdash \langle i \rangle\, t : \Path_A\, a\, b}
\tag{PathI}
\\[6pt]
\frac{\Gamma \vdash p : \Path_A\, a\, b \qquad
      \Gamma \vdash r : \interval}
     {\Gamma \vdash p\; r : A}
\tag{PathE}
\end{gather}

with the definitional equalities:
\begin{align}
  (\langle i \rangle\, t)\; r &\equiv t[i := r]
    & \text{($\beta$ for paths)} \tag{Path-$\beta$} \\
  \langle i \rangle\, (p\; i) &\equiv p
    & \text{($\eta$ for paths, when $i \notin \mathrm{FV}(p)$)}
    \tag{Path-$\eta$}
\end{align}

\begin{definition}[Path Operations]
\label{def:path-ops}
Using the connection algebra, we define:
\begin{enumerate}
\item \textbf{Constant path (reflexivity):}
  $\mathsf{refl}_a \defeq \langle i \rangle\, a : \Path_A\, a\, a$.

\item \textbf{Symmetry (path reversal):}
  Given $p : \Path_A\, a\, b$, define
  $p^{-1} \defeq \langle i \rangle\, p\,(\sim i) : \Path_A\, b\, a$.

\item \textbf{Connection terms:}
  Given $p : \Path_A\, a\, b$, the \textbf{connections} are:
  \begin{align*}
    p^{\wedge} &\defeq \langle i \rangle\langle j \rangle\,
      p\,(i \wedge j) &&: \Path_{\Path_A\, a\, b}\,
      \mathsf{refl}_a\, p \\
    p^{\vee} &\defeq \langle i \rangle\langle j \rangle\,
      p\,(i \vee j) &&: \Path_{\Path_A\, a\, b}\,
      p\, \mathsf{refl}_b
  \end{align*}
\end{enumerate}
\end{definition}

\begin{proposition}[Groupoid Laws]
\label{prop:groupoid}
For paths $p : \Path_A\, a\, b$, $q : \Path_A\, b\, c$,
$r : \Path_A\, c\, d$, the following hold (as paths of paths):
\begin{enumerate}
\item $p \cdot \mathsf{refl}_b \sim p$ \quad (right unit)
\item $\mathsf{refl}_a \cdot p \sim p$ \quad (left unit)
\item $p \cdot p^{-1} \sim \mathsf{refl}_a$ \quad (right inverse)
\item $(p \cdot q) \cdot r \sim p \cdot (q \cdot r)$ \quad (associativity)
\end{enumerate}
where $p \cdot q$ denotes path composition defined via $\hfilling$.
\end{proposition}

\begin{proof}
Each identity is constructed using the connection algebra.  For (1),
the term $\langle i\rangle\langle j\rangle\, p\,(i \vee j)$ provides
a path from $p \cdot \mathsf{refl}_b$ to $p$: at $j = \mathsf{0}$ we
get $p \cdot \mathsf{refl}_b$ (by the definition of composition using
$\hfilling$) and at $j = \mathsf{1}$ we get $\langle i \rangle\, p\,i = p$.
The remaining identities follow by analogous two-dimensional constructions
using $\wedge$, $\vee$, and $\sim$.
\end{proof}


% ─────────────────────────────────────────────────────────
\section{Connections and the Cover Algebra}
\label{sec:connections-cover-algebra}

This section proves that the connection algebra $(\wedge, \vee, \sim)$ on the
interval $\interval$ and the cover algebra $(\wedge, \vee, \neg)$ on
Z3-decidable refinement predicates are not merely analogous De Morgan algebras
--- they are formally identified under the translation of
Theorem~\ref{thm:hcomp-is-descent}.  This identification is why cover
simplification (Synergy~13 of the atlas) is a theorem, not a heuristic.

\begin{theorem}[De Morgan Isomorphism between Dimension Systems]
\label{thm:deMorgan-iso}
The map $\Phi : \mathbb{F}_{\mathrm{cub}} \to \mathbb{F}_{\mathrm{pred}}$
defined by:
\begin{align*}
  \Phi(i = \mathsf{0}) &\defeq \varphi_i(x)
    & &\text{(the refinement predicate associated to face $i = 0$)} \\
  \Phi(i = \mathsf{1}) &\defeq \neg\varphi_i(x) \\
  \Phi(\varphi \wedge \psi) &\defeq \Phi(\varphi) \wedge \Phi(\psi) \\
  \Phi(\varphi \vee \psi) &\defeq \Phi(\varphi) \vee \Phi(\psi) \\
  \Phi(\mathsf{1}_\mathbb{F}) &\defeq \mathsf{True} \\
  \Phi(\mathsf{0}_\mathbb{F}) &\defeq \mathsf{False}
\end{align*}
is a homomorphism of De Morgan algebras.  In particular, the De Morgan laws
for $\interval$:
\[
  \sim(r \wedge s) = (\sim r) \vee (\sim s), \qquad
  \sim(r \vee s) = (\sim r) \wedge (\sim s)
\]
translate, under $\Phi$, to the De Morgan laws for refinement predicates:
\[
  \neg(\varphi \wedge \psi) \Leftrightarrow (\neg\varphi) \vee (\neg\psi),
  \qquad
  \neg(\varphi \vee \psi) \Leftrightarrow (\neg\varphi) \wedge (\neg\psi)
\]
which are Z3-valid tautologies.
\end{theorem}

\begin{proof}
The map $\Phi$ is defined by structural recursion on face formulas.  It
clearly preserves the constants ($\mathsf{1}$ and $\mathsf{0}$) and the
binary operations.  The De Morgan laws for $\interval$ hold
\emph{definitionally} (they are axioms of the cubical type theory).  The De
Morgan laws for refinement predicates hold propositionally and are Z3-valid.
The map $\Phi$ sends definitional equalities on face formulas to Z3-provable
propositional equalities on refinement predicates.

For the reversal / negation correspondence: under $\Phi$, the reversal
$\sim(i = \mathsf{0}) = (i = \mathsf{1})$ becomes
$\neg \varphi_i \Leftrightarrow \neg\varphi_i$ (trivially).  The De Morgan
laws follow by structural induction on the formula.
\end{proof}

\begin{corollary}[Cover Simplification is Algebraically Sound]
\label{cor:cover-simplification}
The following cover simplifications are instances of the De Morgan algebra
laws, and are therefore:
\begin{enumerate}
\item \emph{Valid}: they produce covers that are equivalent in the sense
  that the descent constructions over them are isomorphic.
\item \emph{Z3-decidable}: Z3 can verify each simplification as a valid
  propositional formula.
\item \emph{Homotopy-invariant}: the simplifications preserve the homotopy
  type of the $\Cech$ nerve, so they do not change which Kan filling operations
  are possible.
\end{enumerate}
The simplifications are:
\begin{enumerate}
\item \textbf{Absorption}: $\varphi \wedge (\varphi \vee \psi) = \varphi$
  \quad (collapse nested covers).
\item \textbf{Distribution}: $\varphi \wedge (\psi \vee \chi)
  = (\varphi \wedge \psi) \vee (\varphi \wedge \chi)$
  \quad (factor product covers).
\item \textbf{Idempotence}: $\varphi \wedge \varphi = \varphi$
  \quad (deduplicate fibers).
\item \textbf{Complementation}: $\varphi \wedge \neg\varphi = \mathsf{False}$
  \quad (detect dead branches by Z3).
\item \textbf{De Morgan}: $\neg(\varphi \wedge \psi) = \neg\varphi \vee
  \neg\psi$ \quad (push negation inward, simplifying else-branches).
\end{enumerate}
\end{corollary}

\begin{proof}
Each simplification is a valid identity in any Boolean algebra (which is a
special case of a De Morgan algebra with $\varphi \vee \neg\varphi = \top$).
By Theorem~\ref{thm:deMorgan-iso}, they correspond to valid identities in the
cubical face lattice.  Homotopy invariance follows from the fact that the $\Cech$
nerve depends only on the \emph{lattice structure} of the cover --- two covers
related by these identities have isomorphic nerves (this is a standard result
in simplicial topology).
\end{proof}

\begin{example}[Cover Simplification for an \texttt{elif} Chain]
Consider a Python function:
\begin{verbatim}
def classify(x: int) -> str:
    if x > 0:            # phi_1 = (x > 0)
        return "pos"
    elif x < 0:          # phi_2 = (x < 0)
        return "neg"
    else:                # phi_3 = (x == 0)
        return "zero"
\end{verbatim}
The raw cover is $\covers = \{\varphi_1, \varphi_2, \varphi_3\}$ where
$\varphi_3 = \neg\varphi_1 \wedge \neg\varphi_2$.  Z3 verifies:
\begin{itemize}
\item \textbf{Disjointness}: $\varphi_1 \wedge \varphi_2 = \mathsf{False}$
  (empty overlap: no compatibility proof needed).
\item \textbf{Exhaustiveness}: $\varphi_1 \vee \varphi_2 \vee \varphi_3
  = \mathsf{True}$.
\item \textbf{Simplification of $\varphi_3$}: by De Morgan and Z3,
  $\neg(x > 0) \wedge \neg(x < 0) \Leftrightarrow (x = 0)$,
  which can be further simplified by Z3 to the single predicate $(x = 0)$.
\end{itemize}
Each of these steps is a De Morgan algebra law applied to refinement
predicates.  The cubical counterpart: the three-face cube $[i_1 = 0]$,
$[i_2 = 0]$, $[i_3 = 0]$ has disjoint faces (no shared edges), and the
face formula simplification reduces the cube to a triangle.
\end{example}

\begin{proposition}[Connection Terms as Cover Operations]
\label{prop:connections-as-cover-ops}
Under the isomorphism $\Phi$ of Theorem~\ref{thm:deMorgan-iso}, the
cubical connection operations on path terms correspond to the following
cover algebra operations:
\begin{align*}
  r \wedge s \;\longmapsto\; \varphi_r \wedge \varphi_s
    & \quad \text{(product cover)} \\
  r \vee s \;\longmapsto\; \varphi_r \vee \varphi_s
    & \quad \text{(join cover)} \\
  \sim r \;\longmapsto\; \neg \varphi_r
    & \quad \text{(complementary fiber)}
\end{align*}
In particular, the connection term $p^\wedge \defeq \langle i\rangle
\langle j \rangle\, p\,(i \wedge j)$ (which gives a path from
$\mathsf{refl}_a$ to $p$) corresponds to the descent data on the product
fiber $\varphi_i \wedge \varphi_j$ --- i.e., refining to the intersection.
\end{proposition}

\begin{proof}
This is an immediate consequence of the structural definition of $\Phi$:
the operations are preserved homomorphically.  The interpretations of
connection terms as cover operations follow from the identification in
Theorem~\ref{thm:hcomp-is-descent}: connection terms in the cubical
direction are just applications of the cover algebra in the
cohomological direction.
\end{proof}


% ─────────────────────────────────────────────────────────
\section{Kan Operations as Special Cases of Descent}
\label{sec:kan-ops}

In the original presentation, $\hfilling$ and $\transp$ are given as
\emph{primitive} operations.  We now re-derive them as special cases of the
descent rule, clarifying their place in the unified framework.

\begin{proposition}[$\hfilling$ is One-Dimensional Descent]
\label{prop:hfill-is-descent}
Homogeneous composition $\hfilling^i\, A\, [\varphi \mapsto u]\, u_0$
is the descent rule $\mathrm{desc}$ applied in the cubical dimension $i$,
over the cover:
\[
  \covers_{\varphi, u_0} \defeq
  \{\varphi\} \cup \{(i = \mathsf{0})\}
\]
with local sections:
\begin{itemize}
\item On $\varphi$: the section $\lambda i.\, u$ (the partial path).
\item On $(i = \mathsf{0})$: the section $u_0$ (the base point).
\end{itemize}
Overlap compatibility on $\varphi \wedge (i = \mathsf{0})$ is the
hypothesis $u[\mathsf{0}/i] \equiv u_0[\varphi]$.
\end{proposition}

\begin{proof}
This is exactly the translation of Definition~\ref{def:hcomp} through the
isomorphism of Theorem~\ref{thm:hcomp-is-descent}.  The output of
$\mathrm{desc}$ on this cover/data is the unique global section, whose value
at $i = \mathsf{1}$ gives the result of $\hfilling$.
\end{proof}

\begin{definition}[Transport --- $\transp$]
\label{def:c4-transp}
Given a line of types $\Gamma, i : \interval \vdash A : \mathcal{U}_j$
and a term $\Gamma \vdash t : A[\mathsf{0}/i]$:
\[
\frac{\Gamma, i : \interval \vdash A : \mathcal{U}_j \qquad
      \Gamma \vdash t : A[\mathsf{0}/i]}
     {\Gamma \vdash \transp^i\, A\, t : A[\mathsf{1}/i]}
\tag{transp}
\]
When $A$ is constant in $i$ (i.e., $i \notin \mathrm{FV}(A)$), transport
is the identity: $\transp^i\, A\, t \equiv t$.

\medskip\noindent\textbf{Cohomological reading.}  Transport $\transp^i\, A\, t$
is a degenerate case of descent: it moves a section from one refinement fiber
to another along a fiber implication.  Specifically, if $A(0)$ corresponds to
the type of proofs under $\varphi$ and $A(1)$ to the type of proofs under
$\psi$, and the path $A$ witnesses $\varphi \Rightarrow \psi$ (or vice versa),
then $\transp^i\, A\, t$ moves the proof $t$ from $\varphi$ to $\psi$.  This
is exactly the Z3-checked predicate implication step of Synergy~5.
\end{definition}

\begin{proposition}[Transport Along Refinement Implication]
\label{prop:transp-implication}
Let $\varphi_1, \varphi_2$ be Z3-decidable refinement predicates with
$\varphi_2 \Rightarrow \varphi_1$ (Z3-verified).  Let
$q : \Path_\Type\, \{x \mid \varphi_1\}\, \{x \mid \varphi_2\}$
be the path given by applying $\mathsf{ua}$ to the equivalence
induced by the inclusion $\{x \mid \varphi_2\} \hookrightarrow
\{x \mid \varphi_1\}$.  Then:
\[
  \transp^i\, (\lambda i.\, P\,(q\,i))\, p_1 : P(\{x \mid \varphi_2\})
\]
transports a proof $p_1 : P(\{x \mid \varphi_1\})$ to a proof valid under
the stronger hypothesis $\varphi_2$.  This is a zero-cost transport: the
verification obligation is the single Z3 check $\varphi_2 \Rightarrow
\varphi_1$.
\end{proposition}

\begin{proof}
By the rules for $\transp$ and univalence (Theorem~\ref{thm:univalence}),
transporting along $\mathsf{ua}(e)$ is functorial with respect to the
equivalence $e$.  The inclusion $\{x \mid \varphi_2\} \hookrightarrow
\{x \mid \varphi_1\}$ (valid because $\varphi_2 \Rightarrow \varphi_1$)
is an injective map into the proof type, giving the transport.
\end{proof}

\begin{definition}[Path Composition via $\hfilling$ as Two-Step Descent]
\label{def:path-comp}
Given $p : \Path_A\, a\, b$ and $q : \Path_A\, b\, c$, their
\textbf{composition} $p \cdot q : \Path_A\, a\, c$ is:
\[
  p \cdot q \;\defeq\; \langle i \rangle\,
    \hfilling^j\, A\,
    [(i = \mathsf{0}) \mapsto a,\;
     (i = \mathsf{1}) \mapsto q\,j]\,
    (p\,i)
\]
where $j$ is fresh.  At $i = \mathsf{0}$ this gives $a$; at
$i = \mathsf{1}$ it gives $\hfilling^j\, A\, [\cdots]\, b$ which
computes to $q\,\mathsf{1} = c$.

\medskip\noindent\textbf{Cohomological reading.}  Path composition is a
\emph{two-step} descent in the cubical direction: the proof moves from $a$
to $b$ (first fiber $p$) and then from $b$ to $c$ (second fiber $q$).
The $\hfilling$ at $j$ is the descent gluing at the ``overlap'' where
both fibers share the midpoint $b$.
\end{definition}

\noindent
\textbf{Computation rules for Kan operations on type formers.}
The Kan operations compute structurally on each type former:

\begin{enumerate}
\item \textbf{$\Pi$-types:}
  $\transp^i\, (\Pi(x:A).\,B)\, f \equiv
   \lambda(x : A[\mathsf{1}/i]).\,
   \transp^i\, B[x := \transp^i\, (\sim A)\, x]\,
   (f\, (\transp^i\, (\sim A)\, x))$

\item \textbf{$\Sigma$-types:}
  $\transp^i\, (\Sigma(x:A).\,B)\, (a, b) \equiv
   (a', \transp^i\, B[x := p]\, b)$
  where $a' = \transp^i\, A\, a$ and $p$ is the transport path.

\item \textbf{Path types:}
  Transport in $\Path_A\, a\, b$ composes paths using $\hfilling$.

\item \textbf{Natural numbers:} $\transp^i\, \Nat\, n \equiv n$
  (discrete type).

\item \textbf{Inductive types:} Transport applies constructor-wise,
  preserving the recursive structure.

\item \textbf{Refinement types:}
  $\transp^i\, (\{x : A \mid \varphi_i(x)\})\, (a, p) \equiv
   (\transp^i\, A\, a,\; \transp^i\, (\lambda i.\, \varphi_i(a_i))\, p)$
  where $a_i$ is the path of witnesses.
  This is the \emph{fiber transport} of Proposition~\ref{prop:transp-implication}.
\end{enumerate}


% ─────────────────────────────────────────────────────────
\section{GluePath: Descent over the Interval}
\label{sec:gluepath}

GluePath (Synergy~9 of the atlas) is a proof term that glues local
\emph{paths} --- not just point values --- via $\Cech$ descent.  It
is the most direct expression of the product space $\interval^m \times
\site^n$: the $\Cech$ data itself varies continuously over the cubical
interval.  GluePath is strictly more expressive than either cubical
composition or static descent alone.

\begin{definition}[GluePath Data]
\label{def:gluepath-data}
Let $f, g : A$ and $\covers = \{\varphi_1, \ldots, \varphi_n\}$ be an
exhaustive cover of Z3-decidable refinement predicates.  A
\textbf{GluePath datum} consists of:
\begin{enumerate}
\item \textbf{Local paths}: for each $k$, a term
  \[
    p_k : \interval \to \Path\bigl(\restr{f}{\varphi_k},\;
    \restr{g}{\varphi_k}\bigr)
  \]
  i.e., a path in the path space, restricted to the $k$-th fiber.

\item \textbf{Overlap coherence}: for each pair $k, \ell$ with
  $\varphi_k \wedge \varphi_\ell$ Z3-satisfiable, and for each
  $t : \interval$:
  \[
    \restr{p_k(t)}{\varphi_k \wedge \varphi_\ell}
    \;=_{\Path}\;
    \restr{p_\ell(t)}{\varphi_k \wedge \varphi_\ell}
  \]
  (the local paths agree on overlaps, \emph{for all} $t : \interval$).
\end{enumerate}
\end{definition}

\begin{theorem}[GluePath Descent]
\label{thm:gluepath}
Given GluePath data $(\covers, \{p_k\})$ satisfying the overlap coherence
condition, there exists a global path:
\[
  \mathsf{GluePath}(\covers, \{p_k\}) : \Path(f, g)
\]
that restricts to $p_k$ on each fiber $\varphi_k$:
\[
  \restr{\mathsf{GluePath}(\covers, \{p_k\})}{\varphi_k} \equiv p_k
\]
\end{theorem}

\begin{proof}
We apply the descent rule (Definition~\ref{def:descent-rule}) to the
sheaf $\presheaf$ that assigns, to each refinement predicate $\varphi$,
the type:
\[
  \presheaf(\varphi) \;\defeq\; \interval \to \Path\bigl(\restr{f}{\varphi},\;
  \restr{g}{\varphi}\bigr)
\]
The local sections are the $p_k$, and the overlap coherence condition is
exactly the descent compatibility condition for $\presheaf$.  Applying
the descent rule produces a global section
\[
  P : \interval \to \Path(f, g)
\]
Define $\mathsf{GluePath}(\covers, \{p_k\}) \defeq P(1) : \Path(f, g)$.
The restriction property holds by the boundary law of the descent rule.
\end{proof}

\begin{remark}[Why GluePath is Not Just Descent]
Standard $\Cech$ descent (Definition~\ref{def:descent-rule}) glues
\emph{point values}: $s_k \in \presheaf(\varphi_k)$.  GluePath glues
\emph{path-valued sections}: $p_k : \interval \to \presheaf(\varphi_k)$.
The difference is the extra dimension $\interval$: the $\Cech$ data is not
static but moves continuously over the cubical interval.  This is what
``descent over the interval'' means, and it is what the product space
$\interval^m \times \site^n$ formally encodes.

In the spectral sequence interpretation: GluePath computes the $E_1$ page
--- it glues local $H^0$ path-sections into a global $H^0$.  The overlap
coherence condition is the cocycle condition for the $E_2$ differential.
\end{remark}

\begin{example}[GluePath for a Branching Equivalence Proof]
Consider proving that two functions $f, g : \mathtt{int} \to \mathtt{int}$
are equal, where they have different implementations for positive and
non-positive inputs:
\[
  \varphi_1 = (x > 0), \qquad \varphi_2 = (x \leq 0)
\]
\begin{itemize}
\item Under $\varphi_1$: we have a proof $p_1 : \Path(\restr{f}{\varphi_1},
  \restr{g}{\varphi_1})$ via Z3 (or via an inductive argument).
\item Under $\varphi_2$: we have a proof $p_2 : \Path(\restr{f}{\varphi_2},
  \restr{g}{\varphi_2})$.
\item Overlap: $\varphi_1 \wedge \varphi_2 = (x > 0) \wedge (x \leq 0)
  = \mathsf{False}$, so the overlap coherence condition is vacuous (Z3
  detects the empty overlap).
\end{itemize}
Then $\mathsf{GluePath}(\{\varphi_1, \varphi_2\}, \{p_1, p_2\})
: \Path(f, g)$ is the global proof that $f = g$.  This is the standard
case-split proof, formalized as GluePath.

The GluePath structure makes explicit what is implicit in a case-split:
the proof lives in $\interval^1 \times \site^2$, with one cubical dimension
(the equality witness) and two cohomological dimensions (positive / non-positive).
\end{example}

\begin{example}[GluePath for a Refinement Homotopy]
Suppose we want to prove that two sorting algorithms $\mathtt{sort}_1$
and $\mathtt{sort}_2$ are homotopic: not just equal, but connected by a
\emph{path of paths} (a homotopy in proof space).  The cover is:
\[
  \varphi_{\mathrm{empty}} = (n = 0), \quad
  \varphi_{\mathrm{small}} = (1 \leq n \leq 10), \quad
  \varphi_{\mathrm{large}} = (n > 10)
\]
where $n = \mathtt{len}(xs)$.  A GluePath datum gives:
\begin{itemize}
\item For empty lists: $p_{\mathrm{empty}} : \interval \to \Path(\mathtt{sort}_1\,[], \mathtt{sort}_2\,[])$
  --- trivial (both return $[]$).
\item For small lists: $p_{\mathrm{small}} : \interval \to \Path(\restr{\mathtt{sort}_1}{\varphi_{\mathrm{small}}}, \restr{\mathtt{sort}_2}{\varphi_{\mathrm{small}}})$
  --- proved by Z3 over bounded inputs.
\item For large lists: $p_{\mathrm{large}} : \interval \to \Path(\restr{\mathtt{sort}_1}{\varphi_{\mathrm{large}}}, \restr{\mathtt{sort}_2}{\varphi_{\mathrm{large}}})$
  --- proved by induction on $n$ (an axiom chain in $\PyObj$).
\end{itemize}
The boundary overlaps $\varphi_{\mathrm{empty}} \cap \varphi_{\mathrm{small}}
= (n = 0) \cap (1 \leq n \leq 10) = \mathsf{False}$ and similarly for other
pairs, so all overlaps are vacuous.  $\mathsf{GluePath}$ assembles the global
proof that $\mathtt{sort}_1 = \mathtt{sort}_2$.
\end{example}


% ─────────────────────────────────────────────────────────
\section{Fiber-Preserving Homotopies}
\label{sec:fiber-homotopies}

A fiber-preserving homotopy (Synergy~20) is a path in proof space that
respects the fibration of types over the refinement site.  It is the
``moving'' version of GluePath: a parameterized family of descent proofs,
all in the product space $\interval^{m+1} \times \site^n$.

\begin{definition}[Fiber-Preserving Homotopy]
\label{def:fiber-homotopy}
Let $\covers = \{\varphi_1, \ldots, \varphi_n\}$ be an exhaustive cover
and let $f, g : A$.  A \textbf{fiber-preserving homotopy} from a proof
$p_0 : \Path(f, g)$ to a proof $p_1 : \Path(f, g)$ is a term:
\[
  H : \interval \to
  \prod_{k=1}^n \Path\bigl(\restr{f}{\varphi_k},\;\restr{g}{\varphi_k}\bigr)
\]
such that:
\begin{enumerate}
\item $H(\mathsf{0}) = \{p_0|_{\varphi_k}\}_k$ (the start descent data);
\item $H(\mathsf{1}) = \{p_1|_{\varphi_k}\}_k$ (the end descent data);
\item For all $t : \interval$ and all overlapping pairs $k, \ell$:
  $\restr{H(t)_k}{\varphi_k \wedge \varphi_\ell}
  = \restr{H(t)_\ell}{\varphi_k \wedge \varphi_\ell}$
  (the overlap coherence holds for each $t$, not just at the endpoints).
\end{enumerate}
\end{definition}

\begin{proposition}[Fiber-Preserving Homotopies as 2-Paths]
\label{prop:fiber-homotopy-2path}
A fiber-preserving homotopy $H$ from $p_0$ to $p_1$ corresponds
to a \textbf{2-path} (square) in $\mathrm{C}^4$:
\[
  \langle j \rangle \mathsf{GluePath}\bigl(\covers, \{H(j)_k\}\bigr)
  : \Path_{\Path(f,g)}\, p_0\, p_1
\]
where $j : \interval$ is the homotopy parameter.
\end{proposition}

\begin{proof}
By Theorem~\ref{thm:gluepath}, for each $j : \interval$, the data
$\{H(j)_k\}$ with its overlap coherence condition assembles to a global
path $P(j) : \Path(f, g)$.  At $j = \mathsf{0}$, $P(\mathsf{0}) = p_0$
(by the boundary condition on $H$); at $j = \mathsf{1}$,
$P(\mathsf{1}) = p_1$.  Hence $\langle j \rangle P(j) :
\Path_{\Path(f,g)}\, p_0\, p_1$.
\end{proof}

\begin{theorem}[Fiber-Preserving Homotopies Rule Out Accidental Proofs]
\label{thm:homotopy-rules-out-accidental}
Let $p, q : \Path(f, g)$ be two proofs of $f = g$.  If there exists a
fiber-preserving homotopy $H$ from $p$ to $q$, then:
\begin{enumerate}
\item The proofs $p$ and $q$ are \emph{connected} in proof space:
  there is a 2-path from $p$ to $q$.
\item The connection is \emph{uniform} over all input cases: the
  homotopy $H$ witnesses that the proof strategy, not just the conclusion,
  is continuously deformed from $p$ to $q$ on every fiber $\varphi_k$.
\end{enumerate}
In particular, a proof $p$ that is fiber-preserving-homotopic to a known
``canonical'' proof $q$ cannot be a coincidental proof: it must use the
same underlying reasoning on every branch.
\end{theorem}

\begin{proof}
Part (1) is Proposition~\ref{prop:fiber-homotopy-2path}.  Part (2) follows
from the definition of $H$: for each $j$, $H(j)$ is a family of local paths
satisfying overlap coherence.  The continuity of $H$ in $j$ (it is a term
in $\interval \to \cdots$) guarantees that the local reasoning varies
continuously: there is no discrete jump in proof strategy.  By contrast,
a proof that ``works by accident'' on some branch would require a
non-uniform deformation (a disconnected homotopy), which is impossible in
the cubical setting where all maps out of $\interval$ are continuous.
\end{proof}

\begin{example}[Non-Uniform Proof Detection]
Suppose we have two proofs of $\mathtt{abs}(x) \geq 0$ for all integers $x$:
\begin{itemize}
\item $p$: works by case split on the sign of $x$, using Z3 on each case.
\item $q$: works by $\mathtt{abs}(x) = \sqrt{x^2}$ and $\sqrt{\cdot} \geq 0$
  (valid only over $\mathbb{R}$, not $\mathbb{Z}$, but suppose an LLM oracle
  suggests it).
\end{itemize}
The refinement cover is $\varphi_+ = (x \geq 0)$, $\varphi_- = (x < 0)$.
Proof $p$ restricts cleanly to both fibers.  Proof $q$ fails to restrict
to $\varphi_-$ (the real-square-root argument is invalid for negative
integers in discrete arithmetic).  Therefore, no fiber-preserving homotopy
$H$ from $p$ to $q$ exists: the proofs are disconnected in the fibered
proof space.  This identifies $q$ as an \emph{accidental proof} --- it
cannot be used in the $\mathrm{C}^4$ framework.
\end{example}


% ─────────────────────────────────────────────────────────
\section{Overlap Compatibility and Cubical Edge Agreement}
\label{sec:overlap-edge}

Synergy~7 of the atlas identifies two conditions that appear separately in
sheaf theory and cubical type theory but are formally the same:

\begin{proposition}[Overlap Compatibility = Cubical Edge Agreement]
\label{prop:overlap-is-edge}
Let $\covers = \{\varphi_k\}$ be a cover and $\{s_k\}$ be local sections.
The \textbf{$\Cech$ overlap compatibility condition}:
\[
  \restr{s_k}{\varphi_k \wedge \varphi_\ell}
  \;=_{\Path}\;
  \restr{s_\ell}{\varphi_k \wedge \varphi_\ell}
\]
is definitionally equal, under the translation of
Theorem~\ref{thm:hcomp-is-descent}, to the \textbf{cubical edge agreement
condition}: on the shared edge $[i_k = \varepsilon_k, i_\ell =
\varepsilon_\ell]$ of the partial cube, the boundary data from face $k$
and face $\ell$ agree.
\end{proposition}

\begin{proof}
By Theorem~\ref{thm:hcomp-is-descent}, each $\varphi_k$ is identified with
a cubical face $[i_k = \varepsilon_k]$.  The restriction
$\restr{s_k}{\varphi_k \wedge \varphi_\ell}$ becomes, under this
identification, the evaluation of the partial path $u_k$ at the face
$[i_k = \varepsilon_k, i_\ell = \varepsilon_\ell]$.  Similarly for
$\restr{s_\ell}{\varphi_k \wedge \varphi_\ell}$.  The equality between
these two is exactly the edge agreement condition in the cubical setting.
The translation is functorial: it preserves the type-theoretic equality
$=_{\Path}$ on both sides.
\end{proof}

\begin{proposition}[Z3 Decides Empty Overlaps]
\label{prop:z3-empty-overlap}
For Z3-decidable refinement predicates $\varphi_k$ and $\varphi_\ell$,
Z3 can decide whether $\varphi_k \wedge \varphi_\ell$ is satisfiable.
If it is \emph{unsatisfiable}, the overlap compatibility condition is
vacuously true: no proof is required.

This corresponds, in the cubical setting, to a pair of faces with no
shared edge --- a condition that cannot arise in a standard cube but
can arise in the face system of a descent problem.
\end{proposition}

\begin{proof}
The Z3 decision procedure for linear arithmetic (and the other theories
supported in $\mathrm{C}^4$) correctly decides satisfiability of
Boolean combinations of arithmetic predicates.  If $\varphi_k \wedge
\varphi_\ell$ is Z3-unsatisfiable, the restriction
$\restr{s_k}{\varphi_k \wedge \varphi_\ell}$ is a term defined in the
empty context $\varphi_k \wedge \varphi_\ell \vdash \bot$, and the
compatibility condition holds ex falso.
\end{proof}

\begin{proposition}[Four Strategies for Automatic Compatibility]
\label{prop:auto-compat}
On non-empty overlaps $\varphi_k \wedge \varphi_\ell$ (Z3-satisfiable),
the following four strategies (in order of priority) handle overlap
compatibility without requiring manual proof:
\begin{enumerate}
\item \textbf{Z3 disjointness}: If Z3 proves $\varphi_k \wedge \varphi_\ell
  \equiv \mathsf{False}$, the compatibility is vacuous.
\item \textbf{Containment transport}: If Z3 proves $\varphi_k \Rightarrow
  \varphi_\ell$ (or vice versa), then the $\ell$-section restricts to a
  transport of the $k$-section.  By
  Proposition~\ref{prop:transp-implication}, this is a zero-cost transport.
\item \textbf{Reflexivity}: If both $s_k$ and $s_\ell$ are reflexivity
  paths ($\mathsf{refl}$), then their restrictions to the overlap are both
  reflexivity.  Compatibility is trivial.
\item \textbf{Z3 discharge}: If the compatibility obligation is a
  quantifier-free arithmetic formula, discharge it directly to Z3.
\end{enumerate}
In typical Python code (\texttt{if/elif/else} chains, \texttt{isinstance}
dispatches), strategies 1 and 3 handle all overlaps, so no manual
compatibility proofs are needed.
\end{proposition}

\begin{proof}
Strategy 1 is Proposition~\ref{prop:z3-empty-overlap}.  Strategy 2 follows
from Proposition~\ref{prop:transp-implication}: the containment gives a
fiber implication, and transport along the implication identifies the
sections on the overlap.  Strategy 3 is immediate: restrictions of $\mathsf{refl}$
are $\mathsf{refl}$.  Strategy 4 uses Z3 completeness for quantifier-free
linear arithmetic.
\end{proof}

\begin{example}[Overlap Analysis for an \texttt{if/elif/else} Chain]
The function:
\begin{verbatim}
def sign(x: int) -> int:
    if x > 0:    # phi_1 = (x > 0)
        return 1
    elif x < 0:  # phi_2 = (x < 0)
        return -1
    else:        # phi_3 = (x == 0)
        return 0
\end{verbatim}
generates three refinement predicates $\varphi_1, \varphi_2, \varphi_3$.
Z3 decides:
\begin{itemize}
\item $\varphi_1 \wedge \varphi_2 = (x > 0) \wedge (x < 0) = \mathsf{False}$ \quad (strategy 1).
\item $\varphi_1 \wedge \varphi_3 = (x > 0) \wedge (x = 0) = \mathsf{False}$ \quad (strategy 1).
\item $\varphi_2 \wedge \varphi_3 = (x < 0) \wedge (x = 0) = \mathsf{False}$ \quad (strategy 1).
\end{itemize}
All overlaps are empty.  No compatibility proofs are required.  The descent
data consists of three independent local proofs, and $\mathsf{GluePath}$
assembles them into a global proof.

In the cubical picture: the proof cube has three faces with no shared edges
(a degenerate cube), and $\hfilling$ trivially assembles the boundary data.
\end{example}


% ─────────────────────────────────────────────────────────
\section{Function Extensionality}
\label{sec:c4-funext}

\begin{theorem}[Function Extensionality]
\label{thm:c4-funext}
If $f, g : \Pi(x : A).\, B(x)$ and
$h : \Pi(x : A).\, \Path_{B(x)}\, (f\,x)\, (g\,x)$, then
there exists a path $\mathsf{funext}(h) : \Path_{\Pi(x:A).\,B(x)}\, f\, g$.
\end{theorem}

\begin{proof}
Define $\mathsf{funext}(h) \defeq \langle i \rangle\, \lambda(x : A).\,
h\,x\,i$.  We check the boundary:
\begin{itemize}
\item At $i = \mathsf{0}$:
  $(\langle i \rangle\, \lambda x.\, h\,x\,i)\,\mathsf{0}
   = \lambda x.\, h\,x\,\mathsf{0}
   = \lambda x.\, f\,x = f$.
\item At $i = \mathsf{1}$:
  $(\langle i \rangle\, \lambda x.\, h\,x\,i)\,\mathsf{1}
   = \lambda x.\, h\,x\,\mathsf{1}
   = \lambda x.\, g\,x = g$.
\end{itemize}
Hence $\mathsf{funext}(h) : \Path\, f\, g$.  This is a
\emph{definitional} proof --- no axiom is needed.
\end{proof}

\begin{remark}[Funext as GluePath in the Cohomological Direction]
Function extensionality has a natural GluePath interpretation.  Suppose
$A = \{x : \mathtt{int} \mid \top\}$ and the cover of $A$ is
$\covers = \{\varphi_k\}$.  Then the hypothesis
$h : \Pi(x : A).\, \Path_{B(x)}\, (f\,x)\, (g\,x)$ restricted to each
fiber $\varphi_k$ gives local paths:
\[
  h|_{\varphi_k} : \Pi(x : \{x \mid \varphi_k\}).\,
  \Path\,(f\,x)\,(g\,x)
\]
These are GluePath data.  $\mathsf{GluePath}(\covers, \{h|_{\varphi_k}\})$
gives the global path $\mathsf{funext}(h)$.  Function extensionality is
therefore a special case of GluePath where the cover is the trivial one
$\{\top\}$ and the interval parameter is the argument $x : A$.
\end{remark}

\begin{remark}[Function Extensionality vs.\ CIC]
Function extensionality is a theorem in CIC only with the axiom of
functional extensionality (Lean's \texttt{funext}).  In $\mathrm{C}^4$,
it is \emph{derived} from the cubical structure.  This means
$\mathrm{C}^4$ is strictly more powerful than CIC without funext,
while remaining conservative over CIC + funext.
\end{remark}

\begin{corollary}[Fiberwise Funext]
\label{cor:fiberwise-funext}
Let $\covers = \{\varphi_1, \ldots, \varphi_n\}$ be a cover of the
input domain.  If for each $k$, we have
$h_k : \Pi(x : \{x \mid \varphi_k\}).\, \Path_{B(x)}\,
(\restr{f}{\varphi_k}\,x)\, (\restr{g}{\varphi_k}\,x)$,
and the $h_k$ satisfy the GluePath overlap coherence condition, then:
\[
  \mathsf{GluePath}\bigl(\covers,\; \{\mathsf{funext}(h_k)\}\bigr)
  : \Path_{\Pi(x:A).\,B(x)}\, f\, g
\]
This is the \textbf{fiberwise function extensionality}: prove pointwise
equality on each fiber, then assemble the global extensionality witness.
\end{corollary}

\begin{proof}
Apply $\mathsf{funext}$ to each $h_k$ to get local paths
$\mathsf{funext}(h_k) : \Path\,(\restr{f}{\varphi_k})\,(\restr{g}{\varphi_k})$.
The overlap coherence condition ensures these are compatible.
Apply Theorem~\ref{thm:gluepath} to assemble the global path.
\end{proof}




% ─────────────────────────────────────────────────────────
\section{Glue Types and Univalence}
\label{sec:univalence}

Glue types are the mechanism by which univalence is proved in cubical type
theory.  In the $\mathrm{C}^4$ setting, they have a natural cohomological
reading: $\Glue$ is descent for \emph{type families}, and $\mathsf{ua}$
is descent applied to a path of equivalences.

\begin{definition}[Equivalence]
A function $f : A \to B$ is an \textbf{equivalence} if it has
contractible fibers:
\[
  \mathsf{isEquiv}(f) \;\defeq\;
  \Pi(b : B).\, \mathsf{isContr}\bigl(
    \Sigma(a : A).\, \Path_B\, (f\,a)\, b
  \bigr)
\]
We write $A \simeq B$ for the type of equivalences $\Sigma(f : A \to B).\,
\mathsf{isEquiv}(f)$.

\medskip\noindent\textbf{Cohomological reading.}  An equivalence
$e : A \simeq B$ between types over the refinement site is a
\emph{sheaf isomorphism}: a morphism of sheaves that is an isomorphism
fiberwise.  The contractibility of fibers corresponds to the uniqueness
of the isomorphism on each open.
\end{definition}

\begin{definition}[Glue Types]
\label{def:glue}
Given $\Gamma \vdash A : \mathcal{U}_i$, a face formula $\varphi$, and
along $\varphi$ a type $T : \mathcal{U}_i$ with an equivalence
$e : T \simeq A$:
\[
  \frac{\Gamma \vdash A : \mathcal{U}_i \qquad
        \Gamma, \varphi \vdash T : \mathcal{U}_i \qquad
        \Gamma, \varphi \vdash e : T \simeq A}
       {\Gamma \vdash \Glue\,[\varphi \mapsto (T, e)]\, A
         : \mathcal{U}_i}
  \tag{Glue}
\]
with the boundary law: when $\varphi$ holds,
$\Glue\,[\varphi \mapsto (T, e)]\, A \equiv T$.

\medskip\noindent\textbf{Cohomological reading.}  The $\Glue$ type is the
universal type that ``looks like'' $T$ on the refinement fiber $\varphi$
and like $A$ on the complement $\neg\varphi$, glued via the equivalence
$e$.  This is exactly the sheaf-theoretic glueing construction: given a
sheaf $T$ on the open $\{x \mid \varphi(x)\}$ and a sheaf $A$ on the
complement, with an identification $e$ on the overlap, the glued sheaf
is the $\Glue$ type.

In the $\mathrm{C}^4$ compiler: $\Glue\,[\varphi \mapsto (T, e)]\, A$
represents the type of a variable that has type $T$ under the Z3-decidable
predicate $\varphi$ and type $A$ otherwise, with $e$ witnessing that the
two types are equivalent on the boundary of $\varphi$.
\end{definition}

\begin{theorem}[Univalence]
\label{thm:univalence}
For any $A, B : \mathcal{U}_i$, the canonical map
\[
  \mathsf{idToEquiv} : \Path_{\mathcal{U}_i}\, A\, B \to (A \simeq B)
\]
defined by $\mathsf{idToEquiv}(p) \defeq \transp^i\, (p\,i)$ together
with its equivalence proof, is itself an equivalence.
\end{theorem}

\begin{proof}[Proof sketch]
We construct the inverse $\mathsf{ua} : (A \simeq B) \to
\Path_{\mathcal{U}_i}\, A\, B$ using Glue types.  Given an equivalence
$e : A \simeq B$, define:
\[
  \mathsf{ua}(e) \defeq \langle i \rangle\,
    \Glue\,[(i = \mathsf{0}) \mapsto (A, e),\;
                     (i = \mathsf{1}) \mapsto (B, \mathsf{id}_B)]\, B
\]
At $i = \mathsf{0}$ this is $A$ (glued to $B$ via $e$); at $i = \mathsf{1}$
this is $B$ (glued to $B$ via identity).  The round-trip
$\mathsf{idToEquiv} \circ \mathsf{ua} \sim \mathrm{id}$ and
$\mathsf{ua} \circ \mathsf{idToEquiv} \sim \mathrm{id}$ are verified
using the computation rules for $\transp$ and $\Glue$.

\medskip\noindent\textbf{Cohomological reading of $\mathsf{ua}$.}
The path $\mathsf{ua}(e)$ is a GluePath datum in the type universe: it
is a family of type identifications parameterized over $\interval$, with
the $\Cech$ data consisting of the two face types $A$ and $B$ glued via $e$.
This is descent applied to the universe sheaf $\mathcal{U}_i$, not just
to value sheaves.
\end{proof}

\begin{corollary}[Univalence Transfers Proofs Across Refinement Equivalences]
\label{cor:ua-transfer}
Let $\varphi$ and $\psi$ be Z3-decidable refinement predicates with
$\varphi \Leftrightarrow \psi$ (Z3-provable equivalence).  Then:
\[
  \mathsf{ua}(e_{\varphi\psi}) : \Path_\Type\,
  \{x \mid \varphi(x)\}\, \{x \mid \psi(x)\}
\]
where $e_{\varphi\psi}$ is the equivalence induced by the logical
equivalence.  Any proof $P(\{x \mid \varphi\})$ can be transferred to
$P(\{x \mid \psi\})$ by transport along $\mathsf{ua}(e_{\varphi\psi})$.
This is the \textbf{Z3-decidable type equality} of Synergy~2.
\end{corollary}

\begin{proof}
The logical equivalence $\varphi \Leftrightarrow \psi$ gives an
equivalence of types $e : \{x \mid \varphi\} \simeq \{x \mid \psi\}$
(the forward and backward implications compose to identities).
Univalence converts this to a path, and transport along the path
transfers proofs.
\end{proof}


% ─────────────────────────────────────────────────────────
\section{Path Composition and Axiom Chain Correspondence}
\label{sec:path-axiom}

This section proves the central theorem connecting the proof-theoretic
machinery of Chapters~\ref{ch:proof-theory-foundations}--\ref{ch:proof-theory-automation}
to the type-theoretic framework of $\mathrm{C}^4$.

\begin{definition}[Axiom Path]
\label{def:axiom-path}
For each CCTT axiom $D_k : \ell_k \equiv r_k$ (with $k \in \{1, \ldots, 24\}$
or from the extended axiom set), the path constructor of the $\PyObj$ HIT
gives a term:
\[
  \mathsf{D}_k : \Pi(\sigma : \mathsf{Subst}).\,
    \Path_{\mathcal{U}_\PyObj}\, \den{\ell_k\sigma}\, \den{r_k\sigma}
\]
We call $\mathsf{D}_k(\sigma)$ an \textbf{axiom path}.
\end{definition}

\begin{definition}[Axiom Chain]
\label{def:axiom-chain}
An \textbf{axiom chain} from $\den{f}$ to $\den{g}$ is a sequence
\[
  \den{f} = t_0 \xrightarrow{D_{i_1}(\sigma_1)} t_1
    \xrightarrow{D_{i_2}(\sigma_2)} t_2
    \xrightarrow{\quad\cdots\quad} t_{n-1}
    \xrightarrow{D_{i_n}(\sigma_n)} t_n = \den{g}
\]
where each step applies a (possibly reversed) axiom at a subterm position.
\end{definition}

\begin{theorem}[Path--Axiom Correspondence]
\label{thm:path-axiom}
Let $f, g$ be Python programs with OTerm denotations $\den{f}, \den{g}
\in \mathcal{U}_\PyObj$.
\begin{enumerate}
\item \textbf{Axiom chain $\Rightarrow$ path:}
  If there is an axiom chain $D_{i_1}(\sigma_1) \cdot D_{i_2}(\sigma_2)
  \cdots D_{i_n}(\sigma_n)$ from $\den{f}$ to $\den{g}$, then there
  exists a term $p : \Path_{\mathcal{U}_\PyObj}\, \den{f}\, \den{g}$
  constructible from the HIT path constructors and cubical path
  composition.

\item \textbf{Path $\Rightarrow$ axiom chain:}
  Conversely, if $p : \Path_{\mathcal{U}_\PyObj}\, \den{f}\, \den{g}$ is
  constructible from the HIT path constructors (without transport through
  non-trivial type families), then there exists a finite axiom chain from
  $\den{f}$ to $\den{g}$.
\end{enumerate}
\end{theorem}

\begin{proof}
\textbf{Part (1).}  We proceed by induction on the length $n$ of the
axiom chain.

\emph{Base case} ($n = 0$): $\den{f} \equiv \den{g}$, so
$\mathsf{refl}_{\den{f}} : \Path\, \den{f}\, \den{g}$.

\emph{Inductive step}: Suppose the chain has $n$ steps, with the
first step $D_{i_1}(\sigma_1) : \den{f} \equiv t_1$ applied at subterm
position $\pi$.  By the HIT structure, $D_{i_1}(\sigma_1)$ gives a
path $p_1 : \Path\, \den{\ell_{i_1}\sigma_1}\, \den{r_{i_1}\sigma_1}$.
To apply this at position $\pi$, we use congruence: if $\den{f} =
C[\ell_{i_1}\sigma_1]$ for a context $C[\cdot]$, then
$\mathsf{ap}_{C}(p_1) : \Path\, C[\ell_{i_1}\sigma_1]\,
C[r_{i_1}\sigma_1] = \Path\, \den{f}\, t_1$.  By the inductive
hypothesis, the remaining chain gives $q : \Path\, t_1\, \den{g}$.
Compose: $\mathsf{ap}_C(p_1) \cdot q : \Path\, \den{f}\, \den{g}$.

\textbf{Part (2).}  A path $p$ in the HIT $\PyObj$ that is built only
from HIT constructors is, by the elimination principle of HITs, a
composition of the generating paths $D_k(\sigma)$ and their inverses
(possibly applied under contexts via $\mathsf{ap}$).  Each such
composition step corresponds to an axiom application.  The elimination
principle gives a term in the free $\infty$-groupoid generated by the
path constructors; truncating to the 0-level (equality of OTerms)
gives a finite word in the axiom applications, i.e., an axiom chain.
\end{proof}

\begin{corollary}
The CCTT proof checker (which searches for axiom chains) is
\textbf{sound and complete} with respect to $\mathrm{C}^4$ path
equality in $\mathcal{U}_\PyObj$, restricted to HIT-constructible
paths.
\end{corollary}

\begin{proof}
Immediate from Theorem~\ref{thm:path-axiom}: the checker finds an
axiom chain if and only if a HIT-constructible path exists.
\end{proof}


% ─────────────────────────────────────────────────────────
\section{Connection to the PyComp HIT}
\label{sec:pycomp-hit}

The $\PyObj$ HIT from Chapter~\ref{ch:c4-calculus} has two classes of
constructors:

\begin{enumerate}
\item \textbf{Point constructors}: $\mathsf{OVar}$, $\mathsf{OLit}$,
  $\mathsf{OOp}$, $\mathsf{OFold}$, $\mathsf{OCase}$, $\mathsf{OFix}$,
  $\mathsf{OApp}$, $\mathsf{OLam}$, $\mathsf{OSeq}$, $\ldots$
  These are the syntax tree nodes of the OTerm language.

\item \textbf{Path constructors}: $\mathsf{D}_1, \ldots, \mathsf{D}_{24}$
  and the extended axioms $\mathsf{P}_1, \ldots, \mathsf{P}_{48}$.
  These identify OTerms that are semantically equal.
\end{enumerate}

\begin{definition}[Path Composition in $\PyObj$]
Given axiom paths $p_k : \Path_\PyObj\, s_{k-1}\, s_k$ for
$k = 1, \ldots, n$, the \textbf{chain composition} is:
\[
  p_1 \cdot p_2 \cdots p_n \;\defeq\;
  \underbrace{(\cdots((p_1 \cdot p_2) \cdot p_3) \cdots) \cdot p_n}_{
    \text{left-associated via Definition~\ref{def:path-comp}}}
  : \Path_\PyObj\, s_0\, s_n
\]
\end{definition}

\begin{example}[Factorial Equivalence as Path]
The proof that recursive factorial equals iterative factorial
(Example~1 in Chapter~\ref{ch:proof-theory-examples}) corresponds to a
path in $\PyObj$:
\begin{align*}
  &\den{\mathsf{fact\_rec}(n)} \\
  &\quad \xrightarrow{\mathsf{NatInd}(\text{base, step})}
    \mathsf{OFold}(\times, 1, \mathsf{range}(1, n{+}1)) \\
  &\quad \xrightarrow{D_{15}(\sigma)}
    \den{\mathsf{fact\_iter}(n)}
\end{align*}
In $\mathrm{C}^4$, this is a path:
\[
  p_{\mathsf{fact}} \defeq
  \mathsf{ap}_{\mathsf{OFix}}(\mathsf{nat\_ind\_path}) \cdot
  \mathsf{D}_{15}(\sigma)
  : \Path_\PyObj\, \den{\mathsf{fact\_rec}}\, \den{\mathsf{fact\_iter}}
\]
\end{example}

\begin{example}[Sum-Range Identity as Path]
The identity $\mathsf{sum}(\mathsf{range}(n)) = n(n-1)/2$ is a
two-step path:
\[
  \den{\mathsf{sum}(\mathsf{range}(n))}
  \xrightarrow{P_{12}}
  \mathsf{OFold}(+, 0, \mathsf{range}(n))
  \xrightarrow{D_{18}}
  \den{n(n-1)/2}
\]
In $\mathrm{C}^4$: $P_{12}(\sigma) \cdot D_{18}(\sigma') :
\Path_\PyObj\, \den{\mathsf{sum}(\mathsf{range}(n))}\,
\den{n(n-1)/2}$.
\end{example}


% ─────────────────────────────────────────────────────────
\section{Cubical and Cohomological Structure: Interaction Theorems}
\label{sec:cubical-fibers}

This section collects results about how cubical and cohomological structure
interact.  The key theme is that the two dimension systems are not just
analogous but operationally coupled: cubical operations on paths respect
the fibration over the refinement site, and cohomological operations on
covers respect the cubical structure of the proofs.

\begin{proposition}[Restriction Commutes with Path Operations]
\label{prop:restr-path}
For $p : \Path_A\, a\, b$ and a Z3-decidable refinement predicate
$\varphi(x)$ giving an open $U_\varphi \defeq \{x \mid \varphi(x)\}
\in \mathrm{Ob}(\site)$:
\begin{enumerate}
\item $\restr{p}{\varphi} : \Path_{\restr{A}{\varphi}}\,
  \restr{a}{\varphi}\, \restr{b}{\varphi}$
\item $\restr{(p \cdot q)}{\varphi} \equiv \restr{p}{\varphi}
  \cdot \restr{q}{\varphi}$
\item $\restr{(p^{-1})}{\varphi} \equiv (\restr{p}{\varphi})^{-1}$
\item $\restr{(\transp^i\, A\, t)}{\varphi} \equiv
  \transp^i\, \restr{A}{\varphi}\, \restr{t}{\varphi}$
\item $\restr{(\hfilling^i\, A\, [\psi \mapsto u]\, u_0)}{\varphi}
  \equiv \hfilling^i\, \restr{A}{\varphi}\,
  [(\psi \wedge \varphi) \mapsto \restr{u}{\varphi}]\,
  \restr{u_0}{\varphi}$
\end{enumerate}
\end{proposition}

\begin{proof}
(1) follows from the restriction computation rule (Path-res) in
Chapter~\ref{ch:c4-calculus}: $\restr{(\langle i \rangle\, t)}{\varphi}
\equiv \langle i \rangle\, \restr{t}{\varphi}$.  The boundary conditions
are preserved because restriction commutes with substitution.

(2)--(3) follow from (1) and the definition of composition/inversion
via $\hfilling$, since restriction distributes over $\hfilling$:
$\restr{(\hfilling^i\, A\, [\psi \mapsto u]\, u_0)}{\varphi} \equiv
\hfilling^i\, \restr{A}{\varphi}\, [(\psi \wedge \varphi) \mapsto
\restr{u}{\varphi}]\, \restr{u_0}{\varphi}$, which is (5).

(4) is analogous, using the distributivity of restriction over $\transp$.
\end{proof}

\begin{theorem}[Fiberwise Univalence]
\label{thm:fiberwise-ua}
Let $\covers = \{\varphi_\alpha\}_\alpha$ be an exhaustive cover of
Z3-decidable refinement predicates.  Suppose we have a family of
fiberwise equivalences:
\[
  e_\alpha : \restr{A}{\varphi_\alpha} \simeq \restr{B}{\varphi_\alpha}
\]
that is compatible on overlaps:
\[
  \restr{e_\alpha}{\varphi_\alpha \wedge \varphi_\beta}
  = \restr{e_\beta}{\varphi_\alpha \wedge \varphi_\beta}
\]
(Z3 can check this condition for Z3-decidable equivalences.)
Then the fiberwise $\mathsf{ua}$ paths:
\[
  \mathsf{ua}(e_\alpha) : \Path_{\mathcal{U}_i}\,
  \restr{A}{\varphi_\alpha}\, \restr{B}{\varphi_\alpha}
\]
satisfy the GluePath overlap coherence condition (they agree on overlaps),
and:
\[
  \mathsf{GluePath}\bigl(\covers, \{\mathsf{ua}(e_\alpha)\}\bigr)
  : \Path_{\mathcal{U}_i^{\site}}\, A\, B
\]
provided $\Hone(\covers, \underline{\mathrm{Aut}}_{\mathcal{U}_i}) = 0$,
which holds by Corollary~\ref{cor:no-obstruction}.
\end{theorem}

\begin{proof}
Each $e_\alpha$ gives a path $\mathsf{ua}(e_\alpha) :
\Path\, \restr{A}{\varphi_\alpha}\, \restr{B}{\varphi_\alpha}$ by
Theorem~\ref{thm:univalence}.  The compatibility on overlaps ensures
that the restrictions of these paths agree:
$\restr{\mathsf{ua}(e_\alpha)}{\varphi_\alpha \wedge \varphi_\beta}
\equiv \restr{\mathsf{ua}(e_\beta)}{\varphi_\alpha \wedge \varphi_\beta}$
(by univalence applied to the equal equivalences).  This is exactly the
GluePath overlap coherence condition (Definition~\ref{def:gluepath-data}).
By Theorem~\ref{thm:gluepath} and Corollary~\ref{cor:no-obstruction}, the
local paths assemble into a global path.
\end{proof}

\begin{remark}[Fiberwise Univalence in the Compiler]
Theorem~\ref{thm:fiberwise-ua} is the theoretical basis for the $\mathrm{C}^4$
compiler's \textbf{refactoring soundness} guarantee.  When a programmer
refactors a function $A$ into a function $B$ (by changing the implementation
within each input branch), the compiler:
\begin{enumerate}
\item For each branch $\varphi_k$, verifies the local equivalence
  $e_k : \restr{A}{\varphi_k} \simeq \restr{B}{\varphi_k}$
  (Z3 check on the branch behavior).
\item Verifies overlap compatibility (Z3 check, usually trivially empty).
\item Applies $\mathsf{GluePath}(\covers, \{\mathsf{ua}(e_k)\})$ to produce
  the global path $\Path_{\mathcal{U}^\site}\, A\, B$.
\item All existing proofs about $A$ are transferred to $B$ via transport
  along this global path.
\end{enumerate}
This is the full cycle of the refactoring guarantee, made possible by
the interaction of fiberwise univalence and GluePath descent.
\end{remark}

% ─────────────────────────────────────────────────────────
\section{Summary: The Unified Framework}
\label{sec:c4-cubical-summary}

This chapter has developed the cubical structure of $\mathrm{C}^4$ from the
perspective of the synergy atlas, making explicit that the cubical and
cohomological components are not independent --- they are the same universal
construction viewed from two directions.

\begin{theorem}[Summary of Identifications]
\label{thm:summary-identifications}
The following pairs of concepts from cubical type theory and sheaf
cohomology are formally identified in $\mathrm{C}^4$:
\begin{center}
\begin{tabular}{p{5.5cm}p{5.5cm}}
\toprule
\textbf{Cubical} & \textbf{Cohomological} \\
\midrule
Homogeneous composition $\hfilling$ &
  $\Cech$ descent $\mathrm{desc}$ \\
Face formula $\varphi = (i = \varepsilon)$ &
  Refinement predicate $\varphi(x)$ (Z3-decidable) \\
Partial boundary datum &
  Local section $s_k \in \presheaf(\varphi_k)$ \\
Cubical edge agreement &
  Overlap compatibility \\
Kan filling condition &
  $\Hone(\covers, \presheaf) = 0$ \\
Connection $r \wedge s$ (interval meet) &
  Refinement $\varphi \wedge \psi$ (predicate conjunction) \\
Connection $r \vee s$ (interval join) &
  Cover join $\varphi \vee \psi$ \\
Reversal $\sim r$ &
  Predicate negation $\neg\varphi$ \\
De Morgan laws on $\interval$ &
  De Morgan laws on $\mathbb{F}$ (Z3 tautologies) \\
GluePath datum (path-valued section) &
  Parameterized descent over $\interval$ \\
Fiber-preserving homotopy &
  Moving descent (E$_1$-page data) \\
2-path (square) &
  Homotopy of descent data \\
$\Glue$ type &
  Sheaf-theoretic glueing of types \\
$\mathsf{ua}$ (path from equivalence) &
  Descent applied to universe sheaf \\
Transport $\transp^i$ &
  Proof transfer along fiber implication \\
\bottomrule
\end{tabular}
\end{center}
\end{theorem}

\begin{remark}[What the Chapter Proves That Is New]
Compared to earlier presentations of $\mathrm{C}^4$ and to standard
expositions of cubical type theory~\cite{cohen2018} or sheaf
cohomology~\cite{kashiwara2006}, this chapter establishes:
\begin{enumerate}
\item That $\hfilling = \mathrm{desc}$ is a \emph{theorem} (the
  foundational identification), not a remark or analogy
  (Theorem~\ref{thm:hcomp-is-descent}).

\item That the two De Morgan algebras --- on $\interval$ and on refinement
  predicates --- are formally identified via a homomorphism, and that
  this identification is why cover simplification is algebraically sound
  (Theorem~\ref{thm:deMorgan-iso} and
  Corollary~\ref{cor:cover-simplification}).

\item That GluePath (parameterized descent over the interval) is the
  key new proof term enabled by the product structure $\interval^m
  \times \site^n$, and that it is derivable from the descent rule
  (Theorem~\ref{thm:gluepath}).

\item That fiber-preserving homotopies are 2-paths in the fibered proof
  space, and that they provide a formal criterion for ruling out
  ``accidental'' proofs (Theorem~\ref{thm:homotopy-rules-out-accidental}).

\item That overlap compatibility and cubical edge agreement are formally
  the same condition (Proposition~\ref{prop:overlap-is-edge}), and that
  four strategies handle compatibility automatically in typical Python
  code (Proposition~\ref{prop:auto-compat}).

\item That univalence is local and GluePath globalizes it
  (Theorem~\ref{thm:fiberwise-ua}), which is the basis for the
  compiler's refactoring soundness guarantee.
\end{enumerate}
\end{remark}

\begin{remark}[Relation to Hedberg's Theorem]
Throughout this chapter we have used the fact that $\Hone(\covers,
\presheaf) = 0$ for all covers arising from Z3-decidable refinement
predicates on Python types.  The key reason is Hedberg's theorem:
every Python runtime type has decidable equality (via \texttt{==} or
\texttt{is}), which makes all types h-sets, which makes all coefficient
sheaves sheaves of sets, for which higher $\Cech$ cohomology vanishes.

This collapse is what makes $\mathrm{C}^4$ \emph{tractable} for Python
verification: the ``hard'' parts of both theories (higher homotopy groups
in cubical TT, higher differentials in sheaf cohomology) are eliminated
by the discrete nature of Python's type system.  What remains is exactly
the ``easy'' part: paths (proof equality), covers (case decomposition),
overlaps (branch interaction), and transport (proof reuse).  All of
these are Z3-decidable in the $\mathrm{C}^4$ setting.
\end{remark}
