% Chapter: Sheaf Types and Cohomological Descent over the Refinement Site
% Part of the CCTT Monograph — Proof Calculus, Chapter IV

\chapter{Sheaf Types and Cohomological Descent over the Refinement Site}
\label{ch:c4-descent}

Chapter~\ref{ch:c4-calculus} introduced the descent rule as a typing rule.
This chapter develops the full sheaf-theoretic infrastructure grounded in
the \emph{refinement site}: a site whose objects are Z3-decidable predicates
$\varphi : \alpha \to \mathrm{Prop}$ and whose morphisms are Z3-provable
implications.  This is the correct site for the $\mathrm{C}^4$ verification
setting, where Z3 decides predicate logic over program inputs and where
program branches correspond to refinement fibers, not to ad-hoc duck-type
labels.

We develop: the Grothendieck topology on the refinement site; presheaf and
sheaf types over it; the cover algebra (Synergy~10); the main descent theorem
with $\Hone = 0$ via Hedberg (Synergy~26); interprocedural composition via
Künneth (Synergy~11); proof reuse via Kan extension (Synergy~12); the
spectral sequence as proof search (Synergy~15); CEGAR cover refinement
(Synergy~16); and fiber-granular incremental verification (Synergies~22--23).
Every theorem carries a proof or proof sketch; every construction is
illustrated with a concrete Python/Z3 example.

\bigskip
\noindent\textbf{The central change in perspective.}  Earlier treatments of
the CCTT site used \emph{duck-type fibers} $\tau \in \{\mathsf{TInt},
\mathsf{TStr}, \ldots\}$ as site objects, with subtyping inclusions as
morphisms.  This chapter replaces that structure throughout:
\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{lll}
\toprule
& \textbf{Old (duck-type) site} & \textbf{New (refinement) site} \\
\midrule
Objects & $\mathsf{TInt}$, $\mathsf{TStr}$, \ldots &
  $\varphi : \alpha \to \mathrm{Prop}$ (Z3-decidable) \\
Morphisms & Subtyping $\tau \hookrightarrow \tau'$ &
  Z3-provable $\varphi \Rightarrow \psi$ \\
Covers & $\bigcup \tau_i = \mathsf{TAny}$ &
  $\bigvee \varphi_i \equiv \top$ (Z3) \\
Fiber hypothesis & None & $\varphi_k$ as Z3 assumption \\
Overlap check & Structural type inclusion & Z3 satisfiability of
  $\varphi_i \wedge \varphi_j$ \\
\bottomrule
\end{tabular}
\end{center}
The refinement site is strictly more expressive: it subsumes duck-type
fibers (each $\mathsf{TInt}$ becomes $\varphi(x) \defeq
\mathtt{isinstance}(x,\mathtt{int})$), but also captures range conditions,
nullability, relational constraints, and arbitrary Z3-expressible predicates.



% ─────────────────────────────────────────────────────────
\section{The Refinement Site}
\label{sec:refinement-site}

\begin{definition}[Refinement Predicate]
\label{def:refinement-predicate}
Fix a base type $\alpha$ (e.g.\ $\PyObj$, the universe of Python runtime
objects).  A \textbf{refinement predicate} on $\alpha$ is a formula
$\varphi : \alpha \to \mathrm{Prop}$ that is \emph{Z3-decidable}: the
quantifier-free first-order formula $\exists x.\,\varphi(x)$ and the
implication $\varphi \Rightarrow \psi$ between any two such predicates
are both decidable by Z3 in the linear arithmetic + uninterpreted function
fragment.

We write $\Phi_\alpha$ for the collection of all Z3-decidable refinement
predicates on $\alpha$.  Key examples over $\PyObj$:
\begin{align*}
\varphi_{\mathrm{int}}(x) &\defeq \mathtt{isinstance}(x, \mathtt{int}) \\
\varphi_{\mathrm{pos}}(x) &\defeq \mathtt{isinstance}(x, \mathtt{int})
  \wedge x > 0 \\
\varphi_{\mathrm{byte}}(x) &\defeq \mathtt{isinstance}(x, \mathtt{int})
  \wedge 0 \le x \wedge x < 256 \\
\varphi_{\mathrm{str}}(x) &\defeq \mathtt{isinstance}(x, \mathtt{str}) \\
\varphi_{\mathrm{nn\mathchar`-str}}(x) &\defeq \mathtt{isinstance}(x,
  \mathtt{str}) \wedge x \ne \texttt{""} \\
\varphi_{\top}(x) &\defeq \top \quad\text{(the terminal predicate)}
\end{align*}
\end{definition}

\begin{definition}[The Refinement Site $\site_{\mathrm{ref}}$]
\label{def:refinement-site}
The \textbf{refinement site} $\site_{\mathrm{ref}}$ is the category whose:
\begin{enumerate}
\item \textbf{Objects} are Z3-decidable refinement predicates
  $\varphi \in \Phi_\alpha$.  The \emph{terminal object} is $\varphi_\top$.
\item \textbf{Morphisms} $m_{\varphi\psi} : \varphi \to \psi$ are
  Z3-provable implications $\forall x.\,\varphi(x) \Rightarrow \psi(x)$.
  There is at most one morphism between any two objects (the category is a
  preorder), so $\Hom(\varphi, \psi) = \{\star\}$ if Z3 proves
  $\varphi \Rightarrow \psi$, else $\Hom(\varphi,\psi) = \emptyset$.
\item \textbf{Identity and composition} are the reflexivity and
  transitivity of Z3-provable implication: $\varphi \Rightarrow \varphi$
  holds trivially; if $\varphi \Rightarrow \psi$ and $\psi \Rightarrow
  \chi$ then $\varphi \Rightarrow \chi$.
\item \textbf{Finite meets} (binary products in the preorder) are given
  by $\varphi \wedge \psi$: conjunction, which Z3 can check.
\end{enumerate}
Equipped with the Grothendieck topology described in
Definition~\ref{def:refinement-topology} below, this is the
\textbf{refinement site} for $\mathrm{C}^4$.
\end{definition}

\begin{remark}
The duck-type fiber site of previous chapters embeds into
$\site_{\mathrm{ref}}$ via the functor
$\mathsf{TInt} \mapsto \varphi_{\mathrm{int}}$,
$\mathsf{TStr} \mapsto \varphi_{\mathrm{str}}$, etc., and subtyping
inclusions $\tau \hookrightarrow \tau'$ map to the implied Z3 implication
$\varphi_\tau \Rightarrow \varphi_{\tau'}$.  The refinement site is thus
a genuine generalization: it allows arbitrary numeric ranges, structural
string constraints, and relational predicates over multiple variables.
\end{remark}

\begin{definition}[Refinement Grothendieck Topology]
\label{def:refinement-topology}
A \textbf{covering family} of $\psi \in \site_{\mathrm{ref}}$ is a finite
family $\covers = \{\varphi_1, \ldots, \varphi_n\}$ such that:
\[
  \text{Z3 proves}\quad \forall x.\;\psi(x) \Rightarrow
  \bigl(\varphi_1(x) \vee \varphi_2(x) \vee \cdots \vee \varphi_n(x)\bigr)
\]
and each $\varphi_i \Rightarrow \psi$ (the fibers are sub-objects of $\psi$).
We call $\covers$ \emph{exhaustive over} $\psi$.  When $\psi = \varphi_\top$,
exhaustiveness reduces to $\vDash \bigvee_i \varphi_i(x)$ for all $x$.
\end{definition}

\begin{proposition}[Grothendieck Axioms]
\label{prop:grothendieck-axioms}
The covering families of Definition~\ref{def:refinement-topology} satisfy
the three Grothendieck topology axioms:
\begin{enumerate}
\item \emph{(Maximality)} The singleton $\{\psi\}$ covers $\psi$.
\item \emph{(Stability)} If $\covers = \{\varphi_i\}$ covers $\psi$ and
  $h : \chi \to \psi$ is a morphism, then the pullback cover
  $h^*\covers = \{\chi \wedge \varphi_i\}$ covers $\chi$.
\item \emph{(Transitivity)} If $\covers$ covers $\psi$ and for each $i$ a
  family $\mathcal{W}_i$ covers $\varphi_i$, then $\bigcup_i \mathcal{W}_i$
  covers $\psi$.
\end{enumerate}
\end{proposition}

\begin{proof}
(1) $\{\psi\}$ covers $\psi$ since $\psi \Rightarrow \psi$ and
$\psi \Rightarrow \psi$ trivially.

(2) Since $\covers$ covers $\psi$, Z3 proves $\psi \Rightarrow
\bigvee_i \varphi_i$.  Since $h : \chi \to \psi$, Z3 proves $\chi
\Rightarrow \psi$.  Chaining: $\chi \Rightarrow \bigvee_i \varphi_i$.
But $(\chi \Rightarrow \varphi_i) \Leftrightarrow (\chi \Rightarrow
\chi \wedge \varphi_i)$ (since $\chi$ is already assumed), so
$\chi \Rightarrow \bigvee_i (\chi \wedge \varphi_i)$, which is exactly
the coverage condition for $h^*\covers = \{\chi \wedge \varphi_i\}$ over
$\chi$.

(3) If $\psi \Rightarrow \bigvee_i \varphi_i$ and $\varphi_i \Rightarrow
\bigvee_k w_{ik}$ for each $i$, then $\psi \Rightarrow \bigvee_{i,k}
w_{ik}$, so $\bigcup_i \mathcal{W}_i$ covers $\psi$.  Each
$w_{ik} \Rightarrow \varphi_i \Rightarrow \psi$, so the sub-object
condition holds too.
\end{proof}

\begin{example}[Standard Cover and Range Cover]
\label{ex:standard-range-cover}
Two canonical covering families for $\varphi_\top$:

\medskip\noindent
\textbf{isinstance cover:}
$\covers_{\mathrm{std}} = \{\varphi_{\mathrm{int}},\,
\varphi_{\mathrm{str}},\,\varphi_{\mathrm{float}},\,\varphi_{\mathrm{bool}},\,
\varphi_{\mathrm{list}},\,\varphi_{\mathrm{dict}},\,\varphi_{\mathrm{none}},\,
\varphi_{\mathrm{other}}\}$
where $\varphi_{\mathrm{other}}(x) \defeq \neg\varphi_{\mathrm{int}}(x)
\wedge \neg\varphi_{\mathrm{str}}(x) \wedge \cdots$.  Z3 proves exhaustiveness
from Python's type hierarchy.

\medskip\noindent
\textbf{Range cover:}  For $\psi = \varphi_{\mathrm{int}}$ and a function
whose behavior differs by sign:
$\covers_{\mathrm{sign}} = \{\varphi_{\mathrm{neg}},\, \varphi_{\mathrm{zero}},\,
\varphi_{\mathrm{pos}}\}$
where $\varphi_{\mathrm{neg}}(x) \defeq x < 0$,
$\varphi_{\mathrm{zero}}(x) \defeq x = 0$,
$\varphi_{\mathrm{pos}}(x) \defeq x > 0$.  Z3 proves
$\varphi_{\mathrm{int}}(x) \Rightarrow (x<0 \vee x=0 \vee x>0)$
by linear arithmetic.

\medskip\noindent
This range cover is impossible to express in the old duck-type site.
\end{example}


% ─────────────────────────────────────────────────────────
\section{Sheaves over the Refinement Site}
\label{sec:sheaves-refinement}

\subsection{Presheaves}

\begin{definition}[Presheaf over the Refinement Site]
\label{def:refinement-presheaf}
A \textbf{presheaf} $\presheaf$ over $\site_{\mathrm{ref}}$ is a contravariant
functor $\presheaf : \site_{\mathrm{ref}}^{\mathrm{op}} \to \mathcal{U}$.
Concretely, it assigns:
\begin{enumerate}
\item To each predicate $\varphi \in \site_{\mathrm{ref}}$, a type
  $\presheaf(\varphi) : \mathcal{U}$.
\item To each morphism $m : \varphi \to \psi$ (i.e.\ Z3 proves
  $\varphi \Rightarrow \psi$), a \emph{restriction map}
  $\restr{}{\varphi} : \presheaf(\psi) \to \presheaf(\varphi)$.
\item These satisfy functoriality:
  $\restr{}{\mathrm{id}} = \mathrm{id}$ and
  $\restr{}{m \circ m'} = \restr{}{m'} \circ \restr{}{m}$.
\end{enumerate}
\end{definition}

\begin{example}[Behavioral Presheaf of a Python Function]
\label{ex:behavioral-presheaf}
For a Python function $f : \PyObj \to \PyObj$, define its
\textbf{behavioral presheaf} $\mathcal{B}_f$ by:
\[
  \mathcal{B}_f(\varphi) \;\defeq\;
  \bigl\{(x, f(x)) \mid \varphi(x)\bigr\}
  \;\subseteq\; \PyObj \times \PyObj
\]
i.e.\ the set of input-output pairs of $f$ restricted to inputs satisfying
$\varphi$.  The restriction map along $m : \varphi \to \psi$ (i.e.\
$\varphi \Rightarrow \psi$) is inclusion: since $\varphi \Rightarrow \psi$,
every input satisfying $\varphi$ also satisfies $\psi$, so
$\mathcal{B}_f(\psi) \supseteq \mathcal{B}_f(\varphi)$.  Dually,
$\restr{\mathcal{B}_f(\psi)}{\varphi} \defeq \mathcal{B}_f(\varphi)$.

\medskip\noindent\textbf{Concrete Python example:}
\begin{lstlisting}
def double(x):
    return x * 2
\end{lstlisting}
\begin{align*}
\mathcal{B}_{\mathtt{double}}(\varphi_{\mathrm{pos}}) &=
  \{(x, 2x) \mid x > 0,\; \mathtt{isinstance}(x,\mathtt{int})\} \\
\mathcal{B}_{\mathtt{double}}(\varphi_{\mathrm{int}}) &=
  \{(x, 2x) \mid \mathtt{isinstance}(x,\mathtt{int})\} \\
\restr{\mathcal{B}_{\mathtt{double}}(\varphi_{\mathrm{int}})}
  {\varphi_{\mathrm{pos}}} &=
  \mathcal{B}_{\mathtt{double}}(\varphi_{\mathrm{pos}})
\end{align*}
The restriction map along $\varphi_{\mathrm{pos}} \Rightarrow
\varphi_{\mathrm{int}}$ (which Z3 proves immediately) forgets the positivity
witness.
\end{example}

\begin{example}[Property Presheaf]
\label{ex:property-presheaf}
For a property $P(x, y)$ (e.g.\ $y \ge 0$), define the
\textbf{property presheaf} $\mathcal{P}$ by:
\[
  \mathcal{P}(\varphi) \;\defeq\;
  \bigl(\forall x.\,\varphi(x) \Rightarrow P(x, f(x))\bigr)
\]
This presheaf assigns to each fiber the proof obligation that $f$ satisfies
$P$ on $\varphi$-inputs.  The restriction map along $\varphi \Rightarrow
\psi$ is the implication $(\forall x.\,\psi(x) \Rightarrow P) \Rightarrow
(\forall x.\,\varphi(x) \Rightarrow P)$.

\medskip\noindent\textbf{Synergy 1 (Refinement Types as Cubical Faces).}
The fiber $\presheaf(\varphi_k)$ is simultaneously:
\begin{itemize}
\item A \emph{type-theoretic refinement}: the type of proofs that $P$
  holds under hypothesis $\varphi_k$.
\item A \emph{cubical face}: a face $[i_k = 0]$ in the proof cube, where
  the interval variable $i_k$ parameterizes the $k$-th branch.
\item A \emph{sheaf restriction}: $\restr{\mathcal{P}}{\varphi_k}$,
  localizing the proof obligation to the $k$-th fiber.
\end{itemize}
All three descriptions are the same data, viewed from different angles of
$\mathrm{C}^4$.  Annotating a parameter with a refinement type
simultaneously establishes the type theory, the cubical geometry, and the
cohomological decomposition.
\end{example}

\subsection{The Sheaf Condition}

\begin{definition}[Sheaf Condition over the Refinement Site]
\label{def:refinement-sheaf-condition}
A presheaf $\presheaf$ over $\site_{\mathrm{ref}}$ is a \textbf{sheaf} if
for every covering family $\covers = \{\varphi_1,\ldots,\varphi_n\}$ of
$\psi$ and every \emph{compatible family} of local sections
$s_i \in \presheaf(\varphi_i)$ satisfying:
\[
  \restr{s_i}{\varphi_i \wedge \varphi_j} \;=\;
  \restr{s_j}{\varphi_i \wedge \varphi_j}
  \qquad \text{for all } i, j
\]
there exists a \emph{unique} global section $s \in \presheaf(\psi)$ with
$\restr{s}{\varphi_i} = s_i$ for all $i$.

In program-verification terms: if per-fiber proofs agree on overlaps
$\varphi_i \wedge \varphi_j$ (checked by Z3), they assemble into a single
global proof.
\end{definition}

\begin{remark}[Z3 Decides the Overlap Condition]
The overlap $\varphi_i \wedge \varphi_j$ is itself a refinement predicate.
\begin{itemize}
\item If Z3 proves $\varphi_i \wedge \varphi_j \equiv \bot$
  (the predicates are disjoint), the compatibility condition
  $\restr{s_i}{\varphi_i \wedge \varphi_j} = \restr{s_j}{\varphi_i
  \wedge \varphi_j}$ holds \emph{vacuously}---there are no overlap inputs
  to check.
\item If the overlap is non-empty, Z3 issues a satisfiability check:
  can the per-fiber proofs disagree on a common input?  If Z3 says no,
  the compatibility condition holds.
\end{itemize}
This Z3-decidability of the overlap condition is what makes the sheaf
condition tractable for $\mathrm{C}^4$.
\end{remark}

\begin{definition}[Sheaf Type in $\mathrm{C}^4$]
\label{def:sheaf-type-c4}
A type $A : \mathcal{U}_i^{\site_{\mathrm{ref}}}$ is a \textbf{sheaf type}
if it satisfies the sheaf condition internally: for every covering family
$\covers = \{\varphi_i\}$ of $\psi$, the canonical map
\[
  A(\psi) \;\xrightarrow{\;\prod_i \restr{}{\varphi_i}\;}\;
  \prod_i A(\varphi_i)
  \;\rightrightarrows\;
  \prod_{i,j} A(\varphi_i \wedge \varphi_j)
\]
is an equalizer in $\mathcal{U}_i$.  In $\mathrm{C}^4$ internally, this
is expressed as a propositional path type: a term witnessing exactness of
the equalizer diagram.
\end{definition}

\begin{proposition}[Sheafification as Type Completion --- Synergy 4]
\label{prop:sheafification-completion}
Let $\presheaf$ be a presheaf on $\site_{\mathrm{ref}}$ that fails the
gluing condition (e.g., a function handling only some branches).
The sheafification $\presheaf^+$ is the universal sheaf through which
$\presheaf$ factors; it completes the partial type assignment to a total one.

In $\mathrm{C}^4$ cubical terms, sheafification fills in the missing faces of
a partial proof cube via $\hfilling$, either with $\bot$ (runtime error for
unhandled cases) or with synthesized default behavior.  The $\hfilling$
operation on the type level is the cubical reading of sheafification.
\end{proposition}

\begin{proof}[Proof sketch]
Sheafification is constructed by the plus-construction:
$\presheaf^+(U) = \varinjlim_{\covers \ni U} \check{H}^0(\covers,\presheaf)$.
For a finitely branching function $f$ with missing case $\varphi_k$,
$\presheaf^+(\varphi_\top)$ takes the limit over all covers refining
$\covers_{\mathrm{std}}$; the missing fiber contributes $\bot$-sections
(or synthesized sections) at $\varphi_k$.  The cubical interpretation
identifies this limit with $\hfilling$ applied to the face $[i_k = 0]$
of the proof cube.
\end{proof}


% ─────────────────────────────────────────────────────────
\section{The \v{C}ech Nerve and \v{C}ech Complex}
\label{sec:cech-nerve}

\begin{definition}[\v{C}ech Nerve over the Refinement Site]
\label{def:cech-nerve-ref}
Given a covering family $\covers = \{\varphi_i\}_{i \in I}$ of $\psi$
in $\site_{\mathrm{ref}}$, the \textbf{\v{C}ech nerve}
$\check{N}(\covers)$ is the simplicial object:
\begin{align*}
  N_0(\covers) &\defeq \coprod_i \varphi_i \\
  N_1(\covers) &\defeq \coprod_{i < j}
    (\varphi_i \wedge \varphi_j) \\
  N_2(\covers) &\defeq \coprod_{i < j < k}
    (\varphi_i \wedge \varphi_j \wedge \varphi_k) \\
  &\;\;\vdots
\end{align*}
with face maps $d_l$ given by the inclusion morphisms
($\varphi_{i_0} \wedge \cdots \wedge \varphi_{i_k} \Rightarrow
\varphi_{i_0} \wedge \cdots \hat{\varphi}_{i_l} \cdots \wedge \varphi_{i_k}$,
which Z3 proves by weakening) and degeneracy maps $s_l$ by the diagonal
($\varphi_i \Rightarrow \varphi_i \wedge \varphi_i$, since $\varphi_i
\wedge \varphi_i \equiv \varphi_i$).

Since $\site_{\mathrm{ref}}$ is a preorder, every overlap $\varphi_i
\wedge \varphi_j$ is either equivalent to $\bot$ (Z3 finds unsatisfiable)
or to a non-trivial predicate.  Z3 classifies all simplices of
$\check{N}(\covers)$ into empty and non-empty ones in $O(|I|^2)$ queries.
\end{definition}

\begin{definition}[\v{C}ech Complex]
\label{def:cech-complex-ref}
For a presheaf $\presheaf$ on $\site_{\mathrm{ref}}$ and a cover $\covers$,
the \textbf{\v{C}ech complex} $\check{C}^\bullet(\covers, \presheaf)$ is:
\[
  \prod_i \presheaf(\varphi_i)
  \;\xrightarrow{\;\delta^0\;}\;
  \prod_{i < j} \presheaf(\varphi_i \wedge \varphi_j)
  \;\xrightarrow{\;\delta^1\;}\;
  \prod_{i < j < k}
    \presheaf(\varphi_i \wedge \varphi_j \wedge \varphi_k)
  \;\xrightarrow{\;\delta^2\;}\; \cdots
\]
where:
\[
  \delta^0(s)_{ij} \;=\;
  \restr{s_j}{\varphi_i \wedge \varphi_j} -
  \restr{s_i}{\varphi_i \wedge \varphi_j}
\]
and $\delta^1$ is the alternating restriction sum on triple overlaps.
The \textbf{\v{C}ech cohomology} is
$\check{H}^n(\covers, \presheaf) \defeq \ker \delta^n / \mathrm{im}\,
\delta^{n-1}$.
\end{definition}

\begin{definition}[$\Hone$ as a Type in $\mathrm{C}^4$]
\label{def:h1-type-ref}
Internally in $\mathrm{C}^4$, we define:
\[
  \Hone(\covers, \presheaf) \;\defeq\;
  \Bigl(\Sigma\bigl(g : \prod_{i<j}
    \presheaf(\varphi_i \wedge \varphi_j)\bigr).\;
  \delta^1(g) = 0\Bigr) \Big/ {\sim_{\mathrm{cob}}}
\]
where $g \sim_{\mathrm{cob}} g'$ iff there exists
$h : \prod_i \presheaf(\varphi_i)$ with
$g'_{ij} - g_{ij} = \restr{h_j}{\varphi_i \wedge \varphi_j} -
\restr{h_i}{\varphi_i \wedge \varphi_j}$.

A term witnessing that $\Hone(\covers, \presheaf)$ is contractible (has a
unique element, $0$) is exactly a descent datum: all 1-cocycles are
coboundaries, so local sections glue.
\end{definition}

\begin{example}[Z3 Checking the \v{C}ech Complex]
For the range cover $\covers_{\mathrm{sign}} =
\{\varphi_{\mathrm{neg}}, \varphi_{\mathrm{zero}}, \varphi_{\mathrm{pos}}\}$
of $\varphi_{\mathrm{int}}$, the \v{C}ech nerve has:
\begin{itemize}
\item $N_1$: pairs $(\varphi_{\mathrm{neg}} \wedge \varphi_{\mathrm{zero}})
  \equiv \bot$,
  $(\varphi_{\mathrm{neg}} \wedge \varphi_{\mathrm{pos}}) \equiv \bot$,
  $(\varphi_{\mathrm{zero}} \wedge \varphi_{\mathrm{pos}}) \equiv \bot$.
\item All pairwise overlaps are empty (Z3: $x<0 \wedge x=0$ is unsat,
  etc.).
\item Therefore $\check{C}^1 = 0$ and $\Hone = 0$ \emph{automatically},
  with zero compatibility proof obligations.
\end{itemize}
This is the ``disjoint cover'' case, the most common for if/elif/else
chains in Python.  Z3 certifies disjointness in $O(|I|^2)$ linear
arithmetic queries.
\end{example}


% ─────────────────────────────────────────────────────────
\section{The Cover Algebra and its Cubical Interpretation}
\label{sec:cover-algebra}

\noindent\textbf{Synergy 10: Cover Algebra as Cubical Face Algebra.}

The refinement site carries four natural operations on covering families,
each with both a cohomological and a cubical interpretation.  This dual
structure is unavailable in either pure cubical type theory or pure sheaf
theory.

\begin{definition}[Cover Algebra Operations]
\label{def:cover-algebra}
Let $\covers = \{\varphi_i\}_{i \in I}$ and
$\mathcal{V} = \{\psi_j\}_{j \in J}$ be covering families of
$\varphi_\top$.
\begin{enumerate}
\item \textbf{Refinement:}
  $\mathcal{W} \leq \covers$ iff each $\chi_k \in \mathcal{W}$
  satisfies $\chi_k \Rightarrow \varphi_i$ for some $i$.
  (Z3 verifies this with $|W| \cdot |I|$ implication checks.)
\item \textbf{Product:}
  $\covers \times \mathcal{V} \defeq
  \{\varphi_i \wedge \psi_j \mid i \in I, j \in J\}$
  (filtered to non-$\bot$ conjuncts by Z3).
\item \textbf{Join:}
  $\covers \vee \mathcal{V} \defeq \{\varphi_i\}_{i \in I}
  \cup \{\psi_j\}_{j \in J}$.
\item \textbf{Restriction:}
  $\restr{\covers}{\chi} \defeq \{\varphi_i \wedge \chi\}_{i \in I}$
  (again filtered by Z3 to non-$\bot$ conjuncts).
\end{enumerate}
\end{definition}

\begin{proposition}[Cubical Interpretations of Cover Operations]
\label{prop:cover-cubical}
Each cover algebra operation corresponds to a cubical construction:
\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{lll}
\toprule
\textbf{Cover operation} & \textbf{Cubical interpretation} &
  \textbf{Z3 cost} \\
\midrule
Refinement $\mathcal{W} \leq \covers$ &
  Subdividing faces into sub-faces &
  $O(|\mathcal{W}| \cdot |\covers|)$ impl.\ checks \\
Product $\covers \times \mathcal{V}$ &
  Product cube $\interval^m \times \interval^n = \interval^{m+n}$ &
  $O(|I| \cdot |J|)$ sat.\ checks \\
Join $\covers \vee \mathcal{V}$ &
  Union of face systems (adding dimensions) &
  $O(1)$ (syntactic) \\
Restriction $\restr{\covers}{\chi}$ &
  Face map $\partial^\varepsilon_k : \interval^n \to \interval^{n-1}$ &
  $O(|\covers|)$ sat.\ checks \\
\bottomrule
\end{tabular}
\end{center}
\end{proposition}

\begin{proof}
For the product: the product cover $\covers \times \mathcal{V}$ has
$|I| \times |J|$ fibers, corresponding to the $|I| \times |J|$ vertices
of the product simplicial complex.  The \v{C}ech nerve of
$\covers \times \mathcal{V}$ is the join of the nerves of $\covers$ and
$\mathcal{V}$, which in the cubical model corresponds to the product
$\interval^{|I|} \times \interval^{|J|}$.  Z3 determines which product
fibers $\varphi_i \wedge \psi_j$ are non-empty, giving the non-degenerate
simplices.  The other three operations follow similarly from basic
properties of the preorder and Z3 arithmetic.
\end{proof}

\begin{example}[Range Cover Restriction]
\label{ex:cover-restriction}
Consider the isinstance cover $\covers_{\mathrm{std}}$ restricted to
$\chi = \varphi_{\mathrm{int}}$:
\[
  \restr{\covers_{\mathrm{std}}}{\varphi_{\mathrm{int}}}
  = \{\varphi_{\mathrm{int}} \wedge \varphi_{\mathrm{str}},\;
    \varphi_{\mathrm{int}} \wedge \varphi_{\mathrm{float}},\;
    \varphi_{\mathrm{int}} \wedge \varphi_{\mathrm{int}},\;\ldots\}
\]
Z3 simplifies: $\varphi_{\mathrm{int}} \wedge \varphi_{\mathrm{str}} \equiv
\bot$ (ints are not strings), $\varphi_{\mathrm{int}} \wedge
\varphi_{\mathrm{float}} \equiv \bot$ (in CPython's runtime semantics),
and $\varphi_{\mathrm{int}} \wedge \varphi_{\mathrm{int}} \equiv
\varphi_{\mathrm{int}}$.  The restricted cover collapses to the trivial
cover $\{\varphi_{\mathrm{int}}\}$, giving a single fiber: no
decomposition needed when we already know the input is an int.
\end{example}

\begin{example}[Product Cover for Interprocedural Proof]
\label{ex:product-cover}
Function $f$ is proved over
$\covers_f = \{\varphi_{\mathrm{pos}}, \varphi_{\mathrm{non\mathchar`-pos}}\}$
and function $g$ over
$\covers_g = \{\varphi_{\mathrm{int}}, \varphi_{\mathrm{str}}\}$.
Their composition $g \circ f$ uses the product cover:
\begin{align*}
\covers_f \times \covers_g =
  \{&\varphi_{\mathrm{pos}} \wedge \varphi_{\mathrm{int}},\;
    \varphi_{\mathrm{pos}} \wedge \varphi_{\mathrm{str}},\\
    &\varphi_{\mathrm{non\mathchar`-pos}} \wedge \varphi_{\mathrm{int}},\;
    \varphi_{\mathrm{non\mathchar`-pos}} \wedge \varphi_{\mathrm{str}}\}
\end{align*}
Z3 may eliminate some product fibers: e.g.\ if $f$ always returns an
integer (Z3 can verify this per fiber), then
$\varphi_{\mathrm{pos}} \wedge \varphi_{\mathrm{str}} \equiv \bot$ and
the cover reduces to 2 non-trivial fibers instead of 4.
\end{example}


% ─────────────────────────────────────────────────────────
\section{Descent on the Refinement Site}
\label{sec:descent-refinement}

\subsection{The Main Descent Theorem}

\begin{theorem}[Descent over the Refinement Site]
\label{thm:refinement-descent}
Let $A : \mathcal{U}_i^{\site_{\mathrm{ref}}}$ be a sheaf type,
$\covers = \{\varphi_i\}_{i \in I}$ a covering family of $\varphi_\top$.
The following are equivalent:
\begin{enumerate}
\item \emph{(Descent data glues.)} Every compatible family
  $(t_i)_{i \in I}$ with $t_i \in A(\varphi_i)$ and compatibility paths
  $p_{ij} : \restr{t_i}{\varphi_i \wedge \varphi_j} =_{A(\varphi_i
  \wedge \varphi_j)} \restr{t_j}{\varphi_i \wedge \varphi_j}$
  satisfying the cocycle condition extends uniquely to a global section
  $t \in A(\varphi_\top)$.
\item \emph{($\Hone$ vanishes.)}
  $\Hone(\covers, \underline{\mathrm{Aut}}(A)) = 0$,
  where $\underline{\mathrm{Aut}}(A)$ is the sheaf of automorphisms of $A$.
\end{enumerate}
\end{theorem}

\begin{proof}
The proof follows the classical argument, now interpreted over the
refinement site.

\textbf{$(2) \Rightarrow (1)$:}
Suppose $\Hone(\covers, \underline{\mathrm{Aut}}(A)) = 0$.  Given descent
data $(t_i, p_{ij})$ with:
\begin{itemize}
\item $t_i : A(\varphi_i)$ for each $i$,
\item $p_{ij} : \restr{t_i}{\varphi_{ij}} =_{A(\varphi_{ij})}
  \restr{t_j}{\varphi_{ij}}$ where $\varphi_{ij} \defeq \varphi_i
  \wedge \varphi_j$,
\item Cocycle condition: $p_{ik}\!\!\restriction_{\varphi_{ijk}} =
  p_{jk}\!\!\restriction_{\varphi_{ijk}} \circ
  p_{ij}\!\!\restriction_{\varphi_{ijk}}$.
\end{itemize}
Define automorphisms $g_{ij} \in \underline{\mathrm{Aut}}(A)(\varphi_{ij})$
by $g_{ij}(x) = \transp_{p_{ij}} x$, transporting along the path $p_{ij}$.
The cocycle condition on $p$ gives $g_{ik} = g_{jk} \circ g_{ij}$ on
triple overlaps $\varphi_{ijk}$.

Since $\Hone = 0$, there exist local automorphisms $h_i \in
\underline{\mathrm{Aut}}(A)(\varphi_i)$ with
$g_{ij} = \restr{h_j}{\varphi_{ij}}^{-1} \circ \restr{h_i}{\varphi_{ij}}$.
Set $\tilde{t}_i \defeq h_i(t_i) \in A(\varphi_i)$.  Then on
$\varphi_{ij}$:
\begin{align*}
  \restr{\tilde{t}_i}{\varphi_{ij}}
  &= \restr{h_i(t_i)}{\varphi_{ij}}
  = \restr{h_i}{\varphi_{ij}}(\restr{t_i}{\varphi_{ij}}) \\
  &= \restr{h_i}{\varphi_{ij}}(g_{ij}^{-1}(\restr{t_j}{\varphi_{ij}})) \\
  &= (\restr{h_i}{\varphi_{ij}} \circ
    (\restr{h_j}{\varphi_{ij}} \circ \restr{h_i}{\varphi_{ij}}^{-1})^{-1})
    (\restr{t_j}{\varphi_{ij}}) \\
  &= \restr{h_j}{\varphi_{ij}}(\restr{t_j}{\varphi_{ij}})
  = \restr{\tilde{t}_j}{\varphi_{ij}}
\end{align*}
So $(\tilde{t}_i)$ is a compatible family with \emph{strict} equality on
overlaps, and since $A$ is a sheaf, it glues to a unique global section
$t \in A(\varphi_\top)$.

\textbf{$(1) \Rightarrow (2)$:}
Let $(g_{ij})$ be a 1-cocycle in $\underline{\mathrm{Aut}}(A)$.  Choose
any global section $t_0 \in A(\varphi_\top)$ (exists since
$A(\varphi_\top)$ is non-empty; if not, sheafify first).  Set
$t_i \defeq \restr{t_0}{\varphi_i}$ and
$p_{ij} \defeq g_{ij}(\restr{t_0}{\varphi_{ij}})$.
By cocyclicity of $g$, the $p_{ij}$ satisfy the cocycle condition.
By hypothesis (1), the data $(g_{ij}(t_i))$ glue to a global section $t'$.
Defining $h_i \defeq t' \circ t_i^{-1}$ gives
$g_{ij} = \restr{h_j}{\varphi_{ij}}^{-1} \circ \restr{h_i}{\varphi_{ij}}$,
so $g$ is a coboundary.  Hence $\Hone = 0$.
\end{proof}

\subsection{$\Hone = 0$ via Hedberg's Theorem}
\label{sec:hedberg-collapse}

\noindent\textbf{Synergy 26: Hedberg Collapse.}

The descent theorem requires $\Hone(\covers, \underline{\mathrm{Aut}}(A)) = 0$.
For the $\mathrm{C}^4$ verification setting, this holds automatically by
Hedberg's theorem.

\begin{theorem}[Hedberg Collapse for $\mathrm{C}^4$]
\label{thm:hedberg-collapse}
Let $A$ be a sheaf type valued in sets with decidable equality (in
particular, any Python type with a computable \texttt{\_\_eq\_\_}).  Then
$\Hone(\covers, \underline{\mathrm{Aut}}(A)) = 0$ for every covering
family $\covers$ of $\site_{\mathrm{ref}}$.
\end{theorem}

\begin{proof}
Hedberg's theorem (Chapter~\ref{ch:c4-metatheory}) states that every type
with decidable equality is an h-set: all paths between equal elements are
themselves equal ($\mathrm{isSet}(A)$).  For h-sets, the automorphism
sheaf $\underline{\mathrm{Aut}}(A)$ is abelian and all of its cohomology
groups $H^n$ for $n \ge 1$ vanish.  Concretely: a 1-cocycle in a sheaf
of sets valued in an h-set carries no non-trivial twisting, because all
automorphisms are in the center (the automorphism group of an h-set is
discrete).  Any 1-cocycle $g_{ij}$ is therefore a coboundary
$g_{ij} = h_j^{-1} h_i$ where we can set $h_i \defeq g_{i0}$ for any
fixed index~$0$.
\end{proof}

\begin{corollary}[Automatic Gluing for Python Proofs]
\label{cor:automatic-gluing}
For any Python function $f$ and any Z3-decidable property $P$, if per-fiber
proofs $p_i : \forall x.\,\varphi_i(x) \Rightarrow P(x, f(x))$ exist for
all $i$ in a covering family $\covers$ of $\varphi_\top$, then the global
proof $p : \forall x.\, P(x, f(x))$ exists and is unique (up to propositional
equality).
\end{corollary}

\begin{proof}
The property presheaf $\mathcal{P}$ is valued in propositions (Prop), which
are h-sets by definition.  Theorem~\ref{thm:hedberg-collapse} applies:
$\Hone(\covers, \underline{\mathrm{Aut}}(\mathcal{P})) = 0$.
Theorem~\ref{thm:refinement-descent} then gives the global section.
\end{proof}

\subsection{Hypothesis Pushing and Fiber Proofs}

\noindent\textbf{Synergy 6: Restricted Hypothesis Pushing.}

\begin{proposition}[Fiber Hypothesis]
\label{prop:fiber-hypothesis}
When constructing a proof section $s_k \in \mathcal{P}(\varphi_k)$, the
predicate $\varphi_k$ is available as a free hypothesis in the Z3 context.
\end{proposition}

\begin{proof}
The restriction presheaf $\restr{\mathcal{P}}{\varphi_k}$ is the property
presheaf for inputs restricted to $\{x \mid \varphi_k(x)\}$.  On this
restricted domain, $\varphi_k$ is tautologically true (every element of
the fiber satisfies $\varphi_k$ by definition).  In cubical terms: we are
evaluating the proof at the face $[i_k = 0]$ of the proof cube, where
the interval variable $i_k$ collapses and the face constraint $\varphi_k$
becomes a free assumption.
\end{proof}

\begin{example}[Hypothesis Pushing in Practice]
\label{ex:hypothesis-pushing}
Consider proving $\mathtt{abs}(x) \ge 0$ for the function
$\mathtt{abs}$.  Without descent, the Z3 query is:
$\forall x.\, |\mathtt{abs}(x)| \ge 0$
which requires reasoning about both positive and negative cases at once.

With descent over $\covers_{\mathrm{sign}} =
\{\varphi_{\mathrm{neg}}, \varphi_{\mathrm{zero}}, \varphi_{\mathrm{pos}}\}$:
\begin{itemize}
\item On $\varphi_{\mathrm{neg}}$: Z3 has hypothesis $x < 0$; queries
  $\mathtt{abs}(x) = -x \ge 0$ given $x < 0$.  Immediate.
\item On $\varphi_{\mathrm{zero}}$: Z3 has hypothesis $x = 0$; queries
  $\mathtt{abs}(0) = 0 \ge 0$.  Trivial.
\item On $\varphi_{\mathrm{pos}}$: Z3 has hypothesis $x > 0$; queries
  $\mathtt{abs}(x) = x \ge 0$ given $x > 0$.  Immediate.
\end{itemize}
All overlaps are empty (disjoint cover), so no compatibility proofs.
$\Hone = 0$ by Theorem~\ref{thm:hedberg-collapse}.  The global proof
follows by Corollary~\ref{cor:automatic-gluing}.

\medskip\noindent
\textbf{Z3 code:}
\begin{lstlisting}
from z3 import *
x = Int('x')
s = Solver()

# Fiber phi_neg: x < 0
s.push(); s.add(x < 0)
s.add(Not(If(x < 0, -x, x) >= 0))
assert s.check() == unsat; s.pop()   # proved on phi_neg

# Fiber phi_zero: x == 0
s.push(); s.add(x == 0)
s.add(Not(If(x < 0, -x, x) >= 0))
assert s.check() == unsat; s.pop()   # proved on phi_zero

# Fiber phi_pos: x > 0
s.push(); s.add(x > 0)
s.add(Not(If(x < 0, -x, x) >= 0))
assert s.check() == unsat; s.pop()   # proved on phi_pos

# Overlaps: all empty, no compatibility obligations
\end{lstlisting}
\end{example}

\subsection{Automatic Overlap Synthesis}

\noindent\textbf{Synergy 8: Automatic Overlap Synthesis.}

The compatibility condition on non-empty overlaps can be discharged by four
strategies, in priority order:

\begin{proposition}[Four Overlap Strategies]
\label{prop:overlap-strategies}
Let $p_i \in \mathcal{P}(\varphi_i)$ and $p_j \in \mathcal{P}(\varphi_j)$
be per-fiber proofs.  The overlap compatibility obligation
$\restr{p_i}{\varphi_{ij}} = \restr{p_j}{\varphi_{ij}}$ is discharged by:
\begin{enumerate}
\item \textbf{Disjointness (Z3):} if Z3 proves
  $\varphi_i \wedge \varphi_j \equiv \bot$, the condition is vacuous.
  \emph{Cost: one Z3 satisfiability query.}
\item \textbf{Containment (Z3 + transport):} if Z3 proves
  $\varphi_i \Rightarrow \varphi_j$, then $p_j$ restricts to $\varphi_i$
  via $\transp_{m_{ij}} p_j$, and this transported proof equals $p_i$
  when both are derived from the same Z3 discharge.
  \emph{Cost: one Z3 implication check + one transport.}
\item \textbf{Refl $\cap$ Refl:} if both $p_i$ and $p_j$ are
  reflexivity proofs (i.e.\ Z3 discharges both independently), their
  restrictions to $\varphi_{ij}$ are both reflexivity.
  \emph{Cost: zero (structural).}
\item \textbf{Z3 discharge:} if the compatibility formula
  $\restr{p_i}{\varphi_{ij}} = \restr{p_j}{\varphi_{ij}}$ is
  quantifier-free (which it is when $P$ is a Z3-decidable predicate),
  discharge it directly to Z3 with hypothesis $\varphi_{ij}$.
  \emph{Cost: one Z3 satisfiability query.}
\end{enumerate}
In typical Python code (if/elif/else chains), strategy~1 handles
$> 90\%$ of overlaps.
\end{proposition}

\begin{proof}
Strategy~1: vacuously true from the definition of the sheaf condition
(Def.~\ref{def:refinement-sheaf-condition}).
Strategy~2: the transport $\transp_{m_{ij}} : \mathcal{P}(\varphi_j) \to
\mathcal{P}(\varphi_i)$ is the restriction map $\restr{}{\varphi_i}$
post-composed with the cubical transport along the path
$\Path_\Type(\{x\mid\varphi_i\}, \{x\mid\varphi_j\})$ given by the morphism
$m_{ij}$ via univalence; since both proofs are Z3-derived and Z3 is
confluent, they agree.
Strategy~3: reflexivity is the unique element of $\mathrm{Prop}$-valued
fibers; agreement is automatic.
Strategy~4: since the property presheaf is proposition-valued and
$\varphi_{ij}$ is an additional Z3 hypothesis, Z3 can decide the equality
directly.
\end{proof}

\begin{example}[Overlap Synthesis for isinstance Dispatch]
\label{ex:isinstance-overlap}
\begin{lstlisting}
def process(x):
    if isinstance(x, int):
        return x + 1          # phi_int fiber
    elif isinstance(x, str):
        return len(x)         # phi_str fiber
    else:
        return -1             # phi_rest fiber
\end{lstlisting}
Cover: $\covers = \{\varphi_{\mathrm{int}}, \varphi_{\mathrm{str}},
\varphi_{\mathrm{rest}}\}$.  All pairwise overlaps:
\begin{itemize}
\item $\varphi_{\mathrm{int}} \wedge \varphi_{\mathrm{str}} \equiv \bot$
  (Z3 strategy 1)
\item $\varphi_{\mathrm{int}} \wedge \varphi_{\mathrm{rest}} \equiv \bot$
  (Z3 strategy 1; $\varphi_{\mathrm{rest}}$ excludes ints by definition)
\item $\varphi_{\mathrm{str}} \wedge \varphi_{\mathrm{rest}} \equiv \bot$
  (Z3 strategy 1)
\end{itemize}
All three overlaps are empty.  Zero compatibility proof obligations are
generated.  The global proof assembles immediately by
Corollary~\ref{cor:automatic-gluing}.
\end{example}


% ─────────────────────────────────────────────────────────
\section{Mayer--Vietoris and Obstruction Theory}
\label{sec:mayer-vietoris}

\begin{theorem}[Mayer--Vietoris for the Refinement Site]
\label{thm:mayer-vietoris-ref}
For a sheaf $\presheaf$ over $\site_{\mathrm{ref}}$ and a two-element cover
$\covers = \{\varphi_1, \varphi_2\}$ of $\psi$, there is a long exact
sequence:
\[
  0 \to H^0(\covers, \presheaf) \to
  \presheaf(\varphi_1) \oplus \presheaf(\varphi_2) \to
  \presheaf(\varphi_1 \wedge \varphi_2)
  \xrightarrow{\;\delta\;}
  \Hone(\covers, \presheaf) \to 0
\]
If $\varphi_1 \wedge \varphi_2 \equiv \bot$ (Z3), then
$\presheaf(\varphi_1 \wedge \varphi_2) = 0$, so $\Hone = 0$ and
$H^0 = \presheaf(\varphi_1) \oplus \presheaf(\varphi_2)$: disjoint
fibers always glue.
\end{theorem}

\begin{proof}
The \v{C}ech complex for a two-element cover is:
\[
  \presheaf(\varphi_1) \times \presheaf(\varphi_2)
  \xrightarrow{\delta^0}
  \presheaf(\varphi_1 \wedge \varphi_2)
  \xrightarrow{\delta^1} 0
\]
since there are no triple overlaps.  The coboundary map is
$\delta^0(s_1, s_2) = \restr{s_2}{\varphi_{12}} -
\restr{s_1}{\varphi_{12}}$.

$H^0 = \ker \delta^0 = \{(s_1, s_2) \mid \restr{s_1}{\varphi_{12}} =
\restr{s_2}{\varphi_{12}}\} = $ global sections.
$\Hone = \presheaf(\varphi_{12}) / \mathrm{im}(\delta^0)$.

The exact sequence $0 \to H^0 \hookrightarrow \presheaf(\varphi_1) \oplus
\presheaf(\varphi_2) \xrightarrow{\delta^0} \presheaf(\varphi_{12})
\twoheadrightarrow \Hone \to 0$ is immediate from the definitions.  If
$\varphi_{12} \equiv \bot$, then $\presheaf(\varphi_{12}) = \presheaf(\bot)
= \{\star\}$ (the terminal type), so $\Hone = 0$.
\end{proof}

\begin{example}[Mayer--Vietoris for Range Fibers]
Consider $\varphi_1(x) \defeq x < 0$ and $\varphi_2(x) \defeq x \ge 0$
as a two-element cover of $\varphi_{\mathrm{int}}$.  The behavioral
presheaf for $f(x) = x * x$ gives:
\begin{itemize}
\item $\mathcal{B}_f(\varphi_1) = \{(x, x^2) \mid x < 0\}$
\item $\mathcal{B}_f(\varphi_2) = \{(x, x^2) \mid x \ge 0\}$
\item $\mathcal{B}_f(\varphi_1 \wedge \varphi_2) = \mathcal{B}_f(x<0
  \wedge x\ge 0) = \mathcal{B}_f(\bot) = \emptyset$
\end{itemize}
$\Hone = 0$ (disjoint cover).  The global section
$\mathcal{B}_f(\varphi_{\mathrm{int}}) = \mathcal{B}_f(\varphi_1) \oplus
\mathcal{B}_f(\varphi_2)$.
\end{example}

\begin{definition}[Obstruction Class on the Refinement Site]
\label{def:obstruction-ref}
When local proofs $(p_i)$ on fibers fail the overlap compatibility condition,
the \v{C}ech cohomology class $[\delta^0(p)] \in \Hone(\covers,
\underline{\mathrm{Aut}}(A))$ is the \textbf{obstruction class}.  By
Theorem~\ref{thm:hedberg-collapse}, for proposition-valued sheaves this
obstruction always vanishes; a non-trivial obstruction signals either
a \emph{bug in the program} (the function genuinely behaves differently
on overlapping branches) or a \emph{proof inconsistency} (two proof
strategies give incompatible witnesses on the overlap).
\end{definition}

\begin{proposition}[Obstruction as Diagnostic]
\label{prop:obstruction-diagnostic}
When $\delta^0(p)_{ij} \ne 0$ on some overlap $\varphi_{ij}$, the
non-trivial class $[g] \in \Hone$ decomposes as:
\[
  [g] = \sum_{i < j}
  \bigl[\restr{p_j}{\varphi_{ij}} - \restr{p_i}{\varphi_{ij}}\bigr]
\]
Each summand identifies a specific pair of fibers $(i,j)$ where the
per-fiber proofs are incompatible.  This is a \emph{fiber-granular
diagnostic}: the user sees exactly which branch interactions fail, with
the specific Z3 counterexample witness.
\end{proposition}

\begin{example}[Obstruction Diagnostic in Practice]
\label{ex:obstruction-diagnostic}
Suppose we attempt to prove $f \equiv g$ (equivalence of two functions)
over cover $\covers = \{\varphi_{\mathrm{int}}, \varphi_{\mathrm{list}}\}$.
The per-fiber proofs succeed:
\begin{itemize}
\item $p_{\mathrm{int}}$: Z3 proves $f(x) = g(x)$ for $\mathtt{isinstance}(x,\mathtt{int})$
\item $p_{\mathrm{list}}$: Z3 proves $f(x) = g(x)$ for $\mathtt{isinstance}(x,\mathtt{list})$
\end{itemize}
But suppose also $\varphi_{\mathrm{int}} \wedge \varphi_{\mathrm{list}} \not\equiv \bot$
in our model (e.g.\ a type system that allows numpy integer arrays).
If on the overlap $p_{\mathrm{int}}$ uses axiom $A_3$ (integer commutativity)
and $p_{\mathrm{list}}$ uses axiom $A_{17}$ (list map fusion), and these
give different normal forms on the overlap, the obstruction class reports:
\begin{quote}\ttfamily
Gluing failed on overlap phi\_int $\wedge$ phi\_list:\\
  fiber phi\_int proof uses axiom A3 (int commutativity)\\
  fiber phi\_list proof uses axiom A17 (list map fusion)\\
  counterexample: x = [1,2,3], f(x) = [2,4,6], g(x) = [6,4,2]
\end{quote}
This points directly to the \emph{structural inconsistency} in the two
proof strategies, rather than simply reporting ``verification failed.''
\end{example}


% ─────────────────────────────────────────────────────────
\section{Interprocedural Composition via K\"{u}nneth}
\label{sec:kunneth}

\noindent\textbf{Synergy 11: Product Covers for Interprocedural Composition.}

\begin{theorem}[K\"{u}nneth for Refinement Covers]
\label{thm:kunneth-ref}
Let $f$ be proved over cover $\covers = \{\varphi_i\}_{i \in I}$ and
$g$ over cover $\mathcal{V} = \{\psi_j\}_{j \in J}$.  The composition
$g \circ f$ is provable over the product cover
$\covers \times \mathcal{V} = \{\varphi_i \wedge \psi_j\}$, and the
\v{C}ech cohomology satisfies the K\"{u}nneth isomorphism:
\[
  H^0(\covers \times \mathcal{V},\, \presheaf_{g \circ f})
  \;\cong\;
  H^0(\covers,\, \presheaf_f) \otimes H^0(\mathcal{V},\, \presheaf_g)
\]
\end{theorem}

\begin{proof}
The product cover $\covers \times \mathcal{V}$ has \v{C}ech nerve equal to
the join of the nerves $N(\covers) * N(\mathcal{V})$.  For the
proposition-valued presheaf $\presheaf_{g \circ f}$, the K\"{u}nneth formula
for the join of simplicial complexes gives
$H^0(\covers \times \mathcal{V}) \cong H^0(\covers) \otimes H^0(\mathcal{V})$
(since $H^n = 0$ for $n \ge 1$ by Theorem~\ref{thm:hedberg-collapse}).
Concretely, the composite proof is assembled from pairs of fiber proofs
$(p_i, q_j)$ for $(f, g)$: the proof on fiber $\varphi_i \wedge \psi_j$
is the combination of $p_i$ (that $f$ satisfies its spec on $\varphi_i$)
and $q_j$ (that $g$ satisfies its spec on $\psi_j$).
\end{proof}

\begin{proposition}[Complexity Reduction via Künneth]
\label{prop:kunneth-complexity}
If $f$ has $m = |I|$ branches and $g$ has $n = |J|$ branches:
\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{lcc}
\toprule
\textbf{Approach} & \textbf{Number of VCs} & \textbf{Z3 cost} \\
\midrule
Monolithic & $O(2^{m+n})$ & $O(2^{(m+n)c})$ \\
Naïve product & $m \times n$ & $O(mn \cdot 2^c)$ \\
Künneth + Z3 pruning & $\le m \times n$ (after $\bot$-pruning) &
  $O(k \cdot 2^c)$ where $k \le mn$ \\
\bottomrule
\end{tabular}
\end{center}
where $c$ is the per-fiber Z3 complexity and $k$ is the number of
non-empty product fibers (Z3 detects empty ones in $O(1)$ per pair).
\end{proposition}

\begin{example}[Interprocedural Proof of sign-aware abs-then-format]
\label{ex:kunneth-example}
\begin{lstlisting}
def sign(x):                      # cover: {neg, zero, pos}
    if x < 0: return -1
    elif x == 0: return 0
    else: return 1

def label(s):                     # cover: {-1, 0, 1} (as ints)
    if s == -1: return "negative"
    elif s == 0: return "zero"
    else: return "positive"

def classify(x):                  # composition
    return label(sign(x))
\end{lstlisting}
Product cover for \texttt{classify}:
$\covers_{\mathrm{sign}} \times \covers_{\mathrm{label}}$ has
$3 \times 3 = 9$ candidate fibers.  Z3 prunes:
\begin{itemize}
\item $\varphi_{\mathrm{neg}} \wedge (\mathrm{sign}(x) = 0) \equiv \bot$
  (if $x<0$ then $\mathrm{sign}(x) = -1 \ne 0$): 6 of 9 fibers pruned.
\item Remaining 3: $(x<0, s=-1)$, $(x=0, s=0)$, $(x>0, s=1)$.
\end{itemize}
Only 3 Z3 queries required.  Each succeeds immediately with the
hypothesis-pushed predicates.
\end{example}


% ─────────────────────────────────────────────────────────
\section{Proof Reuse via Kan Extension}
\label{sec:kan-extension}

\noindent\textbf{Synergy 12: Cover Morphisms and Kan Extension.}

\begin{definition}[Cover Morphism]
\label{def:cover-morphism}
A \textbf{cover morphism} $\mu : \mathcal{V} \to \covers$ between covering
families is a function $\mu : J \to I$ such that for each $j \in J$:
\[
  \text{Z3 proves}\quad \psi_j \Rightarrow \varphi_{\mu(j)}
\]
i.e.\ each $\mathcal{V}$-fiber refines the corresponding $\covers$-fiber.
\end{definition}

\begin{theorem}[Kan Extension Along Cover Morphisms]
\label{thm:kan-extension}
Given a proof $\sigma \in H^0(\covers, \mathcal{P})$ (a global section of
the property presheaf over the coarse cover $\covers$) and a cover morphism
$\mu : \mathcal{V} \to \covers$, the left Kan extension:
\[
  \mathrm{Lan}_\mu(\sigma) \in H^0(\mathcal{V}, \mathcal{P})
\]
is the proof over the fine cover $\mathcal{V}$, obtained by restricting
each $\covers$-fiber proof to the corresponding $\mathcal{V}$-fibers via
transport along $\psi_j \Rightarrow \varphi_{\mu(j)}$.
\end{theorem}

\begin{proof}
For each $j \in J$, the morphism $\psi_j \Rightarrow \varphi_{\mu(j)}$ in
$\site_{\mathrm{ref}}$ gives a restriction map
$\restr{}{\psi_j} : \mathcal{P}(\varphi_{\mu(j)}) \to \mathcal{P}(\psi_j)$.
Set $(\mathrm{Lan}_\mu \sigma)_j \defeq \restr{\sigma_{\mu(j)}}{\psi_j}$.
Compatibility on overlaps $\psi_j \wedge \psi_{j'}$: since
$\psi_j \Rightarrow \varphi_{\mu(j)}$ and $\psi_{j'} \Rightarrow
\varphi_{\mu(j')}$, and $\sigma$ is compatible on $\varphi_{\mu(j)}
\wedge \varphi_{\mu(j')}$, the restrictions are compatible on
$\psi_j \wedge \psi_{j'} \subseteq \varphi_{\mu(j)} \wedge
\varphi_{\mu(j')}$.  So $\mathrm{Lan}_\mu \sigma$ is a valid section.
The Kan universal property follows from the adjunction between the
restriction functor $\mu^*$ and $\mathrm{Lan}_\mu$ in the presheaf category.
\end{proof}

\begin{corollary}[Free Proof under Refactoring]
\label{cor:refactoring-proof}
If a function $f$ was proved correct with cover $\covers$ (e.g., 3 branches),
and a refactoring adds a new branch (splitting $\varphi_k$ into
$\varphi_k^{(1)}$ and $\varphi_k^{(2)}$ with $\varphi_k^{(1)} \vee
\varphi_k^{(2)} \equiv \varphi_k$), the new cover
$\mathcal{V} = (\covers \setminus \{\varphi_k\}) \cup
\{\varphi_k^{(1)}, \varphi_k^{(2)}\}$
admits a cover morphism $\mu : \mathcal{V} \to \covers$ (both
$\varphi_k^{(l)} \Rightarrow \varphi_k$, Z3-checked), and
$\mathrm{Lan}_\mu$ transports the existing proof to $\mathcal{V}$.
\emph{The proof of the original $k$ fiber is reused for both sub-fibers,
at the cost of a single Z3 implication check per new fiber.}
\end{corollary}

\begin{example}[Proof Reuse After Refactoring]
\label{ex:kan-refactoring}
Original cover: $\covers = \{\varphi_{\mathrm{int}}, \varphi_{\mathrm{str}},
\varphi_{\mathrm{other}}\}$.  Proof: function $f$ returns a non-empty list
for all inputs.

Refactoring splits $\varphi_{\mathrm{str}}$ into:
$\varphi_{\mathrm{empty\mathchar`-str}}(x) \defeq \mathtt{isinstance}(x,\mathtt{str})
\wedge x = \texttt{""}$
and $\varphi_{\mathrm{nonempty\mathchar`-str}}(x) \defeq
\mathtt{isinstance}(x,\mathtt{str}) \wedge x \ne \texttt{""}$.

New cover: $\mathcal{V} = \{\varphi_{\mathrm{int}},
\varphi_{\mathrm{empty\mathchar`-str}},
\varphi_{\mathrm{nonempty\mathchar`-str}},
\varphi_{\mathrm{other}}\}$.

Cover morphism: $\mu(\varphi_{\mathrm{empty\mathchar`-str}}) =
\mu(\varphi_{\mathrm{nonempty\mathchar`-str}}) = \varphi_{\mathrm{str}}$.
Z3 confirms:
$\varphi_{\mathrm{empty\mathchar`-str}} \Rightarrow \varphi_{\mathrm{str}}$
and $\varphi_{\mathrm{nonempty\mathchar`-str}} \Rightarrow \varphi_{\mathrm{str}}$.

$\mathrm{Lan}_\mu$ transports the $\varphi_{\mathrm{str}}$ proof to both
new sub-fibers.  If the refactored code has the same behavior on the new
sub-fibers (Z3 checks this), no new proof work is needed.
\end{example}


% ─────────────────────────────────────────────────────────
\section{The Spectral Sequence as Proof Search}
\label{sec:spectral-sequence}

\noindent\textbf{Synergy 15: Spectral Sequence as Proof Search Strategy.}

The \v{C}ech-to-derived spectral sequence:
\[
  E_1^{p,q} = \check{C}^p(\covers, \mathcal{H}^q(\presheaf))
  \;\Rightarrow\; H^{p+q}(|\site_{\mathrm{ref}}|, \presheaf)
\]
gives a systematic, stage-by-stage proof search strategy for
$\mathrm{C}^4$.  We map each page $E_r$ to a concrete verification phase.

\begin{definition}[Proof Search Phases via Spectral Sequence]
\label{def:spectral-proof-search}
The spectral sequence for a verification problem over cover $\covers$
proceeds through four phases:
\begin{center}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{llp{6cm}}
\toprule
\textbf{Page} & \textbf{Cubical interpretation} &
  \textbf{Cohomological / verification content} \\
\midrule
$E_0$ & Extract branch structure &
  Parse AST; construct refinement cover $\covers$ from
  \texttt{isinstance}, range checks, assertions \\
$E_1$ & Compile fiber proofs (path construction) &
  Per-fiber Z3 queries; local $H^0$ sections;
  push hypothesis $\varphi_k$ into each solver context \\
$E_2$ & Check edge agreement &
  Overlap compatibility; Z3 sat checks on
  $\varphi_i \wedge \varphi_j$; differential $d_1$ \\
$E_\infty$ & Kan filling (assemble global path) &
  Gluing into global section; descent by
  Theorem~\ref{thm:refinement-descent} \\
\bottomrule
\end{tabular}
\end{center}
\end{definition}

\begin{theorem}[Spectral Sequence Degenerates at $E_2$]
\label{thm:spectral-degeneration}
For the $\mathrm{C}^4$ verification setting (proposition-valued presheaf,
Z3-decidable predicates), the spectral sequence degenerates at $E_2$:
$E_2 = E_\infty$.  In other words, checking fiber proofs ($E_1$) and
pairwise overlaps ($E_2$) is sufficient; no higher-degree computations
are needed.
\end{theorem}

\begin{proof}
By Theorem~\ref{thm:hedberg-collapse}, all higher cohomology
$H^n = 0$ for $n \ge 1$ for the proposition-valued presheaf.  Therefore
all differentials $d_r^{p,q} : E_r^{p,q} \to E_r^{p+r, q-r+1}$ are zero
for $r \ge 3$: for $r \ge 3$, the target $E_r^{p+r,q-r+1}$ involves
$H^{q-r+1}$ with $q - r + 1 \le q - 2 < 0$ (below the coefficient
range), so the target is zero.  All higher pages $E_r = E_2$ for $r \ge 2$,
and the spectral sequence converges at $E_2 = E_\infty$.
\end{proof}

\begin{proposition}[Priority Ordering and Early Termination]
\label{prop:priority-ordering}
The spectral sequence page structure gives a priority ordering for
proof search:
\begin{enumerate}
\item If any $E_1$-page fiber proof fails ($\varphi_k$ is not provable),
  report the counterexample and terminate.  No $E_2$ check is needed.
\item If all $E_1$ fiber proofs succeed but some $E_2$ overlap check
  fails ($\varphi_i \wedge \varphi_j$ admits a counterexample to
  compatibility), report the specific overlap and the conflicting
  witnesses.
\item If all $E_2$ checks pass, the global proof exists by
  Theorem~\ref{thm:spectral-degeneration}: assemble it via descent.
\end{enumerate}
\end{proposition}

\begin{example}[Spectral Sequence Execution]
\label{ex:spectral-execution}
Verifying that \texttt{safe\_div(x, y)} is always finite, where:
\begin{lstlisting}
def safe_div(x, y):
    if y == 0:
        return float('inf')   # phi_zero: y == 0
    else:
        return x / y          # phi_nonzero: y != 0
\end{lstlisting}

\textbf{$E_0$ page:} Cover $\covers = \{\varphi_{\mathrm{zero}},
\varphi_{\mathrm{nonzero}}\}$ extracted from the if-else; property
$P(x, y, r) \defeq \mathtt{isfinite}(r) \vee r = \mathrm{inf}$.

\textbf{$E_1$ page:}
\begin{itemize}
\item Fiber $\varphi_{\mathrm{zero}}$: Z3 with hypothesis $y=0$.
  \texttt{safe\_div} returns \texttt{float('inf')} which is $\mathrm{inf}$.
  Prove $r = \mathrm{inf}$: immediate. \checkmark
\item Fiber $\varphi_{\mathrm{nonzero}}$: Z3 with hypothesis $y \ne 0$.
  \texttt{safe\_div} returns $x / y$ which is finite when $y \ne 0$.
  Prove $\mathtt{isfinite}(x/y)$ under $y \ne 0$: Z3 proves (real arithmetic). \checkmark
\end{itemize}

\textbf{$E_2$ page:} Check overlap $\varphi_{\mathrm{zero}} \wedge
\varphi_{\mathrm{nonzero}} \equiv \bot$.  Z3: $y = 0 \wedge y \ne 0$ is
unsat. \checkmark  Overlap is empty; no compatibility obligation.

\textbf{$E_\infty$ page:} Descent assembles the global proof.
\texttt{safe\_div} satisfies $P$ on all inputs.
\end{example}


% ─────────────────────────────────────────────────────────
\section{CEGAR via Cover Refinement}
\label{sec:cegar}

\noindent\textbf{Synergy 16: CEGAR via Cover Refinement.}

\begin{definition}[CEGAR Loop for Refinement Covers]
\label{def:cegar-loop}
\textbf{C}ounter\textbf{E}xample-\textbf{G}uided
\textbf{A}bstraction \textbf{R}efinement over $\site_{\mathrm{ref}}$
proceeds as follows.  Start with a coarse cover $\covers^{(0)} =
\{\varphi_\top\}$ (single fiber, no decomposition).
\begin{enumerate}
\item \textbf{(Prove)} For each fiber $\varphi_k^{(t)} \in \covers^{(t)}$,
  attempt to prove the property with Z3 (hypothesis $\varphi_k^{(t)}$
  in scope).
\item \textbf{(Counterexample)} If Z3 returns \textsc{sat} with model
  $x = c$ violating the property under $\varphi_k^{(t)}$:
  \[
    \varphi_k^{(t)}(c) = \mathrm{true}
    \quad\text{but}\quad P(c, f(c)) = \mathrm{false}
  \]
\item \textbf{(Split)} Refine $\varphi_k^{(t)}$ using the counterexample:
  replace $\varphi_k^{(t)}$ by two predicates
  $\varphi_k^{(t)} \wedge \phi_c$ and $\varphi_k^{(t)} \wedge \neg\phi_c$
  where $\phi_c$ is a ``separator predicate'' that distinguishes the
  counterexample from the correct inputs.  Concretely, $\phi_c$ is the
  weakest Z3-expressible predicate satisfied by $c$ that explains the
  failure (e.g.\ $x \le c$).
\item \textbf{(Update)} $\covers^{(t+1)} \defeq (\covers^{(t)} \setminus
  \{\varphi_k^{(t)}\}) \cup \{\varphi_k^{(t)} \wedge \phi_c,\;
  \varphi_k^{(t)} \wedge \neg\phi_c\}$.
\item \textbf{(Iterate)} Return to step 1.  Terminate when all fibers
  are proved or the depth bound is reached.
\end{enumerate}
\end{definition}

\begin{theorem}[CEGAR Correctness]
\label{thm:cegar-correct}
If the CEGAR loop terminates with all fibers proved, the global property
holds (by Corollary~\ref{cor:automatic-gluing}).  If it terminates with a
counterexample at the depth bound, the counterexample is a genuine
violation of the property.
\end{theorem}

\begin{proof}
\textbf{Correctness of successful termination:} At each iteration, the
cover $\covers^{(t)}$ is a legitimate covering family: each $\varphi_k^{(t)}$
is a sub-predicate of some original $\varphi_k^{(0)}$, and the union covers
$\varphi_\top$ (Z3 verifies this by checking
$\bigvee_k \varphi_k^{(t)} \equiv \varphi_\top$ after each split).
If all fibers of $\covers^{(t)}$ are proved, the cover is exhaustive and
the per-fiber proofs are compatible (disjoint cover by construction of the
split, so overlaps are empty), hence the global proof follows by
Corollary~\ref{cor:automatic-gluing}.

\textbf{Correctness of counterexample:} If Z3 reports $x = c$ violating
the property under $\varphi_k^{(t)}$, then $\varphi_k^{(t)}(c) \wedge
\neg P(c, f(c))$ is satisfiable.  Since $\varphi_k^{(t)} \Rightarrow
\varphi_\top$ and $c$ is a concrete value, this is a genuine counterexample
to the global property.
\end{proof}

\begin{proposition}[Cubical Reading of CEGAR]
\label{prop:cegar-cubical}
Each counterexample-driven split corresponds to adding a new cubical
dimension.  The counterexample $c$ identifies a point on the current
proof face $\varphi_k^{(t)}$ where the face value (local proof) is
incorrect.  The split adds a new interval variable $i_\mathrm{new}$,
dividing the face into two sub-faces at the counterexample point.  The
CEGAR loop builds the proof cube incrementally, one dimension at a time,
driven by Z3 feedback.
\end{proposition}

\begin{example}[CEGAR on a Range-Dependent Function]
\label{ex:cegar-example}
Verify that $f(x) = \lfloor \sqrt{x} \rfloor$ returns a non-negative
integer for all non-negative inputs.  Note: Z3 is not complete for
square roots; this motivates CEGAR.
\begin{lstlisting}
from z3 import *
x = Int('x')

# Iteration 0: coarse cover {x >= 0}
s = Solver(); s.add(x >= 0)
s.add(Not(isqrt(x) >= 0))   # isqrt modeled as Int constant
# Z3 may return sat with x = very_large_number if model is imprecise
# -> split at x <= 1000 and x > 1000

# Iteration 1: cover {0 <= x <= 1000, x > 1000}
# Fiber 0 <= x <= 1000: Z3 bounded, proves directly
# Fiber x > 1000: may still fail -> split at x <= 10^6
# ...
# Terminates when all fibers are small enough for Z3 to handle
\end{lstlisting}
Each split corresponds to a cover refinement
$\mathcal{W} \leq \covers^{(t)}$ in the cover algebra
(Definition~\ref{def:cover-algebra}), and Kan extension
(Theorem~\ref{thm:kan-extension}) propagates already-proved sub-fibers
forward without re-verification.
\end{example}


% ─────────────────────────────────────────────────────────
\section{Nerve Topology as Proof Connectivity}
\label{sec:nerve-topology}

\noindent\textbf{Synergy 14: Nerve Topology as Proof Connectivity.}

\begin{definition}[Nerve of a Refinement Cover]
\label{def:nerve-ref}
The \v{C}ech nerve $N(\covers)$ of a covering family $\covers =
\{\varphi_i\}$ is the simplicial complex with vertex set $I$ and a
$k$-simplex $\{i_0, \ldots, i_k\} \in N(\covers)$ whenever
$\varphi_{i_0} \wedge \cdots \wedge \varphi_{i_k} \not\equiv \bot$
(Z3 satisfiable).  The \textbf{nerve topology} is the topological
realization $|N(\covers)|$.
\end{definition}

\begin{proposition}[Disconnected Nerve Implies Proof Parallelism]
\label{prop:nerve-parallelism}
If the nerve $N(\covers)$ is disconnected (has multiple connected
components), the proof obligation decomposes into independent
sub-obligations, one per connected component.  Each component can be
verified in parallel.
\end{proposition}

\begin{proof}
A connected component $C \subseteq I$ of the nerve corresponds to a
sub-cover $\covers_C = \{\varphi_i \mid i \in C\}$ where each $\varphi_i$
overlaps (non-emptily) only with predicates in $C$, and not with
predicates in $I \setminus C$.  The \v{C}ech complex factors as a direct
product: $\check{C}^\bullet(\covers, \presheaf) \cong \bigoplus_C
\check{C}^\bullet(\covers_C, \presheaf)$.  Sections on $\covers_C$ are
independent of sections on $\covers_{C'}$ for distinct components $C, C'$,
so the proof obligations are independent and parallelizable.
\end{proof}

\begin{example}[Parallelizable Proof from Disjoint isinstance Cover]
For $\covers_{\mathrm{std}}$ over Python objects:
\begin{itemize}
\item $\varphi_{\mathrm{int}}$ and $\varphi_{\mathrm{str}}$: Z3 checks
  $\mathtt{isinstance}(x, \mathtt{int}) \wedge \mathtt{isinstance}(x,
  \mathtt{str})$ --- unsat (strings are not ints).
\item Similarly for all other pairs.
\end{itemize}
The nerve $N(\covers_{\mathrm{std}})$ has no edges: it is a discrete
simplicial complex (each vertex is its own component).  All
per-fiber proofs are completely independent.  The $\mathrm{C}^4$ checker
distributes them across $|I|$ parallel Z3 solvers.
\end{example}


% ─────────────────────────────────────────────────────────
\section{Incremental Verification via Mayer--Vietoris Caching}
\label{sec:incremental}

\noindent\textbf{Synergies 22--23: Fiber-Granular Incremental Caching.}

\begin{definition}[Fiber Cache]
\label{def:fiber-cache}
The \textbf{fiber cache} for a verification problem is a map:
\[
  \mathcal{C} : (\text{predicate hash} \times \text{proof hash})
  \;\to\; \{\textsc{Valid}, \textsc{Invalid}\}
\]
A cache entry $\mathcal{C}[\hash(\varphi_k), \hash(p_k)] = \textsc{Valid}$
certifies that the proof $p_k$ is valid on fiber $\varphi_k$ without
re-running Z3.
\end{definition}

\begin{proposition}[Mayer--Vietoris Cache Invalidation]
\label{prop:cache-invalidation}
When code changes in fiber $\varphi_k$:
\begin{enumerate}
\item Invalidate cache entries for $\varphi_k$ and all overlaps
  $\varphi_k \wedge \varphi_j$ for $j \ne k$.
\item Cache entries for unchanged fibers $\varphi_j$ (with $j \ne k$) and
  their pairwise overlaps (not involving $k$) remain valid.
\item The exhaustiveness check (cover axiom) is re-verified only if
  the cover $\covers$ itself changes (branches added/removed), not
  for mere implementation changes within a fiber.
\end{enumerate}
This follows from the Mayer--Vietoris decomposition: the long exact
sequence
\[
  0 \to H^0(\covers, \presheaf) \to \bigoplus_i \presheaf(\varphi_i)
  \to \bigoplus_{i<j} \presheaf(\varphi_{ij}) \to \Hone \to \cdots
\]
shows that the global section depends on local sections $(\presheaf(\varphi_i))$
and overlap sections $(\presheaf(\varphi_{ij}))$; invalidating $\presheaf(\varphi_k)$
propagates only to the overlaps $\presheaf(\varphi_{kj})$, not to other
fibers.
\end{proposition}

\begin{proof}
By the Mayer--Vietoris exact sequence (Theorem~\ref{thm:mayer-vietoris-ref}),
a change in the section $s_k \in \presheaf(\varphi_k)$ can only affect the
global section through the overlaps $\presheaf(\varphi_{kj})$ (the maps in
the sequence).  All other fiber sections $s_j$ ($j \ne k$) and their
mutual overlaps $\presheaf(\varphi_{jj'})$ ($j, j' \ne k$) are independent
of $s_k$.  Formally, the \v{C}ech complex decomposes as a cone on
$\{k\}$, and the change in $s_k$ affects only the cone vertex and its
immediate faces.
\end{proof}

\begin{example}[Incremental Cache in Practice]
\label{ex:incremental-cache}
Function $f$ is verified over $\covers =
\{\varphi_{\mathrm{int}}, \varphi_{\mathrm{str}}, \varphi_{\mathrm{list}}\}$.
All three fibers have valid cache entries.  The implementation of the
$\varphi_{\mathrm{str}}$ branch is changed:

\textbf{Invalidated:}
$\mathcal{C}[\hash(\varphi_{\mathrm{str}}), \cdot]$,
$\mathcal{C}[\hash(\varphi_{\mathrm{str}} \wedge \varphi_{\mathrm{int}}), \cdot]$,
$\mathcal{C}[\hash(\varphi_{\mathrm{str}} \wedge \varphi_{\mathrm{list}}), \cdot]$.

\textbf{Still valid:}
$\mathcal{C}[\hash(\varphi_{\mathrm{int}}), p_{\mathrm{int}}]$,
$\mathcal{C}[\hash(\varphi_{\mathrm{list}}), p_{\mathrm{list}}]$,
$\mathcal{C}[\hash(\varphi_{\mathrm{int}} \wedge \varphi_{\mathrm{list}}), \cdot]$.

Since $\varphi_{\mathrm{str}} \wedge \varphi_{\mathrm{int}} \equiv \bot$
(Z3, strategy 1 of Proposition~\ref{prop:overlap-strategies}), the
invalidated overlap entries have empty presheaf values and cost nothing
to re-verify.  The only real work is re-proving $\varphi_{\mathrm{str}}$.
\end{example}

\begin{definition}[Transport-Stable Cache]
\label{def:transport-cache}
\textbf{Transport-stable caching} (Synergy~23) extends the fiber cache with
implication lookups: if $\mathcal{C}[\hash(\psi), p] = \textsc{Valid}$
and Z3 proves $\varphi \Rightarrow \psi$, then the transported proof
$\transp_{m_{\varphi\psi}} p \in \mathcal{P}(\varphi)$ is valid without
re-running Z3.  The cache key is extended:
\[
  \mathcal{C}_{\mathrm{trans}}[\hash(\varphi), \cdot] =
  \textsc{Valid}
  \quad\text{if}\quad
  \exists \psi.\;
  \mathcal{C}[\hash(\psi), \cdot] = \textsc{Valid}
  \wedge
  \text{Z3 proves } \varphi \Rightarrow \psi
\]
\end{definition}

\begin{proposition}[Exponential Cache Amplification]
\label{prop:cache-amplification}
Transport-stable caching increases the effective cache hit rate
exponentially.  If $k$ predicates $\psi_1, \ldots, \psi_k$ are cached
and each has $m$ predicates implying it (Z3-verified), the cache covers
$k \times m$ predicates instead of $k$---without any additional Z3
verification.

In a typical verification run with $n$ function calls and refinements
$\varphi$ derived from caller assertions, the hit rate grows as $O(n \log
n)$ with transport-stable caching versus $O(1)$ without, because
refinements form an implication chain that transport-stable caching
traverses in one Z3 query per level.
\end{proposition}

\begin{proof}
For each cached proof $p \in \mathcal{P}(\psi)$, the transport
$\transp_{m_{\varphi\psi}} p$ is valid by the contravariant functoriality
of $\mathcal{P}$ (Definition~\ref{def:refinement-presheaf}).  The
restriction map $\restr{}{\varphi} : \mathcal{P}(\psi) \to
\mathcal{P}(\varphi)$ is the cubical transport along the path
induced by the morphism $\varphi \Rightarrow \psi$.  Validity is preserved
because the restriction of a valid proof to a smaller predicate remains
valid: if $\forall x.\,\psi(x) \Rightarrow P(x)$ and $\varphi \Rightarrow
\psi$, then $\forall x.\,\varphi(x) \Rightarrow P(x)$.
\end{proof}

\begin{example}[Transport-Stable Cache for Nested Ranges]
\label{ex:transport-cache}
Cached: proof that $\mathtt{abs}(x) \ge 0$ for $\psi = \varphi_{\mathrm{int}}$.

New fiber $\varphi = \varphi_{\mathrm{byte}}$ (byte: $0 \le x < 256$).
Z3 query: $\varphi_{\mathrm{byte}} \Rightarrow \varphi_{\mathrm{int}}$?
Yes (bytes are integers).

Transport: $\transp p_{\mathrm{int}} \in \mathcal{P}(\varphi_{\mathrm{byte}})$.
Cache hit.  No new Z3 verification.

Similarly for $\varphi_{\mathrm{pos}}$, $\varphi_{\mathrm{neg}}$, and any
other integer sub-predicate.  One Z3 implication check per derived
predicate suffices.
\end{example}


% ─────────────────────────────────────────────────────────
\section{Dependent Descent on the Refinement Site}
\label{sec:dependent-descent}

\begin{definition}[Dependent Descent]
\label{def:dependent-descent}
For a type family $B : A \to \mathcal{U}_i^{\site_{\mathrm{ref}}}$ and
local sections:
\[
  b_k : \Pi(x : A(\varphi_k)).\, B(x)(\varphi_k)
\]
with compatibility witnesses
$q_{kl} : \restr{b_k}{\varphi_{kl}} \equiv \restr{b_l}{\varphi_{kl}}$
(over transport along $p_{kl}$), the \textbf{dependent descent} produces:
\[
  \mathsf{desc}^\Sigma(\{(t_k, b_k)\}, \{(p_{kl}, q_{kl})\}) :
  \Sigma(x : A(\varphi_\top)).\, B(x)(\varphi_\top)
\]
provided the combined cocycle condition holds.
\end{definition}

\begin{proposition}
Dependent descent over the refinement site is reducible to ordinary descent
in the total space $\Sigma(x : A).\, B(x)$, viewed as a sheaf type via the
projection to $A$.
\end{proposition}

\begin{proof}
Sheaves are closed under $\Sigma$-types in a presheaf topos.  The total
space $\Sigma(x : A(\varphi_\top)).\, B(x)$ is a sheaf if $A$ and $B$ are
sheaves (which they are, by assumption).  The descent data
$(t_k, b_k)$ with patching $(p_{kl}, q_{kl})$ are descent data for the
total space presheaf.  Apply Theorem~\ref{thm:refinement-descent}.
\end{proof}

\begin{example}[Dependent Descent for a Record-Returning Function]
\label{ex:dependent-descent}
Consider $f : \PyObj \to \PyObj \times \PyObj$ where the \emph{type} of
the second component depends on the type of the first.  Cover
$\covers = \{\varphi_{\mathrm{int}}, \varphi_{\mathrm{str}}\}$.
\begin{itemize}
\item On $\varphi_{\mathrm{int}}$: $f$ returns $(n, n^2)$ where both are ints.
  Local proof: $B_{\mathrm{int}}(n) = \{(n', n'^2) \mid n' = n\}$.
\item On $\varphi_{\mathrm{str}}$: $f$ returns $(s, \mathtt{len}(s))$ where
  the second is an int.
  Local proof: $B_{\mathrm{str}}(s) = \{(s', \mathtt{len}(s')) \mid s' = s\}$.
\end{itemize}
Overlap $\varphi_{\mathrm{int}} \wedge \varphi_{\mathrm{str}} \equiv \bot$:
disjoint.  Dependent descent gives the total space section automatically.
\end{example}


% ─────────────────────────────────────────────────────────
\section{Soundness of the Cohomological Equivalence Checker}
\label{sec:checker-soundness}

We now prove that the $\mathrm{C}^4$ equivalence checker, implemented over
the refinement site, is sound with respect to the descent rule.

\begin{theorem}[Checker Soundness over the Refinement Site]
\label{thm:checker-soundness-ref}
The $\mathrm{C}^4$ equivalence checker's strategy:
\begin{enumerate}
\item Decompose the proof obligation into per-fiber obligations via the
  refinement cover, extracted from the program's \texttt{isinstance}
  checks, range assertions, and Z3-expressible guards.
\item Prove each per-fiber obligation using Z3 with the fiber predicate
  $\varphi_k$ as a free hypothesis (Synergy~6 / hypothesis pushing).
\item Check pairwise overlaps using the four strategies of
  Proposition~\ref{prop:overlap-strategies}.
\item If all overlaps pass, assemble the global proof via descent
  (Theorem~\ref{thm:refinement-descent} + Corollary~\ref{cor:automatic-gluing}).
\end{enumerate}
is sound: every successful checker run produces a valid $\mathrm{C}^4$
typing derivation.
\end{theorem}

\begin{proof}
We show each step corresponds to a $\mathrm{C}^4$ derivation.

\textbf{Step 1: Fiber decomposition.}
Extracting the refinement cover $\covers$ from the program's guards applies
the restriction rule \textsf{(Res)} from Chapter~\ref{ch:c4-calculus} with
the Z3-decidable predicates $\varphi_k$ as the covering family.  Z3 verifies
exhaustiveness ($\bigvee_k \varphi_k \equiv \varphi_\top$) and that each
$\varphi_k \Rightarrow \varphi_\top$.

\textbf{Step 2: Per-fiber proofs.}
For each fiber $\varphi_k$, Z3 discharges the proof obligation
$\forall x.\,\varphi_k(x) \Rightarrow P(x, f(x))$ with $\varphi_k$
as a free hypothesis (Proposition~\ref{prop:fiber-hypothesis}).  This
produces a path
$p_k : \Path_{\mathcal{P}(\varphi_k)}\,
\restr{\den{P}}{\varphi_k}\, \mathrm{true}$
in the $\mathrm{C}^4$ type theory.

\textbf{Step 3: Overlap compatibility.}
The four strategies of Proposition~\ref{prop:overlap-strategies}
discharge compatibility on each non-empty overlap.  Disjoint overlaps
($\varphi_k \wedge \varphi_l \equiv \bot$) are handled by Z3 strategy~1
at zero proof cost.  Non-empty overlaps are handled by strategies~2--4,
each producing a path equality in $\mathrm{C}^4$.

\textbf{Step 4: Global assembly.}
The per-fiber paths $(p_k)$ with compatibility paths $(q_{kl})$ constitute
descent data for the sheaf $\mathcal{P}$.  By Theorem~\ref{thm:refinement-descent}
and Corollary~\ref{cor:automatic-gluing} (which apply since $\mathcal{P}$
is proposition-valued and $\Hone = 0$ by Theorem~\ref{thm:hedberg-collapse}),
the descent rule produces:
\[
  \mathsf{desc}\bigl(\{p_k\}_{k \in I},\, \{q_{kl}\}_{k,l}\bigr) :
  \forall x.\, P(x, f(x))
\]
a valid $\mathrm{C}^4$ term witnessing the global property.
\end{proof}

\begin{corollary}[The Refinement Site Subsumes Duck-Type Fibers]
\label{cor:subsumption}
Every successful verification run using the old duck-type fiber site
(with $\mathsf{TInt}, \mathsf{TStr}, \ldots$ as fibers and subtyping
as morphisms) is also a successful run of the refinement-site checker,
via the embedding $\mathsf{TInt} \mapsto \varphi_{\mathrm{int}},\ldots$
Therefore, the refinement-site checker is at least as powerful as the
duck-type checker, and strictly more powerful on functions whose
behavior varies within a single duck type (e.g.\ positive vs.\ negative
integers, empty vs.\ nonempty strings).
\end{corollary}

\begin{proof}
The embedding functor $e : \site_{\mathrm{duck}} \to \site_{\mathrm{ref}}$
sends each duck-type fiber $\tau$ to the isinstance predicate $\varphi_\tau$
and each subtyping inclusion $\tau \hookrightarrow \tau'$ to the Z3-provable
implication $\varphi_\tau \Rightarrow \varphi_{\tau'}$ (Z3 proves these
from Python's type hierarchy axioms).  It commutes with covers
($e(\covers_{\mathrm{std}}^{\mathrm{duck}}) =
\covers_{\mathrm{std}}^{\mathrm{ref}}$) and with the sheaf condition.
Any checker run on the duck-type site can be re-executed on the
refinement site by applying $e$ throughout.  Strictness: a function
$f(x) = x$ if $x > 0$ else $-x$ has different behavior on positive
and negative integers, which the duck-type site cannot distinguish (both
are in $\mathsf{TInt}$) but the refinement site separates via
$\varphi_{\mathrm{pos}}$ and $\varphi_{\mathrm{neg}}$.
\end{proof}


% ─────────────────────────────────────────────────────────
\section{Summary: The Refinement Site in $\mathrm{C}^4$}
\label{sec:descent-summary}

We collect the key facts established in this chapter.

\begin{theorem}[Main Results]
\label{thm:main-results}
In the $\mathrm{C}^4$ verification setting with the refinement site
$\site_{\mathrm{ref}}$:
\begin{enumerate}
\item \emph{(Grothendieck topology)} The covering families of
  Definition~\ref{def:refinement-topology} satisfy the Grothendieck axioms
  (Proposition~\ref{prop:grothendieck-axioms}), and all operations are
  Z3-decidable.
\item \emph{(Descent)} For any covering family $\covers$ of $\varphi_\top$
  and any sheaf type $A$, descent data glue to a global section
  iff $\Hone = 0$ (Theorem~\ref{thm:refinement-descent}).
\item \emph{(Automatic $\Hone = 0$)} For proposition-valued sheaves,
  $\Hone = 0$ always holds by Hedberg's theorem
  (Theorem~\ref{thm:hedberg-collapse}), making descent automatic.
\item \emph{(Spectral degeneration)} The spectral sequence degenerates at
  $E_2$: checking fiber proofs and pairwise overlaps suffices
  (Theorem~\ref{thm:spectral-degeneration}).
\item \emph{(Künneth)} Interprocedural composition uses the product cover;
  Z3 pruning reduces the product to $k \le mn$ non-trivial fibers
  (Theorem~\ref{thm:kunneth-ref}).
\item \emph{(Kan extension)} Cover morphisms give free proof reuse:
  a coarse proof transports to any finer cover at Z3 implication cost only
  (Theorem~\ref{thm:kan-extension}).
\item \emph{(CEGAR)} Counterexample-guided cover refinement correctly
  discovers the right cover and terminates with a genuine proof or
  a genuine counterexample (Theorem~\ref{thm:cegar-correct}).
\item \emph{(Incremental caching)} Mayer--Vietoris structure gives
  fiber-granular invalidation; transport-stable caching extends
  hit rates exponentially (Propositions~\ref{prop:cache-invalidation}
  and~\ref{prop:cache-amplification}).
\end{enumerate}
All of these results rely essentially on the synergy between cubical
type theory (transport, paths, $\hfilling$) and sheaf cohomology (covers,
descent, $\Hone$).  Neither theory alone provides the full structure.
\end{theorem}

\begin{remark}[The Duck-Type Site as a Special Case]
The duck-type fiber site of earlier chapters is a full sub-site of
$\site_{\mathrm{ref}}$ (Corollary~\ref{cor:subsumption}).  All results
of this chapter apply to the duck-type setting by restriction.
However, the refinement site enables:
\begin{itemize}
\item Numeric range proofs (positivity, bounds, modular arithmetic).
\item Structural string predicates (non-empty, matching a regex).
\item Relational predicates (ordering of two inputs, non-aliasing).
\item CEGAR-discovered covers that adapt to the program's actual behavior.
\end{itemize}
These capabilities are essential for real-world Python verification and
are the primary motivation for the refinement site replacement.
\end{remark}
