\chapter{The Pipeline: Module Name to Certificate}
\label{ch:pipeline}

The \Deppy{} pipeline is the orchestration layer that takes a
sidecar file path and emits a \Lean{} certificate.  Its entry point
is \texttt{deppy.lean.sidecar\_runner.run\_sidecar}; everything
else is a transformation on the \texttt{SidecarModule}.

This chapter walks through every stage in order.  The reader should
expect file-and-line citations; we are documenting the
\emph{actual} implementation.

\section{Architecture overview}

\begin{figure}[h!]
\centering
\begin{tikzpicture}[
  node distance=8mm and 12mm,
  every node/.style={font=\small},
  box/.style={
    draw, rounded corners=2pt, align=center,
    minimum width=33mm, minimum height=8mm,
    fill=blue!4,
  },
  trustbox/.style={
    draw, rounded corners=2pt, align=center,
    minimum width=33mm, minimum height=8mm,
    fill=green!7,
  },
  outbox/.style={
    draw, rounded corners=2pt, align=center,
    minimum width=33mm, minimum height=8mm,
    fill=orange!10,
  },
  arr/.style={->, >=Stealth, thick},
]

\node[box]                            (deppyfile) {\texttt{*.deppy}\\sidecar file};
\node[box, right=of deppyfile]        (parser)    {\texttt{SidecarModule}\\\texttt{.from\_file}};
\node[box, right=of parser]           (sm)        {\texttt{SidecarModule}\\dataclass};

\node[box, below=of deppyfile]        (importlib) {\texttt{importlib}\\\texttt{.import\_module}};
\node[box, below=of parser]           (audit)     {\texttt{audit\_all}\\axiom grounding};
\node[box, below=of sm]               (mech)      {\texttt{mechanize}\\axioms→Lean Props};

\node[box, below=of importlib]        (livesrc)   {\texttt{inspect}\\\texttt{.getsource}};
\node[box, below=of audit]            (bodylink)  {\texttt{link\_all}\\body translation};
\node[box, below=of mech]              (verify)    {\texttt{verify\_all}\\PSDL→ProofTerm};

\node[trustbox, below=of livesrc]     (kernel)    {\texttt{ProofKernel}\\\texttt{.verify}};
\node[trustbox, below=of bodylink]    (z3)        {\texttt{discharge\_foundations}\\Z3 backend};
\node[trustbox, below=of verify]      (counter)   {\texttt{run\_counterexample\_search}};

\node[outbox, below=of kernel]        (annot)     {\texttt{build\_annotation\_preamble}};
\node[outbox, below=of z3]            (cert)      {\texttt{render\_sidecar\_section}};
\node[outbox, below=of counter]       (leanrun)   {\texttt{check\_lean\_module\_source}};

\node[outbox, below=of cert, fill=red!10]  (final) {\texttt{certify\_or\_die}\\\texttt{*.lean}};

\draw[arr] (deppyfile) -- (parser);
\draw[arr] (parser)    -- (sm);
\draw[arr] (sm.south)     -- (mech.north);
\draw[arr] (sm.south)     -- (audit.north);
\draw[arr] (deppyfile) |- ([xshift=-1mm]importlib.east);
\draw[arr] (importlib) -- (livesrc);
\draw[arr] (livesrc)   -- (kernel);
\draw[arr] (audit)     -- (bodylink);
\draw[arr] (mech)      -- (verify);
\draw[arr] (bodylink)  -- (z3);
\draw[arr] (verify)    -- (counter);
\draw[arr] (kernel)    -- (annot);
\draw[arr] (z3)        -- (cert);
\draw[arr] (counter)   -- (leanrun);
\draw[arr] (annot.south) -| (final.west);
\draw[arr] (cert)      -- (final);
\draw[arr] (leanrun.south) -| (final.east);

\end{tikzpicture}
\caption{The \Deppy{} pipeline.  Blue boxes are inputs and parser
stages; green boxes are checking stages with their own trust
contributions; orange boxes are output emitters; the red box is the
final certification gate.}
\label{fig:pipeline}
\end{figure}

The pipeline is data-flow oriented: each stage reads the
\texttt{SidecarModule} (and optionally the outputs of earlier
stages) and produces a structured report.  No stage modifies the
\texttt{SidecarModule}; the result of running the pipeline is a
\texttt{SidecarCertificateSection} containing the rendered \Lean{}
text and a counter-rich report.

\section{Stage 0: Module-name resolution}

The pipeline starts at \texttt{run\_sidecar(deppy\_path, out\_path,
module\_name)} (file: \texttt{deppy/lean/sidecar\_runner.py}, lines
114--185).

\begin{lstlisting}[caption={\texttt{run\_sidecar}: high-level control flow.}]
def run_sidecar(deppy_path, *, out_path=None, module_name=""):
    deppy_path = Path(deppy_path)
    if out_path is None:
        out_path = deppy_path.with_suffix(".lean")
    out_path = Path(out_path)

    sm = SidecarModule.from_file(str(deppy_path))
    target_module = module_name or getattr(sm, "name", "") or ""
    library_summary = _inspect_library(target_module)

    t0 = time.perf_counter()
    section = render_sidecar_section(
        sm,
        module_name=f"Deppy_{deppy_path.stem}",
        library_source_summary=library_summary,
    )
    elapsed_ms = (time.perf_counter() - t0) * 1000.0
    out_path.write_text(section.full_text, encoding="utf-8")

    # Round-trip through Lean toolchain.
    ...
\end{lstlisting}

\texttt{\_inspect\_library} (line 101) calls
\texttt{importlib.import\_module(target\_module)} and reads
\texttt{module.\_\_version\_\_} and
\texttt{module.\_\_file\_\_} for the certificate's provenance
header.  When the import fails, the summary becomes
``\texttt{<module> (not importable)}'' and the certificate proceeds
with empty provenance --- but every verify-block inside that
certificate will fail to resolve its function and the certificate
will record this honestly.

\section{Stage 1: Parsing the sidecar}

\texttt{SidecarModule.from\_file(path)} reads the file and produces a
\texttt{SidecarModule}.  The parser supports both YAML-like
indentation-based blocks and a sequence of top-level declarations
separated by blank lines.

The parser populates:
\begin{itemize}[leftmargin=1.7em]
  \item \verb|sm._foundations: dict[str, FoundationDecl]|
  \item \verb|sm._specs: dict[str, ExternalSpec]|
  \item \verb|sm._axioms: dict[str, AxiomDecl]|
  \item \verb|sm._lemmas: list[LemmaDecl]|
  \item \verb|sm._verifies: list[VerifyDecl]|
  \item \verb|sm._proofs: list[ProofDecl]| (legacy)
  \item \verb|sm._code_types: dict[str, str]|
  \item \verb|sm._lean_imports: list[str]|
  \item \verb|sm._lean_preamble: str|
\end{itemize}

The parser is permissive: unrecognised top-level keys produce a
warning but do not abort.  Malformed indentation produces a
\texttt{SyntaxError} with a line number.

\section{Stage 2: \texttt{render\_sidecar\_section}}

The main rendering pipeline lives in
\texttt{deppy/lean/sidecar\_certificate.py}, function
\texttt{render\_sidecar\_section} (lines 100--296+).  It dispatches
to thirteen sub-emitters, each producing a fragment of the final
\Lean{} text plus a structured report.

The sub-emitters, in execution order:

\begin{enumerate}[leftmargin=1.7em]
  \item \texttt{derive\_all(sm)} --- run the legacy \texttt{by:}-clause
    derivations on each \texttt{ProofDecl}.
  \item \texttt{audit\_all(sm)} --- check the live target library
    for each axiom/spec; compute a \texttt{AxiomGrounding} per
    axiom (\texttt{GROUNDED} / \texttt{TARGET\_MISSING} /
    \texttt{METHOD\_MISSING} / \texttt{NL\_DOC}).
  \item \texttt{mechanize(sm, audit)} --- walk every grounded
    axiom AST and emit concrete \Lean{} \texttt{Prop}s.
  \item \texttt{link\_all(sm, audit, mech)} --- run
    \texttt{deppy.lean.body\_translation} on every method named
    by an axiom; record the translated Lean text per method.
  \item \texttt{attempt\_all(sm, audit, body\_link)} --- attempt
    code-proofs by composing the body translation with the
    cited foundation/axiom; the kernel verifies each, the
    pipeline records the \texttt{VerificationResult}.
  \item \texttt{link\_cubical(body\_link)} --- run the
    intra-procedural cubical AST analysis.
  \item \texttt{verify\_all(sm)} --- the per-\texttt{verify}-block
    pipeline (\S\ref{sec:verify-pipeline}).
  \item \texttt{link\_cohomology(body\_link)} --- run the C⁰/C¹/C²
    cohomology emitter on the linked methods.
  \item \texttt{discharge\_foundations(sm)} --- Z3 each foundation;
    promote to \texttt{Z3\_VERIFIED} when accepted.
  \item \texttt{discharge\_kan\_obligations(body\_link)} --- the
    Kan-fill obligations from the cubical pass.
  \item \texttt{link\_bicomplex(body\_link)} --- the cubical /
    simplicial bicomplex.
  \item \texttt{link\_fibration(sm, audit, body\_link, kernel)}
    --- run \texttt{FibrationVerifier} on
    \texttt{verify} blocks with a \texttt{fibration:} clause.
  \item \texttt{run\_counterexample\_search(sm, foundation\_discharge)}
    --- chase counterexamples for every Z3-attempted but
    unverified foundation.
  \item \texttt{invoke\_all(body\_link)} --- the B1--B16 audit
    sub-systems.
  \item \texttt{probe\_all\_proofterms()} --- C1--C7 audit on the
    \texttt{ProofTerm} class hierarchy.
  \item \texttt{build\_annotation\_preamble(sm)} --- the
    annotation-driven concrete preamble.
\end{enumerate}

After all sub-emitters complete, the function assembles the final
\Lean{} text in this order:

\begin{enumerate}[leftmargin=1.7em]
  \item Header comment with module name and provenance.
  \item Annotation-driven concrete preamble (types, methods,
    properties from binders).
  \item Mechanization preamble (vocabulary from grounded axioms).
  \item Body links (\texttt{def} declarations for each translated
    method, behind comments to avoid colliding with the preamble's
    opaque counterparts).
  \item Foundations.
  \item Sidecar axioms.
  \item Lemmas.
  \item Theorems (one per legacy \texttt{ProofDecl}).
  \item Verify-block sections (one per \texttt{VerifyDecl}).
  \item Cubical analysis section.
  \item Cohomology section.
  \item Bicomplex section.
  \item Counterexample-rejected theorems (when a foundation was
    counter-examined).
  \item Subsystem invocations and proofterm-zoo probes
    (audit appendices).
\end{enumerate}

\section{Stage 3: Audit and grounding}
\label{sec:audit}

\texttt{audit\_all} (file:
\texttt{deppy/lean/sidecar\_source\_audit.py}) walks every
\texttt{AxiomDecl} and tries to resolve its \texttt{target} (a
dotted Python name).  The four possible outcomes:

\begin{description}
  \item[\texttt{GROUNDED}] the target resolves; every method
    referenced in the axiom's statement is callable; the statement
    parses.  The certificate emits a concrete Lean Prop.
  \item[\texttt{TARGET\_MISSING}] the target does not resolve.  The
    axiom is admitted as opaque but flagged.
  \item[\texttt{METHOD\_MISSING}] the target resolves but a
    referenced method does not.  Same outcome as
    \texttt{TARGET\_MISSING}.
  \item[\texttt{NL\_DOC}] the statement is natural-language prose
    (e.g. ``the area is non-negative''), not a Python expression.
    Admitted as opaque, no concrete Prop.
\end{description}

The audit feeds into both \texttt{mechanize} (which only
mechanises \texttt{GROUNDED} axioms) and the certificate header
which counts axioms in each grounding bucket.

\section{Stage 4: Mechanization}
\label{sec:mechanize}

\texttt{mechanize(sm, audit)} (file:
\texttt{deppy/lean/sidecar\_mechanization.py}) is the heart of the
``concrete Props'' transformation.  It runs in two passes:

\paragraph{Pass 1: vocabulary collection.}  A
\texttt{\_VocabCollector} walks each grounded axiom's AST and
records:
\begin{itemize}[leftmargin=1.7em]
  \item \texttt{class\_methods : \{class$\to$\{method$\to$arity\}\}};
  \item \texttt{class\_ctor\_arity};
  \item \texttt{attrs} (every \texttt{.attr} access);
  \item \texttt{free\_fns : \{name$\to$arity\}};
  \item \texttt{prop\_fns} (free fns appearing at top-level
    Prop position);
  \item \texttt{free\_vars};
  \item \texttt{classes}.
\end{itemize}

\paragraph{Pass 2: per-axiom Lean rendering.}  Each grounded axiom's
statement is re-parsed and rendered against the vocabulary.  The
mapping is documented in the docstring (file
\texttt{deppy/lean/sidecar\_mechanization.py}, lines 7--40):

\begin{itemize}[leftmargin=1.7em]
  \item \texttt{ClassName.method(args)} → \texttt{(Class\_method args)}
  \item \texttt{ClassName.attr} → \texttt{(Class\_attr)}
  \item \texttt{var.attr} → \texttt{(dot\_attr var)}
  \item \texttt{var.method(args)} → \texttt{(dot\_method var args)}
  \item \texttt{Class(args)} → \texttt{(Class\_mk args)}
  \item \texttt{func(args)} → \texttt{(func args)}
\end{itemize}

The rendering emits a \texttt{def Sidecar\_X\_prop : Prop := <body>}
(when grounded and the rendering succeeds) or a fallback
\texttt{opaque Sidecar\_X\_prop : Prop} (when not).

\section{Stage 5: Body translation}
\label{sec:body-translation}

\texttt{link\_all} runs
\texttt{deppy.lean.body\_translation.translate\_body} on each method
referenced by an axiom or verify block.

\paragraph{Type-inference pass.}  An \texttt{\_InferTypes} walker
seeds with parameter annotations and propagates types through
operations: \texttt{x.append(...)} forces \texttt{x : list};
\texttt{len(x)} forces \texttt{x : list / dict / str / tuple};
arithmetic operations propagate numeric types; etc.  See
\texttt{infer\_locals\_types} (lines 127--179 of
\texttt{body\_translation.py}).

\paragraph{Translation pass.}  A \texttt{\_BodyT} walker uses the
inferred types to render Lean.  Standard rules:
\begin{itemize}[leftmargin=1.7em]
  \item \texttt{Int}/\texttt{Float}/\texttt{Bool} arithmetic translates
    directly.
  \item \texttt{if/elif/else} → \texttt{if-then-else}.
  \item \texttt{return} bound through to the result.
  \item Recursive self-calls preserved.
  \item \texttt{xs[:n]} → \texttt{xs.take n}; \texttt{xs[n:]} →
    \texttt{xs.drop n}; \texttt{xs[i]} → \texttt{xs.get!}.
  \item \texttt{len(xs)} → \texttt{xs.length} when \texttt{xs : List},
    \texttt{xs.size} when \texttt{HashMap}.
  \item Function calls preserved.
\end{itemize}

When the body uses constructs the translator cannot handle
(\texttt{eval}, dynamic class creation, C-extension
fall-throughs, list comprehensions with multiple clauses, etc.), it
returns \texttt{sorry} with a note.  The certificate flags this in
the \texttt{body\_links\_sorry} count.

\section{Stage 6: \texttt{verify\_all} --- the verify-block pipeline}
\label{sec:verify-pipeline}

\texttt{verify\_all(sm)} (file:
\texttt{deppy/lean/sidecar\_verify.py}) is the most directly relevant
stage for \PSDL{} users.  We summarise its structure here; the
worked example is in \Cref{ch:examples}.

\subsection{Per-block flow}

\texttt{\_process\_one(decl, *, foundations, kernel,
concrete\_prop)} (lines 167+) runs eight steps for each verify
block:

\begin{enumerate}[leftmargin=1.7em]
  \item \textbf{Resolve function.}  \texttt{\_resolve(decl.function)}
    walks the dotted path through \texttt{importlib} and
    \texttt{getattr}.  When it returns \texttt{None}, the outcome
    records ``could not resolve'' and the block aborts.

  \item \textbf{Read source.}  \texttt{inspect.getsource(method)};
    when \texttt{method} is a \texttt{property}, descend to
    \texttt{method.fget}.  The source is dedented via
    \texttt{textwrap.dedent} and SHA-256 hashed for the
    \texttt{source\_hash} field.

  \item \textbf{Translate body.}
    \texttt{deppy.lean.body\_translation.translate\_body(fn\_node,
    type\_context=ctx, predicate\_registry=preds, param\_types=...)}.

  \item \textbf{Compile PSDL (if present).}  When \texttt{decl.psdl\_proof}
    is non-empty, \texttt{deppy.proofs.psdl.compile\_psdl(source,
    foundations=fdict, ctx=ctx)} produces a
    \texttt{PSDLCompileResult}.

  \item \textbf{Fibration descent (if requested).}  When
    \texttt{decl.fibration\_type} is non-empty, the pipeline runs
    \texttt{FibrationVerifier.verify\_per\_fiber(fn\_node, fibre)}
    --- one kernel call per fibre.

  \item \textbf{Cubical analysis (if requested).}  When
    \texttt{decl.use\_cubical} is True, the pipeline runs
    \texttt{build\_cubical\_set(fn\_node)} and records the cell
    counts and Kan candidates.

  \item \textbf{Construct ProofTerm.}  In rough order of preference:
    PSDL's compiled term, when present; otherwise an
    \texttt{Unfold(method) \(\circ\) AxiomInvocation(decl.foundation)};
    otherwise a \texttt{Structural} fallback.

  \item \textbf{Kernel verification.}  Build a \texttt{Judgment}
    over the property's \texttt{RefinementType} and call
    \texttt{kernel.verify(ctx, goal, term)}.  Record the
    \texttt{VerificationResult}'s success bool, trust level,
    axioms used, and message.
\end{enumerate}

\subsection{Theorem emission}

When \texttt{kernel\_success == True}, \texttt{verify\_all} emits a
\Lean{} theorem section:

\begin{lstlisting}[language=lean4]
-- Verify block: dist_nonneg_psdl
-- Function: sympy.geometry.point.Point.distance
-- Property: Point.distance(p, q) >= 0
-- Foundation: Real_sqrt_nonneg
-- Source SHA-256: <hash>
-- Trust level: STRUCTURAL_CHAIN
-- Axioms used: ["Real_sqrt_nonneg"]
theorem Verified_dist_nonneg_psdl :
    Verified_dist_nonneg_psdl_property := by
  intros p q
  -- PSDL compiled output:
  by_cases h : isinstance other Point
  · unfold Point.distance
    exact Foundation_Real_sqrt_nonneg
    have h_<lineno> : sum_zip_sub_sq self other ≥ 0 := by omega
  · exact absurd ‹_› (by simp)
\end{lstlisting}

\texttt{Verified\_dist\_nonneg\_psdl\_property} is the concrete
\texttt{Prop} built by \texttt{build\_annotation\_preamble} from
the \texttt{binders}, \texttt{function}, and \texttt{property}
fields.  See \Cref{ch:lean}.

\section{Stage 7: Foundation discharge (Z3)}
\label{sec:foundation-discharge}

\texttt{discharge\_foundations(sm)} (file:
\texttt{deppy/lean/sidecar\_foundation\_discharge.py}, lines 1--150)
attempts a Z3 discharge of every foundation.  The flow:

\begin{enumerate}[leftmargin=1.7em]
  \item \texttt{\_looks\_arithmetic(stmt)} --- skip statements that
    contain \texttt{sqrt}, \texttt{sin}, \texttt{is\_collinear},
    etc.
  \item \texttt{\_strip\_when\_clause(stmt)} --- split off any
    ``\texttt{... when guard}'' precondition.
  \item \texttt{\_rewrite\_for\_z3(stmt)} --- combine guard and core
    into an implication \texttt{guard => core}.
  \item \texttt{\_collect\_binders(formula)} --- identify free
    variables and assign each to \texttt{int}.
  \item Build a \texttt{Z3Proof(formula, binders)} and call
    \texttt{kernel.verify} on a synthetic \texttt{Judgment}.  When
    Z3 returns \texttt{unsat} on the negation, mark
    \texttt{z3\_verified=True}.
\end{enumerate}

The certificate's foundation section then emits two shapes:

\begin{itemize}[leftmargin=1.7em]
  \item For \texttt{z3\_verified} foundations: a real
    \texttt{theorem Foundation\_X : Foundation\_X\_prop := by
    intros; omega} (or similar tactic).
  \item For unverified foundations: \texttt{axiom Foundation\_X :
    Foundation\_X\_prop} with the statement preserved as a comment.
\end{itemize}

In the \texttt{sympy\_geometry.deppy} run, 7 of 14 foundations
discharge via Z3 (the linear-arithmetic ones); the rest stay
admitted.

\section{Stage 8: Counterexample search}
\label{sec:counter-search}

\texttt{run\_counterexample\_search(sm,
foundation\_discharge\_report)} (file:
\texttt{deppy/lean/sidecar\_counterexample\_link.py}, lines 1--92)
runs only on foundations Z3 \emph{attempted} but did not verify.
For each, it builds a \texttt{Judgment} whose \texttt{type\_} is
the \texttt{RefinementType} with predicate the foundation's
formula, and asks \texttt{CounterexampleFinder} to find a model
satisfying the negation.

When a counterexample is found:
\begin{itemize}[leftmargin=1.7em]
  \item \texttt{outcome.found = True} with a model dict;
  \item the certificate marks the foundation as
    \texttt{REJECTED};
  \item \emph{transitively}: every foundation that depends on it (via
    \texttt{sm.\_foundation\_deps}) is also rejected;
  \item every theorem citing a rejected foundation is replaced in
    the certificate with an honest rejection notice.
\end{itemize}

This is the foundation of the \texttt{rejected\_theorem\_count}
gate.  When non-zero, \texttt{certify\_or\_die} refuses
certification.

\section{Stage 9: Concrete-Props preamble}
\label{sec:concrete-props}

\texttt{build\_annotation\_preamble(sm)} (file:
\texttt{deppy/lean/sidecar\_annotation\_preamble.py}) reads:
\begin{itemize}[leftmargin=1.7em]
  \item every \texttt{verify} block's \texttt{binders};
  \item every \texttt{verify} block's \texttt{function} dotted path;
  \item every \texttt{verify} block's \texttt{property};
  \item the top-level \texttt{code\_types} dict;
  \item the top-level \texttt{lean\_imports} and
    \texttt{lean\_preamble} text.
\end{itemize}

It emits:

\begin{enumerate}[leftmargin=1.7em]
  \item \texttt{abbrev <T> : Type := Int} for each unique
    binder type (the flat-Int simplification we acknowledge in
    \Cref{ch:lean}).
  \item \texttt{axiom <name> : <signature>} for each entry in
    \texttt{code\_types}.
  \item \texttt{def <Verified\_X\_property> : Prop := <translated>}
    for each verify block whose property translates to a non-trivial
    Lean expression.
\end{enumerate}

The verifying theorems then have concrete goal types that elaborate
against the PSDL-emitted tactics.  The
\texttt{annotation\_preamble.concrete\_props} dict maps verify-block
names to \texttt{(prop\_name, prop\_body)} for use by
\texttt{verify\_all}.

\section{Stage 10: Lean round-trip}
\label{sec:lean-roundtrip}

\texttt{check\_lean\_module\_source(text, timeout\_s=60)} (file:
\texttt{deppy/lean/lean\_runner.py}) writes the certificate to a
temporary file, invokes the \texttt{lean} binary on it, and
captures stdout/stderr/exit-code.

\begin{lstlisting}[caption={Outline of \texttt{check\_lean\_module\_source}.}]
def check_lean_module_source(source, *, timeout_s=60):
    with NamedTemporaryFile(suffix=".lean", mode="w") as fp:
        fp.write(source)
        fp.flush()
        result = subprocess.run(
            ["lean", fp.name],
            capture_output=True,
            timeout=timeout_s,
            text=True,
        )
    return LeanResult(
        ok=result.returncode == 0,
        stdout=result.stdout,
        stderr=result.stderr,
    )
\end{lstlisting}

When \Lean{} accepts (exit 0), \texttt{lean\_runner\_ok = True} and
the certificate's trust level for \texttt{LeanProof} sub-terms
elevates to \texttt{KERNEL}.

When \Lean{} rejects, \texttt{lean\_runner\_ok = False}.  The
\texttt{certify\_or\_die} gate then refuses certification, listing
``\texttt{lean\_runner round-trip rejected}'' as a failure.

\subsection{Caveats}

The \Lean{} binary must be on the user's \texttt{PATH}.  When it
isn't, \texttt{check\_lean\_module\_source} catches the
\texttt{FileNotFoundError} and \texttt{lean\_runner\_invoked =
False}, but \texttt{certify\_or\_die} does not refuse certification
on this basis alone --- the user can still rely on the kernel-level
verification for soundness, just not on the round-trip guarantee.

The 60s default timeout is on the conservative side.  For the
\texttt{sympy\_geometry.deppy} run, \Lean{} typically returns in
under 10 seconds; for larger sidecars (\texttt{torch}, \texttt{numpy})
the timeout may need raising.

\section{Stage 11: \texttt{certify\_or\_die}}
\label{sec:certify-or-die}

The final gate is \texttt{\_certify\_or\_die(s)} (file:
\texttt{deppy/lean/sidecar\_runner.py}, lines 62--98):

\begin{lstlisting}[caption={The \texttt{certify\_or\_die} verdict.}]
def _certify_or_die(s) -> tuple[bool, list[str]]:
    failures: list[str] = []
    if s.soundness_gate_fail > 0:
        failures.append(
            f"soundness gate fired in {s.soundness_gate_fail} proof(s)")
    if s.counterexamples_found > 0:
        failures.append(
            f"{s.counterexamples_found} counterexample(s) found "
            "for admitted foundations")
    if getattr(s, "rejected_theorem_count", 0) > 0:
        failures.append(
            f"{s.rejected_theorem_count} theorem(s) cited "
            "counterexample-rejected foundations")
    if getattr(s, "hash_drift_count", 0) > 0:
        failures.append(
            f"{s.hash_drift_count} verify block(s) failed "
            "expects_sha256 hash check")
    if getattr(s, "counterexample_expectation_violations", 0) > 0:
        failures.append(
            f"{s.counterexample_expectation_violations} verify block(s) "
            "expected counterexample but none found")
    return (len(failures) == 0, failures)
\end{lstlisting}

A certificate is \texttt{CERTIFIED} only when every gate is clean:
\begin{itemize}[leftmargin=1.7em]
  \item No \texttt{Structural} leaf in a non-vacuous position
    (soundness gate).
  \item No counterexample on any admitted foundation.
  \item No theorem citing a rejected foundation.
  \item No hash-drift on any \texttt{expects\_sha256:}-tagged
    verify block.
  \item No \texttt{expecting\_counterexample: true} block where
    the search came up empty.
  \item \Lean{}~4 round-trip exit 0 (when invoked).
\end{itemize}

When any gate fires, \texttt{certify\_or\_die} returns False; the
certificate is still emitted (so the user can debug), but
\texttt{rr.certified = False} and the runner prints an
``\texttt{✗ NOT CERTIFIED}'' summary.

\section{Output reporting}

\texttt{print\_report(rr)} (lines 188--282) prints a structured
human-readable summary.  The report (paraphrased) lists:
\begin{itemize}[leftmargin=1.7em]
  \item Foundations declared (trust surface).
  \item Sidecar axioms by grounding bucket.
  \item Body links: total, exact, sorry, error.
  \item Code-proofs and the soundness gate.
  \item Cubical analysis: methods, total cells, Kan candidates.
  \item Cohomology counts: C⁰, C¹, C².
  \item Bicomplex: built, edges, faces, Euler χ.
  \item Fibration descent: outcomes / successes.
  \item Foundation Z3 discharge: total / attempted / verified.
  \item Counterexample search: foundations attempted /
    counterexamples found.
  \item Subsystem invocations and ProofTerm-zoo probes.
  \item Verify blocks: total / kernel-verified.
  \item Specs (informational), proofs total / derived / cited /
    sorry.
  \item Per-classification counts.
  \item \texttt{lean\_runner} verdict.
  \item \texttt{certify\_or\_die} verdict.
\end{itemize}

For \texttt{sympy\_geometry.deppy} a typical run prints
\texttt{14 foundations, 38 axioms, 11 verify blocks, 4 body links},
\texttt{7 of 14 foundations discharged via Z3}, \texttt{0
counterexamples found}, and either \texttt{✓ CERTIFIED} or a list
of certify failures.

\section{Where each \texttt{ProofTerm} lives in the pipeline}

A summary mapping the kernel \texttt{ProofTerm} subclasses to the
pipeline stages that construct them:

\begin{center}
\begin{tabularx}{\linewidth}{@{}lX@{}}
\toprule
\textbf{Term} & \textbf{Constructed by} \\
\midrule
\texttt{Refl}, \texttt{Sym}, \texttt{Trans}, \texttt{Cong} &
PSDL refl/sym/trans/cong; legacy chain proofs. \\
\texttt{Cut}, \texttt{Assume} &
PSDL block-chaining; explicit \texttt{forall(\dots)}. \\
\texttt{Z3Proof} &
PSDL \texttt{assert ..., "z3"}; foundation discharge. \\
\texttt{LeanProof} &
PSDL \texttt{lean("\dots")}; \texttt{by\_lean:} on a verify block. \\
\texttt{NatInduction}, \texttt{ListInduction}, \texttt{Cases} &
PSDL \texttt{while}, \texttt{for}, \texttt{match}; explicit
\texttt{induction(\dots)}. \\
\texttt{DuckPath}, \texttt{ConditionalDuckPath}, \texttt{FiberedLookup} &
\texttt{deppy.equivalence}; \texttt{DuckPath.auto\_construct};
PSDL \texttt{dispatch(\dots)}. \\
\texttt{EffectWitness}, \texttt{ConditionalEffectWitness} &
PSDL \texttt{try/except}; \texttt{lift}/\texttt{bind}/\texttt{expect}. \\
\texttt{TransportProof} &
PSDL \texttt{transport(\dots)}; cubical-effects transport. \\
\texttt{Fiber} &
PSDL \texttt{if isinstance}; verify-block fibration descent. \\
\texttt{Patch} &
\texttt{kan\_fill} sugar; the patching module. \\
\texttt{Cocycle}, \texttt{KanFill}, \texttt{CechGlue} &
The cohomology pipeline; the bicomplex linker. \\
\texttt{Univalence} &
\texttt{UnivalenceEngine.from\_equiv}; \texttt{equiv\_to\_path}. \\
\texttt{AxiomInvocation} &
PSDL \texttt{apply}; legacy chain/citation proofs. \\
\texttt{Unfold}, \texttt{Rewrite} &
PSDL \texttt{inline}, \texttt{unfold}, \texttt{rewrite}; legacy
\texttt{rewrite then cite}. \\
\texttt{Structural} &
PSDL \texttt{raise NotImplementedError}; safety pipeline
escape hatches. \\
\bottomrule
\end{tabularx}
\end{center}

\section{Cache and incremental verification}

The \texttt{deppy/pipeline/cache.py} module provides a content-keyed
cache for: \Zthree{} discharge results,
\texttt{check\_lean\_module\_source} results, body-translation
results, and per-axiom mechanizations.  The cache key is the
SHA-256 of the input text, so identical re-runs are essentially
free.

\texttt{deppy/pipeline/incremental.py} sketches a cross-run
incremental verifier that re-runs only verify blocks whose
function source SHA-256 has changed.  This is what would let a
maintainer rebuild a certificate after a one-line change without
re-running the entire pipeline.  The current implementation is
opt-in and not exercised by the canonical \texttt{run\_sidecar}
entry point.

\section{Summary}

The pipeline:
\begin{itemize}[leftmargin=1.7em]
  \item starts at a \texttt{.deppy} path and a target Python module
    name;
  \item parses the sidecar into a \texttt{SidecarModule};
  \item runs a sequence of report-producing transformations
    (audit, mechanize, link, verify, foundations, counterexample,
    \dots);
  \item assembles a \Lean{}~4 certificate;
  \item round-trips the certificate through the \Lean{} toolchain;
  \item gates the \texttt{CERTIFIED} verdict on a small set of
    explicit failure conditions in \texttt{certify\_or\_die}.
\end{itemize}

The next chapter dives into the \Lean{} target layer: how the
generated text is structured, how the trust chain bottoms at
concrete \Lean{} \texttt{Prop}s, and how the body-translation and
mechanization passes interact with the preamble.
