\chapter{Soundness: What Is Verified, What Is Admitted}
\label{ch:soundness}

This chapter audits the soundness story.  We give the trust surface
in detail, document the soundness gates, and articulate what
\Deppy{}'s certifications mean.

We do not claim metamathematical soundness; we claim
\emph{engineering soundness} (Theorem~\ref{thm:engsound}) and
document the trust surface so a careful user can decide whether
the certifications are sufficient for their use case.

\section{The trust surface}
\label{sec:trust-surface}

The trust surface of a \Deppy{} certification is the union of:

\begin{enumerate}[leftmargin=1.7em]
  \item \textbf{\Lean{}'s kernel.}  The certificate's \Lean{} text
    is round-tripped through the \texttt{lean} binary.  When
    \Lean{} accepts, we accept.  We do not duplicate \Lean{}'s
    type-checking.

  \item \textbf{\Zthree.}  Every formula discharged via
    \texttt{Z3Proof} is decided by \Zthree's SMT engine.  When
    \Zthree{} returns \texttt{unsat} on the negation, we accept.

  \item \textbf{CPython's faithful execution.}  We trust that
    CPython runs the source bytes \texttt{inspect.getsource}
    returned.  If CPython misimplements (or
    \texttt{inspect.getsource} returns a different source than what
    actually runs), the certification is meaningless.

  \item \textbf{The user's foundations.}  Each
    \texttt{foundation} that does not Z3-verify is admitted on
    the user's authority.  Their statements are reproduced in the
    certificate; a reviewer can refuse a foundation by editing the
    sidecar.

  \item \textbf{The user's sidecar axioms.}  Each \texttt{axiom}
    in a \texttt{spec} block is admitted on the user's authority.
    The certificate exposes them as \Lean{} \texttt{axiom}
    declarations with the original Python statement preserved as
    a comment.

  \item \textbf{The user's structural leaves.}  Each
    \texttt{Structural("\dots")} term in a \texttt{ProofTerm}
    contributes a structural assertion.  The soundness gate
    flags non-vacuous occurrences; vacuous occurrences (e.g.\
    exception-fibre vacuity) are accepted.

  \item \textbf{The user's escape hatches.}  Each PSDL
    \texttt{lean("\dots raw \dots")} or \texttt{by\_lean: |}
    block is raw \Lean{} text that \Deppy{}'s kernel does not
    inspect.  \Lean{}'s elaborator is the only check.

  \item \textbf{The user's monkey-patch discipline.}  We trust
    the user not to monkey-patch a class while a verified
    certificate is in scope.  The optional
    \texttt{expects\_sha256:} hash is a partial mitigation.
\end{enumerate}

\section{What is verified}

The verifier has positive results in the following sense:

\paragraph{Foundations Z3 discharges.}  When the foundation discharge
report's \texttt{z3\_verified == True}, the foundation is provable in
\Zthree's theory.  Trust \texttt{Z3\_VERIFIED}.  In
\texttt{sympy\_geometry.deppy}, 7 of 14 foundations achieve this.

\paragraph{PSDL-compiled \texttt{ProofTerm}s the kernel accepts.}
When \texttt{kernel.verify} returns \texttt{success=True} for a
PSDL-compiled term, the term has been checked at the trust level
returned (the \texttt{min} over all leaves).  This is a real
machine check on the proof tree.

\paragraph{\Lean{} certificates that round-trip.}  When the \Lean{}
runner returns exit 0 on the certificate, the entire collection of
declarations type-checks in \Lean{}'s elaborator and kernel.  The
verify-block theorems in particular have proof bodies the kernel
accepts.

\paragraph{Counterexample-cleared foundations.}  Foundations that
\Zthree{} attempted but did not verify, and for which counterexample
search returned no model, are at least \emph{not provably false}
under the user's binder typing (typically \texttt{int}).  They stay
admitted, but with the additional negative evidence.

\paragraph{Body translations marked exact.}  When
\texttt{translate\_body} returns \texttt{exact=True} with no
\texttt{sorry}, the function body has been faithfully expressed in
\Lean{} terms.  The certificate emits the translation behind a
comment for audit.

\paragraph{Type-grounded axioms.}  When an axiom's
\texttt{AxiomGrounding} is \texttt{GROUNDED}, all the methods,
classes, and attributes its statement references resolve on the
live target.  This is not a soundness verification --- a grounded
axiom can still be false --- but it is a sanity check that the
axiom is talking about something real.

\section{What is admitted}

The verifier has negative results in the following sense:

\paragraph{Non-discharged foundations.}  Z3 cannot discharge
\texttt{sqrt} / \texttt{sin} / non-linear / quantifier-alternating
foundations.  These admit at \texttt{AXIOM\_TRUSTED}.  In
\texttt{sympy\_geometry.deppy}: 7 of 14.

\paragraph{Sidecar axioms.}  Every \texttt{axiom} declaration
admits.  The certificate flags grounding (does it resolve?), but
not truth.  In \texttt{sympy\_geometry.deppy}: 38 axioms.

\paragraph{Lemmas without \texttt{lean\_proof}.}  Lemmas declared
without an explicit Lean proof admit as axioms.

\paragraph{\texttt{Structural} leaves in vacuous positions.}  The
soundness gate accepts these.  They are recorded but do not
contribute to a counterexample search.

\paragraph{Body translations that produced \texttt{sorry}.}  When
\texttt{translate\_body} cannot handle a control-flow shape
(e.g.\ try/except, dynamic dispatch, generator comprehension), the
returned Lean text contains \texttt{sorry}.  The certificate
counts these in \texttt{body\_links\_sorry}.  A verify block whose
target's body translation produces \texttt{sorry} cannot be
\texttt{KERNEL}-trust verified.

\paragraph{Method calls through descriptors / \texttt{\_\_getattr\_\_}.}
The pipeline does not trace through \texttt{\_\_getattr\_\_}-based
proxies in the body translation pass; method calls on the proxy
emit a \texttt{getattr} application that the certificate's preamble
declares as opaque.  \texttt{FiberedLookup} in the kernel is the
proper way to verify these, but the verify-block path does not
currently engage it automatically.

\paragraph{Non-arithmetic Z3 obligations in PSDL.}  When PSDL emits
\texttt{Z3Proof} for a non-arithmetic formula, Z3 returns
\texttt{unknown} or fails to encode; the kernel rejects the
\texttt{Z3Proof} and the verify block fails.  This is a \emph{good}
behaviour --- failing rather than silently accepting --- but it
constrains the kinds of PSDL assertions that close.

\section{Soundness gates}
\label{sec:gates}

\Deppy{} runs several gates that can refuse certification.

\subsection{Soundness gate (\texttt{soundness\_gate\_fail})}

The soundness gate scans every emitted \texttt{ProofTerm} for
\texttt{Structural} leaves in non-vacuous positions.  A position
is \emph{vacuous} when:
\begin{itemize}[leftmargin=1.7em]
  \item it is the exception-fibre branch of a \texttt{Fiber} or
    \texttt{ConditionalEffectWitness} where the precondition
    excludes the value;
  \item it is the empty case of an \texttt{empty proof block} that
    the compiler eliminated;
  \item it is a \texttt{show\_goal} / \texttt{hint} / \texttt{pure}
    structural marker.
\end{itemize}

When a non-vacuous \texttt{Structural} leaf is detected (e.g.\ the
user wrote \texttt{raise NotImplementedError} in the value-fibre
branch), the gate fires.  \texttt{certify\_or\_die} refuses
certification.

In the current \texttt{sympy\_geometry} run, the soundness gate has
zero failures.

\subsection{Counterexample gate (\texttt{counterexamples\_found})}

When any admitted foundation has a counterexample, the gate fires.
\texttt{certify\_or\_die} refuses certification.

In the current run, 0 counterexamples found.

\subsection{Rejected-theorem gate (\texttt{rejected\_theorem\_count})}

When any theorem cites a foundation that has a counterexample,
the theorem is rejected.  The gate fires when this count is non-zero.

In the current run, 0 rejected theorems.

\subsection{Hash-drift gate (\texttt{hash\_drift\_count})}

When a verify block declares \texttt{expects\_sha256: <hex>} and
the live source's hash differs, the gate fires.

In the current run, no verify block uses \texttt{expects\_sha256}.

\subsection{Counterexample-expectation gate
(\texttt{counterexample\_expectation\_violations})}

When a verify block declares
\texttt{expecting\_counterexample: true} and the search
\emph{fails} to find one, the gate fires.  This is for testing
the counterexample finder itself.

In the current run, no such block.

\subsection{\Lean{} round-trip gate}

When the \Lean{} runner returns non-zero exit, the round-trip is
considered rejected.  \texttt{certify\_or\_die} refuses
certification only if the runner was successfully invoked; if the
\texttt{lean} binary cannot be found, the gate is skipped.

\section{What a \texttt{CERTIFIED} verdict means}

Putting the gates together, \texttt{CERTIFIED} means:

\begin{enumerate}[leftmargin=1.7em]
  \item No \texttt{Structural} leaf is in a non-vacuous position.
  \item No admitted foundation has a counterexample under the
    declared binder typing.
  \item No theorem cites a counterexample-rejected foundation.
  \item No verify block's \texttt{expects\_sha256} hash mismatched.
  \item No \texttt{expecting\_counterexample} was unfulfilled.
  \item \Lean{}~4 (when invoked) accepted the entire certificate.
\end{enumerate}

It does \emph{not} mean:
\begin{itemize}[leftmargin=1.7em]
  \item every cited foundation is actually true (some admit);
  \item every cited axiom is actually true (some admit);
  \item the property holds in any computational sense beyond what
    the bindings encode (e.g.\ a property quantifying over
    \texttt{Point := Int} elaborates over integers, not over the
    actual SymPy \texttt{Point} class).
\end{itemize}

The certification is therefore relative: ``relative to the
foundations and axioms admitted, and the binder types declared,
the property holds''.

\section{The role of \texttt{certify\_or\_die}}

\texttt{certify\_or\_die} (file:
\texttt{deppy/lean/sidecar\_runner.py}, lines 62--98) is the
central gate.  Its design is deliberately conservative:

\paragraph{Refuse on any uncertainty.}  When Z3 returns
\texttt{unknown}, the foundation is not Z3-verified.  When the
\Lean{} runner cannot be invoked, the round-trip is not certified.
When a structural leaf is in a non-vacuous position, the proof is
not sound.

\paragraph{Be specific about why.}  Each failure is recorded as a
human-readable string.  The runner's output lists every failure;
the user can fix one at a time.

\paragraph{Do not silently accept.}  There is no ``accept on the
basis of the kernel-only check, even if \Lean{} disagrees''.  The
\Lean{} round-trip is required when \texttt{lean} is on the
\texttt{PATH}.

\section{Limitations}
\label{sec:limits}

We close with the explicit list of limitations.

\subsection{Type system}

\begin{itemize}[leftmargin=1.7em]
  \item No universe polymorphism in the generated certificate.
    Everything is in \texttt{Type 0}.
  \item Class types are \texttt{abbrev T : Type := Int} aliases
    --- the flat-Int simplification.
  \item Refinement types are syntactically supported but the
    kernel does not check refinement predicates beyond the
    \Zthree{} encoder's reach.
  \item No coinductive specs.
  \item No higher inductive types.
\end{itemize}

\subsection{Body translation}

\begin{itemize}[leftmargin=1.7em]
  \item No \texttt{try / except} translation; emits \texttt{sorry}.
  \item No \texttt{while} translation; emits \texttt{sorry}.
  \item No \texttt{generator} or \texttt{yield} translation.
  \item No dynamic class creation.
  \item No \texttt{exec} / \texttt{eval}.
  \item No C-extension fall-throughs (\texttt{inspect.getsource}
    fails).
\end{itemize}

\subsection{Z3}

\begin{itemize}[leftmargin=1.7em]
  \item No transcendentals.
  \item No nested $\forall \exists$.
  \item No floating-point semantics (we use \texttt{Real}, not
    IEEE 754).
  \item No string reasoning beyond fixed alphabets.
\end{itemize}

\subsection{\Lean{} target}

\begin{itemize}[leftmargin=1.7em]
  \item Mechanization preamble is flat-\texttt{Int}; richer typed
    preambles are future work.
  \item Method bodies are emitted as \texttt{def := sorry}
    placeholders alongside the actual translation as comments
    (avoids elaboration ambiguity but loses unfold power).
  \item No \texttt{Mathlib} dependency by default; foundations
    involving real arithmetic admit.
  \item Round-trip is single-shot \texttt{lean} invocation; no
    interactive feedback loop.
\end{itemize}

\subsection{PSDL}

\begin{itemize}[leftmargin=1.7em]
  \item No partial-proof credit (LLM iteration affordance: future
    work).
  \item No tactic-suggestion mode.
  \item No counterexample-to-obligation feedback.
  \item Quantifier scoping is by simple variable shadowing; no
    explicit De Bruijn manipulation.
\end{itemize}

\subsection{Cohomology}

\begin{itemize}[leftmargin=1.7em]
  \item Cohomology cocycles are \emph{advisory} in the current
    implementation; they do not contribute trust to verify-blocks.
  \item No actual computation of $H^k$ groups; only Euler χ.
\end{itemize}

\subsection{Pipeline}

\begin{itemize}[leftmargin=1.7em]
  \item Single-pass; no incremental verification by default.
  \item \Lean{}'s 60s default timeout is hard-coded; longer
    runs may need adjustment.
  \item No parallelism between sub-emitters
    (\texttt{deppy/pipeline/simple\_parallel.py} sketches
    multiprocessing but is not engaged by \texttt{run\_sidecar}).
\end{itemize}

\section{Where to put trust}

A user reading a \Deppy{} certificate should:

\begin{enumerate}[leftmargin=1.7em]
  \item \textbf{Verify the certificate round-trips.}  Run
    \texttt{lean <certificate>.lean} and check exit 0.  This
    confirms the \Lean{} kernel accepts the entire body of
    declarations.

  \item \textbf{Audit foundations.}  Each \texttt{axiom Foundation\_X}
    is admitted.  Read its statement; if you do not believe it,
    refuse it.

  \item \textbf{Audit sidecar axioms.}  Each \texttt{axiom
    Sidecar\_X} is admitted.  Sympy axioms about geometric
    facts are typically defensible; library-specific axioms
    on a closed-source library are more questionable.

  \item \textbf{Read the report.}  Note the
    \texttt{soundness\_gate\_fail}, the
    \texttt{counterexamples\_found}, the
    \texttt{rejected\_theorem\_count}.  All zero is the goal.

  \item \textbf{Check binder typing.}  When a verify block declares
    \texttt{binders: x: int}, the verified property quantifies over
    \texttt{int}, not over arbitrary numeric types.  Be aware that
    a property verified over \texttt{int} may fail over
    \texttt{float} or \texttt{Decimal}.

  \item \textbf{Check the source hash.}  When a verify block
    declares \texttt{expects\_sha256:}, the certificate is pinned
    to a specific source byte sequence.  Re-running on a different
    \Sympy{} version requires re-hashing.

  \item \textbf{Beware monkey-patching.}  No monkey-patch detection
    runs at certificate-issuance time; do not patch verified
    classes.
\end{enumerate}

\section{Soundness, formally restated}

To close the chapter we restate Theorem~\ref{thm:engsound}:

\begin{theorem}[Engineering soundness, restated]
For every verify block whose certificate prints
\texttt{CERTIFIED}, the following hold:

\begin{enumerate}[leftmargin=1.7em]
  \item \textbf{Kernel acceptance.}  The PSDL-compiled
    \texttt{ProofTerm} (or the legacy by-clause derivation) was
    accepted by \texttt{ProofKernel.verify} returning
    \texttt{success=True} at the reported trust level.

  \item \textbf{SMT discharge.}  Every \texttt{Z3Proof} sub-term
    was checked by \Zthree{} (or by cached lookup) and returned
    \texttt{unsat} on the negation.

  \item \textbf{\Lean{} round-trip.}  When \texttt{lean} was on the
    \texttt{PATH}, the certificate's \Lean{} text was compiled
    successfully (exit 0).

  \item \textbf{Foundation status visible.}  Every cited foundation
    has a status declaration in the certificate (real
    \texttt{theorem} for Z3-discharged, \texttt{axiom} for
    admitted).  No foundation is silently introduced.

  \item \textbf{No silent admissions.}  No \texttt{Structural} leaf
    is in a non-vacuous position
    (\texttt{soundness\_gate\_fail = 0}); no theorem cites a
    counterexample-refuted foundation.

  \item \textbf{No false claims under the binder typing.}  The
    counterexample finder did not produce a model violating any
    cited foundation under its declared binders.
\end{enumerate}
\end{theorem}

This is the formal-but-engineering soundness statement \Deppy{}
makes.  The trust surface is large (it includes \Lean{}'s kernel,
\Zthree{}, CPython, and the user's admissions), but it is
\emph{enumerated}.  Reasoning about whether to trust a given
\Deppy{} certification reduces to reasoning about whether to trust
the surface enumerated in this chapter.

\section{Summary}

\begin{itemize}[leftmargin=1.7em]
  \item The trust surface is the union of \Lean's kernel, \Zthree's
    answer, CPython's faithful execution, and the user's
    admissions (foundations, axioms, structural leaves).
  \item \texttt{certify\_or\_die} runs five gates: soundness,
    counterexamples, rejected theorems, hash drift, counterexample
    expectation; plus the \Lean{} round-trip when invoked.
  \item \texttt{CERTIFIED} means relative-soundness: relative to
    the admissions, the property holds at the kernel-and-Lean
    level.
  \item Limitations are explicit and enumerated.
  \item A user reading a certificate should audit the foundations,
    sidecar axioms, binder types, and source hash; the certificate
    exposes everything for that audit.
\end{itemize}

The next chapter situates \Deppy{} among its conceptual neighbours
in related work and concludes.
